### 2021-02-02 README.Rmd
2nd commit of package structure

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
<!-- badges: end -->

The goal of ppseq is to ...

## Installation

You can install the development version of ppseq from GitHub with:

``` r
remotes::install_github("zabore/ppseq")
```

## Example

This is a basic example which shows you how to solve a common problem:

``` r
library(ppseq)
## basic example code
```

### 2021-02-02 R/temp.R

@@ -1,0 +1,22 @@
+#' Temporary R file to push the R folder to GitHub
+#'
+#' @description This function doesn't do anything useful, and I just created it 
+#' so that the R folder would be pushed to GitHub as empty folders don't get 
+#' committed. Will delete once we add real functions.
+#'
+#' @param x a numeric value
+#' @param y a numeric value
+#' 
+#' @return Returns the sum of \code{x} and \code{y}
+#'
+#' @examples
+#'
+#' temp(1, 10)
+#' temp(2, NA)
+#'
+#' @export
+
+temp <- function(x, y) {
+  sumxy <- sum(x, y, na.rm = TRUE)
+  return(sumxy)
+}

### 2021-02-02 covid-files/comp-Posterior.R

@@ -1,0 +1,111 @@
+
+#-----------------------------------------------------
+#  Beta distribution comparison 
+# Prob beta(c,d) < beta(a,b)
+#-----------------------------------------------------
+
+#### Posterior Design #####
+
+ # prob y1 < y0
+ posterior_y0.gt.y1 <- function(n0, n1, y0, y1, delta=0, prior=c(0.5,0.5)){ ## y0 standard     y1 experimental
+  SAMP <- 10000
+  out <- sum( rbeta(SAMP, prior[1] + y0, prior[2] + n0 - y0) > ( rbeta(SAMP, prior[1] + y1, prior[2] + n1 - y1) + delta ) ) / SAMP
+  return(out)
+ }
+
+find.Boundary <- function(j, theta, Tab, delta=0, prior=c(0.5,0.5) ){
+
+ Y=Tab$n1[j]; STOP <- FALSE
+ temp <- posterior_y0.gt.y1( n0=Tab$n0[j], n1=Tab$n1[j], y0=Tab$y0[j], y1=Y, delta=delta, prior=prior )
+ if( is.na(temp) ){ temp <- 0; print(paste0("NA - ",Tab[j,])) }
+ if( temp > theta ){ STOP <- TRUE }
+ while( !STOP ){
+  Y <- Y - 1
+  temp <- posterior_y0.gt.y1( n0=Tab$n0[j], n1=Tab$n1[j], y0=Tab$y0[j], y1=Y, delta=delta, prior=prior )
+  if( is.na(temp) ){ temp <- 0 }
+  if( Y == 0 ){ STOP <- TRUE }
+  if( temp > theta ){ STOP <- TRUE }
+ }
+
+ if( temp > theta ){ out <- Y 
+ } else{ out <- NA }
+
+ return( c(j, out) )
+}
+
+library(parallel)
+library(foreach)
+library(doParallel)
+
+###### Paralllel Computing Setup #########################
+ no_cores <- min(detectCores() - 1, 90)
+
+ if( grepl("Windows", sessionInfo()[4]$running) ){ 
+  cl <- makeCluster(no_cores); registerDoParallel(cl) 
+ } else{
+  library(doMC)
+  registerDoMC(no_cores)
+ }
+
+
+###########
+Max <- 2000
+N.tab <- expand.grid( list( seq(100,1500,100), seq(100,1500,100) ) )
+N.tab <- N.tab[-which( rowSums(N.tab) > Max ), ]
[…]

### 2021-02-02 covid-files/comp-pp.R

@@ -1,0 +1,173 @@
+
+#-----------------------------------------------------
+#  Beta distribution comparison 
+# Prob beta(c,d) < beta(a,b)
+#-----------------------------------------------------
+
+#### Posterior Design #####
+
+ # prob y1 < y0 | Data #
+ posterior_y0.gt.y1 <- function(y, n0, n1, delta=0, prior=c(0.5,0.5)){ ## y0 standard     y1 experimental
+  SAMP <- 5000
+  out <- sum( rbeta(SAMP, prior[1] + y[1], prior[2] + n0 - y[1]) > ( rbeta(SAMP, prior[1] + y[2], prior[2] + n1 - y[2]) + delta ) ) / SAMP
+  return(out)
+ }
+
+ # pred y1 < y0 | Data #
+ # d = y0,y1,n0,n1
+ pred_y0.gt.y1 <- function(d, N0, N1, theta=0.95, delta=0, prior=c(0.5,0.5)){
+
+  samp.Trial <- function(p, n){ return( rbinom( 1, n, p) ) }
+
+  SAMP <- 5000
+  p0 <- rbeta(SAMP, prior[1] + d[1], prior[2] + d[3] - d[1])
+  p1 <- rbeta(SAMP, prior[1] + d[2], prior[2] + d[4] - d[2])
+
+  if( d[3] < N0 ){ Y0 <- d[1] + sapply(p0, FUN=samp.Trial, n=N0-d[3])  
+  } else{ Y0 <- rep(d[1], SAMP); N0 <- d[3] }
+  if( d[4] < N1 ){ Y1 <- d[2] + sapply(p1, FUN=samp.Trial, n=N1-d[4])
+  } else{ Y1 <- rep(d[2], SAMP); N1 <- d[4] }
+
+  y <- cbind(Y0,Y1)
+  
+  post <- apply( y, MARGIN=1, FUN=posterior_y0.gt.y1, n0=N0, n1=N1, delta=0, prior=c(0.5,0.5) )
+  return( sum( as.numeric( post > theta  ) )/SAMP )
+ }
+
+
+ sim.pp.Trial <- function(n, p, N, theta, delta, prior){
+
+  sim.pp.Trial.1 <- function(p0, p1, n0, n1, N0, N1, theta, delta, prior){
+
+   rbinom. <- function(size, n=1, prob){ return( rbinom(n=n,size=size,prob=prob) ) }
+
+   y0 <- rbinom.(size=n0, prob=p0 ) 
[…]

### 2021-02-02 README.Rmd
setting up Travis-CI and code coverage badge

@@
 <!-- badges: start -->
+[![codecov](https://codecov.io/gh/zabore/ppseq/branch/master/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq)
 [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)

### 2021-02-17 README.Rmd
fixing a couple of build warnings

--- 2021-02-02 README.Rmd
@@
 [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+[![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=master)](https://travis-ci.com/zabore/ppseq)
+[![codecov](https://codecov.io/gh/zabore/ppseq/branch/master/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq)
 [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)

### 2021-02-17 R/calc_posterior.R

@@ -18,6 +18,7 @@
 #' @param direction "greater" (default) if interest is in p(y1 > y0) and "less" 
 #' if interest is in p(y1 < y0). For one-sample case, "greater" if interest is 
 #' in p(y1 > p) and "less" if interest is in p(y1 < p)
+#' @param p The target value to compare to in the one-sample case
 #' @param delta clinically meaningful difference between groups. 
 #' Typically 0 (default).
 #' @param prior hyperparameters of beta distribution, Beta(0.5, 0.5) is default
@@ -55,8 +56,8 @@ calc_posterior <- function(y, n, direction = "greater", p = NULL,
   
   if(length(y) == 2) {
     
-    rb0 <- rbeta(S, prior[1] + y[1], prior[2] + n[1] - y[1])
-    rb1 <- rbeta(S, prior[1] + y[2], prior[2] + n[2] - y[2]) 
+    rb0 <- stats::rbeta(S, prior[1] + y[1], prior[2] + n[1] - y[1])
+    rb1 <- stats::rbeta(S, prior[1] + y[2], prior[2] + n[2] - y[2]) 
     
     out <- ifelse(direction == "greater", 
                   mean(rb1 > rb0 + delta), 
@@ -65,7 +66,7 @@ calc_posterior <- function(y, n, direction = "greater", p = NULL,
     
   } else if(length(y) == 1) {
     
-    rb1 <- rbeta(S, prior[1] + y, prior[2] + n - y) 
+    rb1 <- stats::rbeta(S, prior[1] + y, prior[2] + n - y) 
     
     out <- ifelse(direction == "greater", mean(rb1 > p), mean(rb1 < p))
   }

### 2021-02-17 README.Rmd
Fixing the build badge

--- 2021-02-02 README.Rmd
@@
 [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+[![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=main)](https://travis-ci.com/zabore/ppseq)
+[![codecov](https://codecov.io/gh/zabore/ppseq/branch/master/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq)
 [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)

### 2021-03-02 README.Rmd
starting first vignette

--- 2021-02-02 README.Rmd
@@
 <!-- badges: start -->
+[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
+<!-- [![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=main)](https://travis-ci.com/zabore/ppseq) -->
 [![codecov](https://codecov.io/gh/zabore/ppseq/branch/master/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq)
-[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
 [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)

### 2021-03-03 R/calc_predictive.R
setup of one-sample expansion vignette

@@ -12,11 +12,10 @@
 #' @param y vector of length two containing number of responses observed so far
 #' c(y0, y1) for two-sample case; integer of number of responses observed so far
 #' for one-sample case
-#' @param n vector of length two containing the sample size sor far c(n0, n1) 
+#' @param n vector of length two containing the sample size so far c(n0, n1) 
 #' for two-sample case; integer of sample size so far for one-sample case
 #' @param direction "greater" (default) if interest is in p(y1 > y0) and "less" 
-#' if interest is in p(y1 < y0). For one-sample case, "greater" if interest is 
-#' in p(y1 > p) and "less" if interest is in p(y1 < p)
+#' if interest is in p(y1 < y0). 
 #' @param p The target value to compare to in the one-sample case
 #' @param delta clinically meaningful difference between groups. 
 #' Typically 0 (default).
@@ -24,7 +23,7 @@
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
 #' @param seed set the seed for the random number generation (default NULL)
-#' @param N the total planned sample size at the end of the trial, c(N0, N!)
+#' @param N the total planned sample size at the end of the trial, c(N0, N1)
 #' for two-sample case; integer of total planned sample size at end of trial
 #' for one-sample case
 #' @param theta The target posterior probability, default is 0.95

### 2021-03-03 R/sim_single_trial.R

@@ -1,0 +1,74 @@
+#' Simulate a single trial with posterior probability monitoring
+#'
+#' @description This function is meant to be used in the context of a 
+#' clinical trial with a binary endpoint. Interest is in the response rate. 
+#' The goal is to simulate response counts based on the number of patients at 
+#' accrued at each interim look, and calculate the posterior predictive 
+#' probability of success at the end of a trial, given the data available at 
+#' each interim analysis. 
+#'
+#' @param prob vector of length two containing the probability of response in 
+#' the standard of care and experimental arm c(p0, p1) for the two-sample case; 
+#' vector of response probability for one-sample case
+#' @param n matrix containing the total number of patients accrued so far at 
+#' each interim look in the standard of care (column 1) and experimental 
+#' (column 2) arms for two-sample case; vector of sample size accrued so far 
+#' at each interim look for one-sample case
+#' @param direction "greater" (default) if interest is in p(y1 > y0) and "less" 
+#' if interest is in p(y1 < y0). 
+#' @param p The target value to compare to in the one-sample case 
+#' (i.e. the unacceptable response rate)
+#' @param delta clinically meaningful difference between groups. 
+#' Typically 0 (default).
+#' @param prior hyperparameters of prior beta distribution.
+#' Beta(0.5, 0.5) is default
+#' @param S number of samples, default is 5000
+#' @param seed set the seed for the random number generation (default NULL)
+#' @param N the total planned sample size at the end of the trial, c(N0, N1)
+#' for two-sample case; integer of total planned sample size at end of trial
+#' for one-sample case
+#' @param theta The target posterior probability, default is 0.95
+#' 
+#' @return Returns the posterior predictive probability of interest
+#'
+#' @examples
+#'
+#' # Two-sample case
+#' calc_predictive(y = c(7, 12), n = c(50, 50), seed = 1, N = c(100, 100))
+#' 
+#' # One-sample case
+#' calc_predictive(y = 14, n = 50, p = 0.2, delta = NULL, seed = 1, N = 100)
+#'
+#' @export
+
+sim_single_trial <- function(prob, n, direction = "greater", p = NULL, 
+                             delta = 0, prior = c(0.5,0.5), S = 5000, 
+                             seed = NULL, 
+                             N, theta = 0.95){
+  
+  set.seed(seed)
+  
+  if(length(prob) == 2) {
+    
+  } else if(length(prob) == 1) {
+    
[…]

### 2021-03-03 R/sim_single_trial.R
adding an import

@@ -29,15 +29,8 @@
 #' for one-sample case
 #' @param theta The target posterior probability, default is 0.95
 #' 
-#' @return Returns the posterior predictive probability of interest
-#'
-#' @examples
-#'
-#' # Two-sample case
-#' calc_predictive(y = c(7, 12), n = c(50, 50), seed = 1, N = c(100, 100))
-#' 
-#' # One-sample case
-#' calc_predictive(y = 14, n = 50, p = 0.2, delta = NULL, seed = 1, N = 100)
+#' @return Returns a matrix of number of responses, sample size, and posterior
+#' predictive probability at each interim look for the one-sample case
 #'
 #' @export
 

### 2021-03-04 R/calc_posterior.R
some function cleanups

@@ -1,29 +1,29 @@
 #' Calculate a single posterior probability
 #'
 #' @description This function is meant to be used in the context of a 
-#' clinical trial with a binary endpoint. Interest is in the response rate. 
-#' For the two-sample case, the total number of responses for the 
-#' standard-of-care arm is y0 and the total number of responses for the 
-#' experimental arm is y1. The function samples from the posterior beta 
-#' distribution based on the data and the prior beta hyperparameters, and
-#' returns the posterior probability that y1 is greater than (or less than) y0
-#' given the data. The one-sample case is also available, in which a target p
+#' clinical trial with a binary endpoint. For the two-sample case, the total 
+#' number of events in the standard-of-care arm is y0 and the total number of 
+#' events in the experimental arm is y1. The function samples from the posterior 
+#' beta distribution based on the data and the prior beta hyperparameters, and
+#' returns the posterior probability that p1 is greater than (or less than) p0
+#' given the data. The one-sample case is also available, in which a target p0
 #' must be specified and the function returns the posterior probability that
-#' y1 is greater than (or less than) p given the data.
+#' p is greater than (or less than) p0 given the data.
 #'
 #' @param y vector of length two containing total responses c(y0, y1) for 
-#' two-sample case; integer of total responses for one-sample case
+#' two-sample case; integer of total responses y for one-sample case
 #' @param n vector of length two containing total sample size c(n0, n1) for 
-#' two-sample case; integer of total sample size for one-sample case
-#' @param direction "greater" (default) if interest is in p(y1 > y0) and "less" 
-#' if interest is in p(y1 < y0). For one-sample case, "greater" if interest is 
-#' in p(y1 > p) and "less" if interest is in p(y1 < p)
-#' @param p The target value to compare to in the one-sample case
+#' two-sample case; integer of total sample size n for one-sample case
+#' @param direction "greater" (default) if interest is in p(p1 > p0) and "less" 
+#' if interest is in p(p1 < p0) for two-sample case. For one-sample case, 
[…]

### 2021-03-04 R/sim_single_trial.R

@@ -1,53 +1,91 @@
 #' Simulate a single trial with posterior probability monitoring
 #'
 #' @description This function is meant to be used in the context of a 
-#' clinical trial with a binary endpoint. Interest is in the response rate. 
-#' The goal is to simulate response counts based on the number of patients at 
+#' clinical trial with a binary endpoint. The goal is to simulate event counts 
+#' from the binomial distribution based on the number of patients 
 #' accrued at each interim look, and calculate the posterior predictive 
-#' probability of success at the end of a trial, given the data available at 
-#' each interim analysis. 
+#' probability of success (or futility) at the end of a trial, 
+#' given the data available at each interim analysis. 
 #'
-#' @param prob vector of length two containing the probability of response in 
+#' @param prob vector of length two containing the probability of event in 
 #' the standard of care and experimental arm c(p0, p1) for the two-sample case; 
-#' vector of response probability for one-sample case
+#' integer of event probability for one-sample case
 #' @param n matrix containing the total number of patients accrued so far at 
 #' each interim look in the standard of care (column 1) and experimental 
 #' (column 2) arms for two-sample case; vector of sample size accrued so far 
 #' at each interim look for one-sample case
-#' @param direction "greater" (default) if interest is in p(y1 > y0) and "less" 
-#' if interest is in p(y1 < y0). 
-#' @param p The target value to compare to in the one-sample case 
-#' (i.e. the unacceptable response rate)
+#' @param direction "greater" (default) if interest is in P(p1 > p0) and "less" 
+#' if interest is in P(p1 < p0) for two-sample case. For one-sample case, 
+#' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
+#' @param p0 The target value to compare to in the one-sample case
 #' @param delta clinically meaningful difference between groups. 
 #' Typically 0 (default).
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
-#' @param seed set the seed for the random number generation (default NULL)
+#' @param seed set the seed for the random number generation. Default is NULL.
 #' @param N the total planned sample size at the end of the trial, c(N0, N1)
[…]

### 2021-03-10 R/calc_posterior.R
updated vignette draft

@@ -12,8 +12,8 @@
 #'
 #' @param y vector of length two containing total responses c(y0, y1) for 
 #' two-sample case; integer of total responses y for one-sample case
-#' @param n vector of length two containing total sample size c(n0, n1) for 
-#' two-sample case; integer of total sample size n for one-sample case
+#' @param n vector of length two containing the sample size so far c(n0, n1) 
+#' for two-sample case; integer of sample size so far n for one-sample case
 #' @param direction "greater" (default) if interest is in p(p1 > p0) and "less" 
 #' if interest is in p(p1 < p0) for two-sample case. For one-sample case, 
 #' "greater" if interest is in p(p > p0) and "less" if interest is in p(p < p0).

### 2021-03-10 R/sim_single_trial.R

@@ -13,7 +13,9 @@
 #' @param n matrix containing the total number of patients accrued so far at 
 #' each interim look in the standard of care (column 1) and experimental 
 #' (column 2) arms for two-sample case; vector of sample size accrued so far 
-#' at each interim look for one-sample case
+#' at each interim look for one-sample case. If only a single look will be done
+#' at the end of the trial, this can be an integer specifying the total sample
+#' size at the end of the trial N.
 #' @param direction "greater" (default) if interest is in P(p1 > p0) and "less" 
 #' if interest is in P(p1 < p0) for two-sample case. For one-sample case, 
 #' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
@@ -61,14 +63,16 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
     y0 <- stats::rbinom(n = 1, size = n[1, 1], prob = prob[1]) 
     y1 <- stats::rbinom(n = 1, size = n[1, 2], prob = prob[2]) 
     
-    for(i in 2:nrow(n)) { 
-      
-      y0 <- c(y0, y0[length(y0)] + stats::rbinom(n = 1, 
-                                                 size = n[i, 1] - n[i - 1, 1], 
-                                                 prob = prob[1])) 
-      y1 <- c(y1, y1[length(y1)] + stats::rbinom(n = 1, 
-                                                 size = n[i, 2] - n[i - 1, 2], 
-                                                 prob = prob[2])) 
+    if(length(n) > 1) {
+      for(i in 2:nrow(n)) { 
+        
+        y0 <- c(y0, y0[length(y0)] + stats::rbinom(n = 1, 
+                                                   size = n[i, 1] - n[i - 1, 1], 
+                                                   prob = prob[1])) 
+        y1 <- c(y1, y1[length(y1)] + stats::rbinom(n = 1, 
+                                                   size = n[i, 2] - n[i - 1, 2], 
+                                                   prob = prob[2])) 
+      }
     }
     
     pp <- purrr::pmap_dbl(list(y0, y1, n[, 1], n[, 2]), 
@@ -84,13 +88,17 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
   } else if(length(prob) == 1) {
     
     y1 <- stats::rbinom(n = 1, size = n[1], prob = prob) 
-    
-    for(i in 2:length(n)) { 
-      
-      y1 <- c(y1, y1[length(y1)] + stats::rbinom(n = 1, 
[…]

### 2021-03-17 R/calc_posterior.R
working on seq sim

@@ -23,23 +23,23 @@
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
-#' @param seed set the seed for the random number generation. Default is NULL.
 #' 
 #' @return Returns the posterior probability of interest
 #'
 #' @examples
+#' 
+#' set.seed(123)
 #'
 #' # Two-sample case
-#' calc_posterior(y = c(14, 23), n = c(100, 100), seed = 1)
+#' calc_posterior(y = c(14, 23), n = c(100, 100))
 #' 
 #' # One-sample case
-#' calc_posterior(y = 27, n = 100, p0 = 0.2, delta = NULL, seed = 1)
+#' calc_posterior(y = 27, n = 100, p0 = 0.2, delta = NULL)
 #'
 #' @export
 
 calc_posterior <- function(y, n, direction = "greater", p0 = NULL, 
-                           delta = 0, prior = c(0.5,0.5), S = 5000, 
-                           seed = NULL) { 
+                           delta = 0, prior = c(0.5,0.5), S = 5000) { 
   
   if(length(y) != length(n)) 
     stop("y and n must be the same length")
@@ -56,9 +56,7 @@ calc_posterior <- function(y, n, direction = "greater", p0 = NULL,
   
   if(length(y) == 2 & is.null(delta))
     stop("delta must be specified for the two-sample case")
-  
-  set.seed(seed)
-  
+
   if(length(y) == 2) {
     
     rb0 <- stats::rbeta(S, prior[1] + y[1], prior[2] + n[1] - y[1])

### 2021-03-17 R/calc_predictive.R

@@ -22,7 +22,6 @@
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
-#' @param seed set the seed for the random number generation. Default is NULL.
 #' @param N the total planned sample size at the end of the trial, c(N0, N1)
 #' for two-sample case; integer of total planned sample size at end of trial N
 #' for one-sample case
@@ -33,18 +32,19 @@
 #' @return Returns the posterior predictive probability of interest
 #'
 #' @examples
+#' 
+#' set.seed(123)
 #'
 #' # Two-sample case
-#' calc_predictive(y = c(7, 12), n = c(50, 50), seed = 1, N = c(100, 100))
+#' calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100))
 #' 
 #' # One-sample case
-#' calc_predictive(y = 14, n = 50, p0 = 0.2, delta = NULL, seed = 1, N = 100)
+#' calc_predictive(y = 14, n = 50, p0 = 0.2, delta = NULL, N = 100)
 #'
 #' @export
 
 calc_predictive <- function(y, n, direction = "greater", p0 = NULL, 
                             delta = 0, prior = c(0.5,0.5), S = 5000, 
-                            seed = NULL, 
                             N, theta = 0.95) {
   
   if(length(y) != length(n)) 
@@ -65,9 +65,7 @@ calc_predictive <- function(y, n, direction = "greater", p0 = NULL,
   
   if(length(y) != length(N))
     stop("y and N must be the same length")
-  
-  set.seed(seed)
-  
+
   if(length(y) == 2) {
     
     rb0 <- stats::rbeta(S, prior[1] + y[1], prior[2] + n[1] - y[1])
@@ -89,7 +87,7 @@ calc_predictive <- function(y, n, direction = "greater", p0 = NULL,
     
     post <- purrr::map2_dbl(Y0, Y1, ~calc_posterior(
       y = c(.x, .y), n = N,  direction = direction, p0 = p0, delta = delta, 
-      prior = prior, S = S, seed = seed)
+      prior = prior, S = S)
       )
 
   } else if(length(y) == 1) {
@@ -105,8 +103,7 @@ calc_predictive <- function(y, n, direction = "greater", p0 = NULL,
     
     post <- purrr::map_dbl(Y, calc_posterior, n = N, 
                            direction = direction, p0 = p0, 
-                           delta = delta, prior = prior, S = S, 
[…]

### 2021-03-17 R/calibrate_posterior_threshold.R

@@ -1,0 +1,87 @@
+#' Calibrate the posterior probability threshold 
+#'
+#' @description This function is meant to be used in the context of a 
+#' clinical trial with a binary endpoint. For a vector of possible posterior 
+#' decision thresholds, the function simulates many trials and then calculates
+#' the average number of times the posterior probability exceeds a given 
+#' threshold. In a null case, this will result in the type I error at a given
+#' threshold. In an alternative case, this will result in the power at a given
+#' threshold.
+#'
+#' @param prob vector of length two containing the probability of event in 
+#' the standard of care and experimental arm c(p0, p1) for the two-sample case; 
+#' integer of event probability for one-sample case
+#' @param N vector of length two containing the total sample size c(N0, N1) 
+#' for two-sample case; integer of sample size so far N for one-sample case
+#' @param direction "greater" (default) if interest is in p(p1 > p0) and "less" 
+#' if interest is in p(p1 < p0) for two-sample case. For one-sample case, 
+#' "greater" if interest is in p(p > p0) and "less" if interest is in p(p < p0).
+#' @param p0 The target value to compare to in the one-sample case
+#' @param delta clinically meaningful difference between groups. 
+#' Typically 0 (default).
+#' @param prior hyperparameters of prior beta distribution.
+#' Beta(0.5, 0.5) is default
+#' @param S number of samples drawn from the posterior, and number of simulated
+#' trials. Default is 5000
+#' @param theta The target posterior probability thresholds to consider. 
+#' Integer or vector.
+#' 
+#' @return Returns a tibble with the threshold(s) and associated proportion of 
+#' positive trials.
+#' 
+#' @examples 
+#' 
+#' set.seed(123)
+#' 
+#' # One-sample case
+#' calibrate_posterior_threshold(prob = 0.1, N = 50, direction = "greater", 
+#' p0 = 0.1, delta = NULL, prior = c(0.5,0.5), S = 5000, theta = c(0.9, 0.95))
+#' 
+#' # Two-sample case
+#' calibrate_posterior_threshold(prob = c(0.1, 0.1), N = c(50, 50),
+#' direction = "greater", p0 = NULL, delta = 0, prior = c(0.5,0.5), 
+#' S = 5000, theta = c(0.9, 0.95))
+#'
+#' @export
+
+calibrate_posterior_threshold <- function(prob, N,
+                                          direction = "greater", p0 = NULL, 
[…]

### 2021-03-17 R/sim_single_trial.R

@@ -13,8 +13,12 @@
 #' @param n matrix containing the total number of patients accrued so far at 
 #' each interim look in the standard of care (column 1) and experimental 
 #' (column 2) arms for two-sample case; vector of sample size accrued so far 
-#' at each interim look for one-sample case. If only a single look will be done
-#' at the end of the trial, this can be an integer specifying the total sample
+#' at each interim look for one-sample case. The last value should be equal to 
+#' the total sample size at the end of the trial. 
+#' If only a single look will be done
+#' at the end of the trial, this can be a vector specifying the total sample 
+#' size c(N0, N1) for the two-sample case or an integer specifying the total 
+#' sample size N for the one-sample case
 #' size at the end of the trial N.
 #' @param direction "greater" (default) if interest is in P(p1 > p0) and "less" 
 #' if interest is in P(p1 < p0) for two-sample case. For one-sample case, 
@@ -25,22 +29,35 @@
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
-#' @param seed set the seed for the random number generation. Default is NULL.
 #' @param N the total planned sample size at the end of the trial, c(N0, N1)
 #' for two-sample case; integer of total planned sample size at end of trial N
 #' for one-sample case
 #' @param theta The target posterior probability. e.g. Efficacy decision if 
 #' P(p1 > p0) > theta for the two-sample case with greater direction. 
-#' Default is 0.95.
+#' Default is 0.95. Can be a vector if interest is in selecting from among a 
+#' variety of thresholds.
 #' 
-#' @return Returns a matrix of number of responses, sample size, and posterior
-#' predictive probability at each interim look
+#' @return Returns a tibble with theta, number of responses, sample size, and 
+#' posterior probability, and posterior predictive probability at each interim 
+#' look
+#' 
+#' @examples
+#' 
+#' # One-sample case
+#' set.seed(123)
+#' sim_single_trial(prob = 0.3, n = seq(5, 50, 5), direction = "greater", p = 0.1,
+#' delta = NULL, prior = c(0.5,0.5), S = 5000, N = 50, theta = 0.95)
+#' 
+#' # Two-sample case
+#' set.seed(123)
+#' sim_single_trial(prob = c(0.1, 0.3), n = cbind(seq(5, 50, 5), seq(5, 50, 5)), 
[…]

### 2021-03-17 R/sim_single_trial.R
just committing last week's updates

@@ -114,8 +114,8 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
                              S = S, N = N, theta = ..5)
     )
     
-    res <- arrange(
-      select(
+    res <- dplyr::arrange(
+      dplyr::select(
         tibble::add_column(crossargs, pp = rep(pp, length(theta)), ppp = ppp),
         theta, everything()
       ),
@@ -155,8 +155,8 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
                              S = S, N = N, theta = ..3)
     )
     
-    res <- arrange(
-      select(
+    res <- dplyr::arrange(
+      dplyr::select(
         tibble::add_column(crossargs, pp = rep(pp, length(theta)), ppp = ppp),
         theta, everything()
       ),

### 2021-03-19 R/globals.R
adding calibrate_pp_ppp_threshold function

@@ -1,0 +1,1 @@
+utils::globalVariables(c("n0", "n1", "positive"))

### 2021-03-23 R/calibrate_posterior_threshold.R
cleaning up one-sample vignette

@@ -26,8 +26,8 @@
 #' @param theta The target posterior probability thresholds to consider. 
 #' Integer or vector.
 #' 
-#' @return Returns a tibble with the threshold(s) and associated proportion of 
-#' positive trials.
+#' @return Returns a tibble with the posterior probability threshold(s) and 
+#' associated proportion of positive trials.
 #' 
 #' @examples 
 #' 
@@ -80,8 +80,8 @@ calibrate_posterior_threshold <- function(prob, N,
   proportion_positive <- purrr::map_dbl(theta, ~mean(posts > .x))
   
   return(tibble::tibble(
-    theta = theta, 
-    proportion_positive = proportion_positive)
+    pp_threshold = theta, 
+    prop_pos = proportion_positive)
   )
   
 }

### 2021-03-23 R/calibrate_pp_ppp_threshold.R

@@ -255,7 +255,9 @@ calibrate_pp_ppp_threshold <- function(prob, n,
                           eval_thresh(x, pp_threshold, ppp_threshold,
                                       direction = direction, p0 = p0, 
                                       delta = delta, prior = prior, 
-                                      S = S, N = N)))
+                                      S = S, N = N), 
+                        .options = furrr::furrr_options(seed = TRUE)), 
+      .options = furrr::furrr_options(seed = TRUE))
   
   res_df <- 
     dplyr::bind_rows(
@@ -265,19 +267,25 @@ calibrate_pp_ppp_threshold <- function(prob, n,
   
   if(length(prob) == 2) {
     res_summary <- 
-      dplyr::summarize(
-        dplyr::group_by(res_df, pp_threshold, ppp_threshold),
-        mean_n0 = mean(n0),
-        mean_n1 = mean(n1),
-        prop_pos = mean(positive),
-        prop_stopped = mean(sum(n0, n1) < sum(N)))
+      dplyr::ungroup(
+        dplyr::summarize(
+          dplyr::group_by(res_df, pp_threshold, ppp_threshold),
+          mean_n0 = mean(n0),
+          mean_n1 = mean(n1),
+          prop_pos = mean(positive),
+          prop_stopped = mean(sum(n0, n1) < sum(N))
+          )
+        )
   } else if(length(prob) == 1) {
     res_summary <- 
-      dplyr::summarize(
-        dplyr::group_by(res_df, pp_threshold, ppp_threshold),
-        mean_n1 = mean(n1),
-        prop_pos = mean(positive),
-        prop_stopped = mean(n1 < N))
+      dplyr::ungroup(
+        dplyr::summarize(
+          dplyr::group_by(res_df, pp_threshold, ppp_threshold),
+          mean_n1 = mean(n1),
+          prop_pos = mean(positive),
+          prop_stopped = mean(n1 < N)
+          )
+        )
   }
   return(res_summary)
 }

### 2021-03-23 R/sim_single_trial.R

@@ -36,8 +36,9 @@
 #' Default is 0.95. Can be a vector if interest is in selecting from among a 
 #' variety of thresholds.
 #' 
-#' @return Returns a tibble with theta, number of responses, sample size, and 
-#' posterior probability, and posterior predictive probability at each interim 
+#' @return Returns a tibble with pp_threshold (i.e. theta, the target posterior 
+#' probability), number of responses, sample size,  
+#' posterior probability, and posterior predictive probability at each 
 #' look
 #' 
 #' @examples
@@ -103,7 +104,7 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
                                 y1 = rep(y1, length(theta)), 
                                 n0 = rep(n[, 1], length(theta)),
                                 n1 = rep(n[, 2], length(theta)),
-                                theta = rep(theta, each = length(pp)))
+                                pp_threshold = rep(theta, each = length(pp)))
     
     ppp <- purrr::pmap_dbl(crossargs, 
                            ~calc_predictive(
@@ -115,10 +116,11 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
     
     res <- dplyr::arrange(
       dplyr::select(
-        tibble::add_column(crossargs, pp = rep(pp, length(theta)), ppp = ppp),
-        theta, dplyr::everything()
+        tibble::add_column(crossargs, pp = rep(pp, length(theta)), 
+                           ppp = ppp),
+        pp_threshold, dplyr::everything()
       ),
-      theta
+      pp_threshold
     )
     
   } else if(length(prob) == 1) {
@@ -143,8 +145,8 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
     )
     
     crossargs <- tibble::tibble(y1 = rep(y1, length(theta)), 
-                                n = rep(n, length(theta)),
-                                theta = rep(theta, each = length(pp)))
+                                n1 = rep(n, length(theta)),
+                                pp_threshold = rep(theta, each = length(pp)))
     
     ppp <- purrr::pmap_dbl(crossargs, 
                            ~calc_predictive(
@@ -157,9 +159,9 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
     res <- dplyr::arrange(
       dplyr::select(
         tibble::add_column(crossargs, pp = rep(pp, length(theta)), ppp = ppp),
-        theta, dplyr::everything()
[…]

### 2021-03-26 R/calibrate_pp_ppp_threshold.R
removing the future::plan to test again on server

@@ -243,8 +243,7 @@ calibrate_pp_ppp_threshold <- function(prob_null, prob_alt, n,
                                        delta = 0, prior = c(0.5, 0.5), 
                                        S = 5000, N, nsim = 1000,
                                        pp_threshold, ppp_threshold) {
-  future::plan("future::multisession")
-  
+
   sim_dat_null <- 
     purrr::map(1:nsim, ~sim_dat1(prob = prob_null, n = n))
   

### 2021-03-31 R/globals.R
update calibrate_pp_ppp_threshold and vignette

@@ -1,1 +1,4 @@
-utils::globalVariables(c("n0", "n1", "positive", "pp_threshold"))
+utils::globalVariables(c("n0", "n1", "positive", "pp_threshold",
+                         "ppp", "y0", "y1", "n0_null", "n1_null",
+                         "positive_null", "n0_alt", "n1_alt",
+                         "positive_alt"))

### 2021-04-06 R/calibrate_thresholds.R
adding generics

@@ -209,20 +209,24 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold,
 #' @param ppp_threshold the posterior probability threshold of interest for 
 #' futility monitoring
 #' 
-#' @return A tibble containing the posterior probability threshold 
+#' @return A list containing a 
+#' 1) a tibble 'res_summary' containing the posterior probability threshold 
 #' (pp_threshold), the predictive probability threshold (ppp_threshold), 
 #' the mean sample size (mean_n0 and mean_n1 for two-sample case; mean_n1 for
 #' one-sample case), and the proportion of positive trials.
+#' 2) 'call_list' containing the original function call
+#' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the 
+#' original function call
 #' 
 #' The proportion of positive trials will be a measure of the type I error 
 #' for a null setting, and a measure of the power in the alternative setting.
 #' 
 #' @examples 
 #' 
-#' set.seed(123)
-#' 
 #' # One-sample case
 #' \dontrun{
+#' set.seed(123)
+#' 
 #' calibrate_thresholds(prob_null = 0.1, prob_alt = 0.3, 
 #' n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL, 
 #' prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000, 
@@ -379,149 +383,3 @@ calibrate_thresholds <- function(prob_null, prob_alt, n,
 #                      prior = c(0.5, 0.5), S = 500, N = 25, nsim = 10, 
 #                      pp_threshold = c(0.9, 0.95, 0.96, 0.98), 
 #                      ppp_threshold = seq(0.05, 0.2, 0.05))
-
-
-#' Print method for \code{calibrate_thresholds} objects
-#' 
-#' @description By default prints only the res_summary table from an object of 
-#' class 'calibrate_thresholds'
-#' 
-#' @param x an object of class 'calibrate_thresholds', usually returned by the 
-#' \code{calibrate_thresholds} function
-#' 
-#' @return 
-#' Returns a tibble
-#' 
-#' @export
-
-print.calibrate_thresholds <- function(x, ...) {
-  print(x$res_summary)
-}
-
-
-#' Plot method for \code{calibrate_thresholds} objects
-#' 
-#' @param x an object of class 'calibrate_thresholds', usually returned by the 
-#' \code{calibrate_thresholds} function
-#' 
[…]

### 2021-04-06 R/globals.R

@@ -1,4 +1,6 @@
 utils::globalVariables(c("n0", "n1", "positive", "pp_threshold",
                          "ppp", "y0", "y1", "n0_null", "n1_null",
                          "positive_null", "n0_alt", "n1_alt",
-                         "positive_alt"))
+                         "positive_alt",
+                         "prop_pos_null", "prop_pos_alt", "prop_stopped_null",
+                         "Youden's index", "mean_n1_null", "design"))

### 2021-04-06 R/optimize_design.R

@@ -1,0 +1,90 @@
+#' Function to setup usage of `optimize_design.calibrate_thresholds`
+#' 
+#' @param x an object of class 'calibrate_thresholds', usually returned by the 
+#' \code{calibrate_thresholds} function
+#' @param type1_range a vector specifying the minimum and maximum acceptable
+#' type I error. Specify NULL to return the full range of resulting type I 
+#' error. Defaults to c(0.05, 0.1)
+#' @param minimum_power a numeric between 0 and 1 specifying the minimum 
+#' acceptable power. Specify NULL to return the full range of resulting power.
+#' Defaults to 0.8.
+#' @param ... ignored
+#' 
+#' @export
+
+optimize_design <- function(x, 
+                            type1_range = c(0.05, 0.1), 
+                            minimum_power = 0.8, ...) {
+  UseMethod("optimize_design")
+}
+
+
+#' Custom optimization method for \code{calibrate_thresholds} objects
+#' 
+#' @description Determines the optimal designs based on a variety of criteria
+#' 
+#' @param x an object of class 'calibrate_thresholds', usually returned by the 
+#' \code{calibrate_thresholds} function
+#' @param type1_range a vector specifying the minimum and maximum acceptable
+#' type I error. Specify NULL to return the full range of resulting type I 
+#' error. Defaults to c(0.05, 0.1)
+#' @param minimum_power a numeric between 0 and 1 specifying the minimum 
+#' acceptable power. Specify NULL to return the full range of resulting power.
+#' Defaults to 0.8.
+#' @param ... ignored
+#' 
+#' @examples 
+#' 
+#' \dontrun{
+#' set.seed(123)
+#' 
+#' cal_tbl <- calibrate_thresholds(prob_null = 0.1, prob_alt = 0.3, 
+#' n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL, 
+#' prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000, 
+#' pp_threshold = c(0.9, 0.95, 0.96, 0.98), 
+#' ppp_threshold = seq(0.05, 0.2, 0.05))
+#' 
+#' optimize_design(cal_tbl)
+#' }
+#' 
+#' @export
+
+optimize_design.calibrate_thresholds <- function(x, 
+                                                 type1_range = c(0.05, 0.1), 
+                                                 minimum_power = 0.8, ...) {
+  
+  if(any(class(x) =="calibrate_thresholds") == FALSE)
[…]

### 2021-04-06 R/plot.R

@@ -1,0 +1,122 @@
+#' Plot method for \code{calibrate_thresholds} objects
+#' 
+#' @description Returns two interactive \code{plotly} plots to compare results
+#' from various designs generated from a call to \code{calibrate_thresholds}
+#' based on various criteria, and to assist in selecting an optimal design. 
+#' 
+#' @param x an object of class 'calibrate_thresholds', usually returned by the 
+#' \code{calibrate_thresholds} function
+#' @param type1_range a vector specifying the minimum and maximum acceptable
+#' type I error. Specify NULL to return the full range of resulting type I 
+#' error. Defaults to c(0.05, 0.1)
+#' @param minimum_power a numeric between 0 and 1 specifying the minimum 
+#' acceptable power. Specify NULL to return the full range of resulting power.
+#' Defaults to 0.8.
+#' @param ... unused
+#' 
+#' @export
+
+plot.calibrate_thresholds <- function(x, 
+                                      type1_range = c(0.05, 0.1), 
+                                      minimum_power = 0.8, ...) {
+  
+  if(any(class(x) =="calibrate_thresholds") == FALSE)
+    stop("x must be class 'calibrate_thresholds', usually an object returned from a call to the function ppseq::calibrate_thresholds()")
+  
+  if(length(type1_range) != 2 & !is.null(type1_range))
+    stop("type1_range must be a vector of length 2 defining the range of acceptable type I error, or can be set equal to NULL to return all values of type I error")
+  
+  if(length(minimum_power) != 1 & !is.null(minimum_power))
+    stop("minimum_power must be a numeric value of length 1 defining the minimum acceptable power, or can be set equal to NULL to return all values of power")
+  
+  if(!is.numeric(minimum_power) & !is.null(minimum_power))
+    stop("minimum_power must be a numeric value of length 1 or NULL")
+  
+  if(!is.numeric(type1_range) & !is.null(type1_range))
+    stop("type1_range must be a numeric vector of length 2 or NULL")
+  
+  plot_x <- 
+    dplyr::mutate(
+      dplyr::filter(x$res_summary, 
+                    prop_pos_null >= type1_range[1] & 
+                      prop_pos_null <= type1_range[2] &
[…]

### 2021-04-06 R/print.R

@@ -1,0 +1,71 @@
+#' Print method for \code{calibrate_thresholds} objects
+#' 
+#' @description By default prints only the res_summary table from an object of 
+#' class 'calibrate_thresholds'. The table can be limited to a range of type 1 
+#' error and a minimum value of power using the arguments 'type1_range' and
+#' 'minimum_power' respectively.
+#' 
+#' @param x an object of class 'calibrate_thresholds', usually returned by the 
+#' \code{calibrate_thresholds} function
+#' @param type1_range a vector specifying the minimum and maximum acceptable
+#' type I error. Specify NULL to return the full range of resulting type I 
+#' error. Defaults to c(0.05, 0.1)
+#' @param minimum_power a numeric between 0 and 1 specifying the minimum 
+#' acceptable power. Specify NULL to return the full range of resulting power.
+#' Defaults to 0.8.
+#' @param ... ignored
+#' 
+#' @return 
+#' Returns a tibble
+#'
+#' @examples 
+#' 
+#' \dontrun{
+#' set.seed(123)
+#' 
+#' cal_tbl <- calibrate_thresholds(prob_null = 0.1, prob_alt = 0.3, 
+#' n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL, 
+#' prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000, 
+#' pp_threshold = c(0.9, 0.95, 0.96, 0.98), 
+#' ppp_threshold = seq(0.05, 0.2, 0.05))
+#' 
+#' print(cal_tbl)
+#' print(cal_tbl, type1_range = NULL, minimum_power = NULL)
+#' }
+#' 
+#' @export
+
+print.calibrate_thresholds <- function(x, 
+                                       type1_range = c(0.05, 0.1), 
+                                       minimum_power = 0.8, 
+                                       ...) {
+  
+  if(any(class(x) =="calibrate_thresholds") == FALSE)
+    stop("x must be class 'calibrate_thresholds', usually an object returned from a call to the function ppseq::calibrate_thresholds()")
+  
+  if(length(type1_range) != 2 & !is.null(type1_range))
+    stop("type1_range must be a vector of length 2 defining the range of acceptable type I error, or can be set equal to NULL to return all values of type I error")
+  
+  if(length(minimum_power) != 1 & !is.null(minimum_power))
[…]

### 2021-04-13 README.Rmd
updated vignette draft
cleaning up one-sample vignette
cleaning up one-sample and starting two-sample vignette

--- 2021-02-17 README.Rmd
@@
 [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-[![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=main)](https://travis-ci.com/zabore/ppseq)
-[![codecov](https://codecov.io/gh/zabore/ppseq/branch/master/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq)
-[![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
+[![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=main)](https://travis-ci.com/zabore/ppseq) [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
+[![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
 <!-- badges: end -->
 
-The goal of ppseq is to ...
+The goal of ppseq is to provide functions that aid in designing clinical trials with sequential posterior probability monitoring incorporated.
 
@@
 
-``` r
+```r
 remotes::install_github("zabore/ppseq")
 ```
-
-## Example
-
-This is a basic example which shows you how to solve a common problem:
-
-``` r
-library(ppseq)
-## basic example code
-```

### 2021-04-13 R/cal_tbl-data.R

@@ -1,0 +1,26 @@
+#' Output from a call to \code{calibrate_threshold}
+#'
+#' This .rda file contains output from a call to \code{calibrate_threshold}. 
+#' See the vignette titled "One-sample expansion cohort" for a description of
+#' the input parameters used, or run \code{cal_tbl$calibrate_thresholds_inputs}
+#' to see a list of the original function inputs. For use in testing global 
+#' functions and in vignettes.
+#'
+#' @docType data
+#'
+#' @usage data(cal_tbl)
+#'
+#' @format A list containing a 
+#' 1) a tibble 'res_summary' containing the posterior probability threshold 
+#' (pp_threshold), the predictive probability threshold (ppp_threshold), 
+#' the ceiling of the mean sample size (mean_n0 and mean_n1 for two-sample case; 
+#' mean_n1 for one-sample case), and the proportion of positive trials.
+#' 2) 'call_list' containing the original function call
+#' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the 
+#' original function call
+#' 4) 'simon_res' gives results from a Simon's two-stage minimax design with
+#' unacceptable rate 0.1, acceptable rate 0.3, and theoretical type I error of
+#' 0.05 and power of 0.8
+#'
+#' @keywords datasets
+"cal_tbl"

### 2021-04-13 R/calibrate_thresholds.R

@@ -212,8 +212,8 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold,
 #' @return A list containing a 
 #' 1) a tibble 'res_summary' containing the posterior probability threshold 
 #' (pp_threshold), the predictive probability threshold (ppp_threshold), 
-#' the mean sample size (mean_n0 and mean_n1 for two-sample case; mean_n1 for
-#' one-sample case), and the proportion of positive trials.
+#' the ceiling of the mean sample size (mean_n0 and mean_n1 for two-sample case; 
+#' mean_n1 for one-sample case), and the proportion of positive trials.
 #' 2) 'call_list' containing the original function call
 #' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the 
 #' original function call
@@ -312,12 +312,12 @@ calibrate_thresholds <- function(prob_null, prob_alt, n,
       dplyr::ungroup(
         dplyr::summarize(
           dplyr::group_by(res_df, pp_threshold, ppp_threshold),
-          mean_n0_null = mean(n0_null),
-          mean_n1_null = mean(n1_null),
+          mean_n0_null = ceiling(mean(n0_null)),
+          mean_n1_null = ceiling(mean(n1_null)),
           prop_pos_null = mean(positive_null),
           prop_stopped_null = mean(sum(n0_null, n1_null) < sum(N)),
-          mean_n0_alt = mean(n0_alt),
-          mean_n1_alt = mean(n1_alt),
+          mean_n0_alt = ceiling(mean(n0_alt)),
+          mean_n1_alt = ceiling(mean(n1_alt)),
           prop_pos_alt = mean(positive_alt),
           prop_stopped_alt = mean(sum(n0_alt, n1_alt) < sum(N))
           )
@@ -337,10 +337,10 @@ calibrate_thresholds <- function(prob_null, prob_alt, n,
       dplyr::ungroup(
         dplyr::summarize(
           dplyr::group_by(res_df, pp_threshold, ppp_threshold),
-          mean_n1_null = mean(n1_null),
+          mean_n1_null = ceiling(mean(n1_null)),
           prop_pos_null = mean(positive_null),
           prop_stopped_null = mean(n1_null < N),
-          mean_n1_alt = mean(n1_alt),
+          mean_n1_alt = ceiling(mean(n1_alt)),
           prop_pos_alt = mean(positive_alt),
           prop_stopped_alt = mean(n1_alt < N)
           )

### 2021-04-13 R/globals.R

@@ -3,4 +3,9 @@ utils::globalVariables(c("n0", "n1", "positive", "pp_threshold",
                          "positive_null", "n0_alt", "n1_alt",
                          "positive_alt",
                          "prop_pos_null", "prop_pos_alt", "prop_stopped_null",
-                         "Youden's index", "mean_n1_null", "design"))
+                         "Youden's index", "mean_n1_null", "design",
+                         "ppp_threshold", "mean_n1_alt", "n_dist_metric",
+                         "Type I error", "Power", "Design", 
+                         "Average N under the null", 
+                         "Average N under the alternative",
+                         "Distance to min(N under null) and max(N under alt)"))

### 2021-04-13 R/plot.R

@@ -36,84 +36,61 @@ plot.calibrate_thresholds <- function(x,
     stop("type1_range must be a numeric vector of length 2 or NULL")
   
   plot_x <- 
-    dplyr::mutate(
-      dplyr::filter(x$res_summary, 
-                    prop_pos_null >= type1_range[1] & 
-                      prop_pos_null <= type1_range[2] &
-                      prop_pos_alt >= minimum_power),
-      `Youden's index` = prop_pos_alt + prop_stopped_null - 1
+    dplyr::rename(
+      dplyr::mutate(
+        dplyr::filter(
+          x$res_summary, 
+          prop_pos_null >= type1_range[1] & 
+            prop_pos_null <= type1_range[2] &
+            prop_pos_alt >= minimum_power
+        ),
+        `Youden's index` = prop_pos_alt + prop_stopped_null - 1,
+        Design = paste0("Posterior threshold ", pp_threshold, 
+                        " and predictive threshold ", ppp_threshold),
+        n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 + 
+                           (mean_n1_alt - max(mean_n1_alt))^2)^(1/2)
+      ),
+      `Type I error` = prop_pos_null,
+      Power = prop_pos_alt,
+      `Average N under the null` = mean_n1_null,
+      `Average N under the alternative` = mean_n1_alt,
+      `Distance to min(N under null) and max(N under alt)` = n_dist_metric
     )
   
   p1 <- 
-    ggplot2::ggplot(plot_x, ggplot2::aes(x = prop_stopped_null, 
-                                         y = prop_pos_alt, 
-                                         color = `Youden's index`)) + 
-    ggplot2::geom_point() + 
-    ggplot2::scale_x_reverse(limits = c(1, 0)) +
-    ggplot2::ylim(0, 1) + 
-    ggplot2::labs(
-      x = "Proportion stopped under the null (Specificity)",
-      y = "Proportion positive under alternative (Sensitiivity)"
-    ) 
-  
-  simon_lower <- clinfun::ph2simon(pu = x$inputs$prob_null, 
-                                   pa = x$inputs$prob_alt, 
-                                   ep1 = type1_range[1], 
-                                   ep2 = 1 - minimum_power)
-  
-  xopt_lower <- simon_lower$out[
[…]

### 2021-04-13 tests/testthat.R

@@ -1,0 +1,4 @@
+library(testthat)
+library(ppseq)
+
+test_check("ppseq")

### 2021-04-14 tests/testthat/test-calc_predictive.R
adding first two tests

@@ -1,0 +1,26 @@
+test_that("no errors or warnings with correct use", {
+  expect_error(calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100),
+                               S = 100), NA)
+  expect_warning(calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100),
+                                 S = 100), NA)
+  
+  expect_error(calc_predictive(y = 14, n = 50, p0 = 0.2, delta = NULL, N = 100, 
+                               S = 100), NA)
+  expect_warning(calc_predictive(y = 14, n = 50, p0 = 0.2, delta = NULL, 
+                                 N = 100, S = 100), NA)
+})
+
+test_that("errors when expected", {
+  expect_error(calc_predictive(y = c(7, 12), n = 50, N = c(100, 100), S = 100), 
+               "*")
+  expect_error(calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100),
+                               p0 = 0.1, delta = 0, S = 100), "*")
+  expect_error(calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100),
+                               S = 100, direction = "south"), "*")
+  expect_error(calc_predictive(y = 14, n = 50, delta = NULL, N = 100, 
+                               S = 100), "*")
+  expect_error(calc_predictive(y = c(7, 12), n = c(50, 50), N = c(100, 100),
+               S = 100, delta = NULL), "*")
+  expect_error(calc_predictive(y = c(7, 12), n = c(50, 50), N = 100,
+                               S = 100), "*")
+})

### 2021-04-15 R/optimize_design.R
updating optimize_design and plot to handle two-sample case

@@ -70,21 +70,40 @@ optimize_design.calibrate_thresholds <- function(x,
   
   options(tibble.width = Inf)
   
-  plot_x <- 
-    dplyr::mutate(
-      dplyr::filter(x$res_summary, 
-                    prop_pos_null >= type1_range[1] & 
-                      prop_pos_null <= type1_range[2] &
-                      prop_pos_alt >= minimum_power),
-      `Youden's index` = prop_pos_alt + prop_stopped_null - 1
-    )
+  if(length(x$inputs$prob_null) == 2) {
+    opt_x <- 
+      dplyr::mutate(
+        dplyr::filter(x$res_summary, 
+                      prop_pos_null >= type1_range[1] & 
+                        prop_pos_null <= type1_range[2] &
+                        prop_pos_alt >= minimum_power),
+        mean_n_null = (mean_n0_null + mean_n1_null) / 2,
+        mean_n_alt = (mean_n0_alt + mean_n1_alt) / 2,
+        ab_dist_metric = ((prop_pos_null - 0)^2 + 
+                            (prop_pos_alt - 1)^2)^(1/2),
+        n_dist_metric = ((mean_n_null - min(mean_n_null))^2 + 
+                           (mean_n_alt - max(mean_n_alt))^2)^(1/2)
+      )
+    
+  } else if(length(x$inputs$prob_null) == 1) {
+    opt_x <- 
+      dplyr::mutate(
+        dplyr::filter(x$res_summary, 
+                      prop_pos_null >= type1_range[1] & 
+                        prop_pos_null <= type1_range[2] &
+                        prop_pos_alt >= minimum_power),
+        ab_dist_metric = ((prop_pos_null - 0)^2 + 
+                            (prop_pos_alt - 1)^2)^(1/2),
+        n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 + 
+                           (mean_n1_alt - max(mean_n1_alt))^2)^(1/2)
+      )
+  }
+  
   
   list(
-    "The design that maximizes specificity (i.e. the proportion stopped under the null) is:" = 
-      plot_x[which.max(plot_x$prop_stopped_null), ],
-    "The design that maximizes sensitivity (i.e. the proportion positive under the alternative) is:" = 
-      plot_x[which.max(plot_x$prop_pos_alt), ],
-    "The design that maximizes Youden's index is:" =
-      plot_x[which.max(plot_x$`Youden's index`), ]
[…]

### 2021-04-15 R/plot.R

@@ -35,36 +35,68 @@ plot.calibrate_thresholds <- function(x,
   if(!is.numeric(type1_range) & !is.null(type1_range))
     stop("type1_range must be a numeric vector of length 2 or NULL")
   
-  plot_x <- 
-    dplyr::rename(
-      dplyr::mutate(
-        dplyr::filter(
-          x$res_summary, 
-          prop_pos_null >= type1_range[1] & 
-            prop_pos_null <= type1_range[2] &
-            prop_pos_alt >= minimum_power
+  if(length(x$inputs$prob_null) == 2) {
+    plot_x <- 
+      dplyr::rename(
+        dplyr::mutate(
+          dplyr::filter(
+            x$res_summary, 
+            prop_pos_null >= type1_range[1] & 
+              prop_pos_null <= type1_range[2] &
+              prop_pos_alt >= minimum_power
+          ),
+          Design = paste0("Posterior threshold ", pp_threshold, 
+                          " and predictive threshold ", ppp_threshold),
+          ab_dist_metric = ((prop_pos_null - 0)^2 + 
+                              (prop_pos_alt - 1)^2)^(1/2),
+          mean_n_null = (mean_n0_null + mean_n1_null) / 2,
+          mean_n_alt = (mean_n0_alt + mean_n1_alt) / 2,
+          n_dist_metric = ((mean_n_null - min(mean_n_null))^2 + 
+                             (mean_n_alt - max(mean_n_alt))^2)^(1/2)
         ),
-        `Youden's index` = prop_pos_alt + prop_stopped_null - 1,
-        Design = paste0("Posterior threshold ", pp_threshold, 
-                        " and predictive threshold ", ppp_threshold),
-        n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 + 
-                           (mean_n1_alt - max(mean_n1_alt))^2)^(1/2)
-      ),
-      `Type I error` = prop_pos_null,
-      Power = prop_pos_alt,
-      `Average N under the null` = mean_n1_null,
-      `Average N under the alternative` = mean_n1_alt,
-      `Distance to min(N under null) and max(N under alt)` = n_dist_metric
-    )
+        `Type I error` = prop_pos_null,
+        Power = prop_pos_alt,
+        `Average N under the null` = mean_n_null,
+        `Average N under the alternative` = mean_n_alt,
+        `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
[…]

### 2021-04-26 R/globals.R
updating the plot and optimize functions

@@ -8,4 +8,6 @@ utils::globalVariables(c("n0", "n1", "positive", "pp_threshold",
                          "Type I error", "Power", "Design", 
                          "Average N under the null", 
                          "Average N under the alternative",
-                         "Distance to min(N under null) and max(N under alt)"))
+                         "Distance to min(N under null) and max(N under alt)",
+                         "mean_n0_null", "mean_n0_alt", "mean_n_null",
+                         "mean_n_alt", "ab_dist_metric", "Distance to (0, 1)"))

### 2021-04-26 R/optimize_design.R

@@ -72,38 +72,70 @@ optimize_design.calibrate_thresholds <- function(x,
   
   if(length(x$inputs$prob_null) == 2) {
     opt_x <- 
-      dplyr::mutate(
-        dplyr::filter(x$res_summary, 
-                      prop_pos_null >= type1_range[1] & 
-                        prop_pos_null <= type1_range[2] &
-                        prop_pos_alt >= minimum_power),
-        mean_n_null = (mean_n0_null + mean_n1_null) / 2,
-        mean_n_alt = (mean_n0_alt + mean_n1_alt) / 2,
-        ab_dist_metric = ((prop_pos_null - 0)^2 + 
-                            (prop_pos_alt - 1)^2)^(1/2),
-        n_dist_metric = ((mean_n_null - min(mean_n_null))^2 + 
-                           (mean_n_alt - max(mean_n_alt))^2)^(1/2)
-      )
+      dplyr::rename(
+        dplyr::mutate(
+          dplyr::filter(x$res_summary, 
+                        prop_pos_null >= type1_range[1] & 
+                          prop_pos_null <= type1_range[2] &
+                          prop_pos_alt >= minimum_power),
+          mean_n_null = (mean_n0_null + mean_n1_null) / 2,
+          mean_n_alt = (mean_n0_alt + mean_n1_alt) / 2,
+          ab_dist_metric = ((prop_pos_null - 0)^2 + 
+                              (prop_pos_alt - 1)^2)^(1/2),
+          n_dist_metric = ((mean_n_null - min(mean_n_null))^2 + 
+                             (mean_n_alt - max(mean_n_alt))^2)^(1/2)
+          ),
+        `Type I error` = prop_pos_null,
+        Power = prop_pos_alt,
+        `Average N under the null` = mean_n_null,
+        `Average N under the alternative` = mean_n_alt,
+        `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
+        `Distance to (0, 1)` = ab_dist_metric)
     
   } else if(length(x$inputs$prob_null) == 1) {
     opt_x <- 
-      dplyr::mutate(
-        dplyr::filter(x$res_summary, 
-                      prop_pos_null >= type1_range[1] & 
-                        prop_pos_null <= type1_range[2] &
-                        prop_pos_alt >= minimum_power),
-        ab_dist_metric = ((prop_pos_null - 0)^2 + 
-                            (prop_pos_alt - 1)^2)^(1/2),
[…]

### 2021-04-26 R/plot.R

@@ -88,15 +88,31 @@ plot.calibrate_thresholds <- function(x,
       )
   }
   
+  plot_ab <- 
+    dplyr::slice(
+      dplyr::group_by(
+        dplyr::arrange(plot_x, 
+                       `Distance to (0, 1)`, -pp_threshold, -ppp_threshold),
+        `Distance to (0, 1)`), 
+      1)
+  
+  plot_nn <-
+    dplyr::slice(
+      dplyr::group_by(
+        dplyr::arrange(plot_x, 
+                       `Distance to min(N under null) and max(N under alt)`, 
+                       -pp_threshold, -ppp_threshold),
+        `Distance to min(N under null) and max(N under alt)`),
+      1)
   
   p1 <- 
-    ggplot2::ggplot(plot_x, 
+    ggplot2::ggplot(plot_ab, 
                     ggplot2::aes(
                       x = `Type I error`, 
                       y = Power, 
                       color = `Distance to (0, 1)`,
                       Design = Design)) + 
-    ggplot2::geom_jitter(width = 0.01, height = 0.01) +
+    ggplot2::geom_point() +
     ggplot2::ylim(0, 1) + 
     ggplot2::xlim(0, 1) +
     ggplot2::labs(
@@ -107,13 +123,13 @@ plot.calibrate_thresholds <- function(x,
     ggplot2::theme_bw()
   
   p2 <- 
-    ggplot2::ggplot(plot_x, 
+    ggplot2::ggplot(plot_nn, 
                     ggplot2::aes(
                       x = `Average N under the null`, 
                       y = `Average N under the alternative`,
                       color = `Distance to min(N under null) and max(N under alt)`,
                       Design = Design)) + 
-    ggplot2::geom_jitter(width = 0.5, height = 0.5) + 
+    ggplot2::geom_point() +
     ggplot2::ylim(min(plot_x$`Average N under the null`), 
                   max(plot_x$`Average N under the alternative`)) +
     ggplot2::xlim(min(plot_x$`Average N under the null`), 
@@ -126,8 +142,6 @@ plot.calibrate_thresholds <- function(x,
     ggplot2::scale_color_viridis_c() +
     ggplot2::theme_bw()
   
-  # gridExtra::grid.arrange(p1, p2, ncol = 2)
-  plotly::ggplotly(p1)
-  plotly::ggplotly(p2)
+  list(plotly::ggplotly(p1), plotly::ggplotly(p2))
 
 }

### 2021-05-26 R/calc_decision_rules.R
sim_single_trial parallelized.

@@ -28,8 +28,8 @@
 #' 
 #' @return Returns a tibble with n at each look and r. Stop the trial at that
 #' look if the number of observed responses is <=r. At the end of the trial,
-#' the treatment is considered promising if the number of observed responses is 
-#' >r.
+#' the treatment is considered promising if the number of observed responses 
+#' is >r.
 #'
 #' @examples
 #' 

### 2021-05-26 R/sim_single_trial.R

@@ -106,13 +106,16 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
                                 n1 = rep(n[, 2], length(theta)),
                                 pp_threshold = rep(theta, each = length(pp)))
     
-    ppp <- purrr::pmap_dbl(crossargs, 
-                           ~calc_predictive(
-                             y = c(..1, ..2), 
-                             n = c(..3, ..4), 
-                             p0 = p0, delta = delta, prior = prior,
-                             S = S, N = N, theta = ..5)
-    )
+    ppp <- furrr::future_pmap_dbl(
+      crossargs, 
+      ~calc_predictive(
+        y = c(..1, ..2), 
+        n = c(..3, ..4), 
+        p0 = p0, delta = delta, prior = prior,
+        S = S, N = N, theta = ..5),
+      .options = furrr_options(
+        seed = TRUE,
+        chunk_size = ceiling(floor(nrow(crossargs)/future::nbrOfWorkers()))))
     
     res <- dplyr::arrange(
       dplyr::select(
@@ -148,14 +151,17 @@ sim_single_trial <- function(prob, n, direction = "greater", p0 = NULL,
                                 n1 = rep(n, length(theta)),
                                 pp_threshold = rep(theta, each = length(pp)))
     
-    ppp <- purrr::pmap_dbl(crossargs, 
-                           ~calc_predictive(
-                             y = ..1, 
-                             n = ..2, 
-                             p0 = p0, delta = delta, prior = prior,
-                             S = S, N = N, theta = ..3)
-    )
-    
+    ppp <- furrr::future_pmap_dbl(
+      crossargs, 
+      ~calc_predictive(
+        y = ..1, 
+        n = ..2, 
+        p0 = p0, delta = delta, prior = prior,
+        S = S, N = N, theta = ..3),
+      .options = furrr_options(
+        seed = TRUE,
+        chunk_size = ceiling(nrow(crossargs)/future::nbrOfWorkers())))
+
     res <- dplyr::arrange(
       dplyr::select(
         tibble::add_column(crossargs, pp = rep(pp, length(theta)), ppp = ppp),

### 2021-06-04 R/calc_decision_rules.R
new one_sample_expansion vignette and other minor changes

@@ -26,17 +26,19 @@
 #' @param ppp The target predictive probability. e.g. Stop the trial if the 
 #' predictive probability falls below this target.
 #' 
-#' @return Returns a tibble with n at each look and r. Stop the trial at that
+#' @return Returns a tibble with n at each look and r. Stop the trial at that 
 #' look if the number of observed responses is <=r. At the end of the trial,
-#' the treatment is considered promising if the number of observed responses is 
-#' >r.
+#' the treatment is considered promising if the number of observed responses 
+#' is >r.
 #'
 #' @examples
 #' 
+#' \dontrun{
 #' set.seed(123)
 #' 
 #' # One-sample case
 #' calc_decision_rules(seq(5, 25, 5), p0 = 0.1, N = 25, theta = 0.86, ppp = 0.2)
+#' }
 #'
 #' @export
 
@@ -45,8 +47,7 @@ calc_decision_rules <- function(n, direction = "greater", p0,
                                 N, theta, ppp) {
 
   if((is.null(p0) & is.null(delta)) | (!is.null(p0) & !is.null(delta)))
-    stop("Exactly one of delta or p0 must be specified for the two-sample and 
-         one-sample case, respectively")
+    stop("Exactly one of delta or p0 must be specified for the two-sample and one-sample case, respectively")
   
   if(!direction %in% c("greater", "less")) 
     stop('direction must be either "greater" or "less"')
@@ -58,10 +59,11 @@ calc_decision_rules <- function(n, direction = "greater", p0,
   )
   
   # initialize the ytest parameter (will have 1 added in loop so will really start at 0)
-  ytest <- -1
+ 
   
   # Loop over each value of n and for each, continue while the predictive probability for a given ytest is less than the ppp threshold
   for(i in n) {
+    ytest <- -1
     pred <- 0
     while(pred < ppp) {
       ytest <- ytest + 1

### 2021-06-04 R/globals.R

@@ -8,6 +8,7 @@ utils::globalVariables(c("n0", "n1", "positive", "pp_threshold",
                          "Type I error", "Power", "Design", 
                          "Average N under the null", 
                          "Average N under the alternative",
-                         "Distance to min(N under null) and max(N under alt)",
+                         "Distance to optimal efficiency",
                          "mean_n0_null", "mean_n0_alt", "mean_n_null",
-                         "mean_n_alt", "ab_dist_metric", "Distance to (0, 1)"))
+                         "mean_n_alt", "ab_dist_metric", 
+                         "Distance to optimal accuracy"))

### 2021-06-04 R/one_sample_cal_tbl-data.R

@@ -1,11 +1,11 @@
 #' Output from a one-sample call to \code{calibrate_thresholds}
 #'
 #' This .rda file contains output from a one-sample call to 
-#' \code{calibrate_threshold}. 
+#' \code{calibrate_thresholds()}. 
 #' See the vignette titled "One-sample expansion cohort" for a description of
 #' the input parameters used, or run 
 #' \code{one_sample_cal_tbl$calibrate_thresholds_inputs}
-#' to see a list of the original function inputs. For use in testing global 
+#' to see a list of the original function inputs. For use in testing  
 #' functions and in vignettes.
 #'
 #' @docType data
@@ -15,14 +15,15 @@
 #' @format A list containing a 
 #' 1) a tibble 'res_summary' containing the posterior probability threshold 
 #' (pp_threshold), the predictive probability threshold (ppp_threshold), 
-#' the ceiling of the mean sample size (mean_n0 and mean_n1 for two-sample case; 
-#' mean_n1 for one-sample case), and the proportion of positive trials.
+#' the mean sample size (mean_n0 and mean_n1 for two-sample case; 
+#' mean_n1 for one-sample case), and the proportion of positive trials under 
+#' the null and alternative response rates.
 #' 2) 'call_list' containing the original function call
 #' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the 
 #' original function call
-#' 4) 'simon_res' gives results from a Simon's two-stage minimax design with
-#' unacceptable rate 0.1, acceptable rate 0.3, and theoretical type I error of
-#' 0.05 and power of 0.8
+#' 4) 'protocol_res' gives results from the protocol-specified design of
+#' the atezolizumab case study, as detailed in the vignette titled 
+#' "One-sample expansion cohort"
 #'
 #' @keywords datasets
 "one_sample_cal_tbl"

### 2021-06-04 R/optimize_design.R

@@ -132,10 +132,14 @@ optimize_design.calibrate_thresholds <- function(x,
   
   
   list(
-    "The design that minimizes Euclidean distance to (0, 1) on the type I error by power plot is:" =
-      opt_ab[opt_ab$`Distance to (0, 1)` == min(opt_ab$`Distance to (0, 1)`), ],
-    "The design that minimizes Euclidean distance to the point with minimum average sample size under the null and maximum average sample size under the alternative is:" = 
+    "Optimal accuracy design:" =
+      opt_ab[opt_ab$`Distance to (0, 1)` == min(opt_ab$`Distance to (0, 1)`), 
+             c("pp_threshold", "ppp_threshold", "Type I error", "Power",
+               "Average N under the null", "Average N under the alternative")],
+    "Optimal efficiency design:" = 
       opt_nn[opt_nn$`Distance to min(N under null) and max(N under alt)` == 
-               min(opt_nn$`Distance to min(N under null) and max(N under alt)`), ]
+               min(opt_nn$`Distance to min(N under null) and max(N under alt)`), 
+             c("pp_threshold", "ppp_threshold", "Type I error", "Power",
+               "Average N under the null", "Average N under the alternative")]
   )
 }

### 2021-06-04 R/plot.R

@@ -12,13 +12,17 @@
 #' @param minimum_power a numeric between 0 and 1 specifying the minimum 
 #' acceptable power. Specify NULL to return the full range of resulting power.
 #' Defaults to 0.8.
+#' @param plotly a logical indicator of whether you want the plots returned as 
+#' interactive plotly plots or non-interactive ggplots
 #' @param ... unused
 #' 
 #' @export
 
 plot.calibrate_thresholds <- function(x, 
                                       type1_range = c(0.05, 0.1), 
-                                      minimum_power = 0.8, ...) {
+                                      minimum_power = 0.8, 
+                                      plotly = FALSE,
+                                      ...) {
   
   if(any(class(x) =="calibrate_thresholds") == FALSE)
     stop("x must be class 'calibrate_thresholds', usually an object returned from a call to the function ppseq::calibrate_thresholds()")
@@ -72,8 +76,8 @@ plot.calibrate_thresholds <- function(x,
               prop_pos_null <= type1_range[2] &
               prop_pos_alt >= minimum_power
           ),
-          Design = paste0("Posterior threshold ", pp_threshold, 
-                          " and predictive threshold ", ppp_threshold),
+          Design = paste0("Posterior threshold = ", pp_threshold, 
+                          " and predictive threshold = ", ppp_threshold),
           ab_dist_metric = ((prop_pos_null - 0)^2 + 
                               (prop_pos_alt - 1)^2)^(1/2),
           n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 + 
@@ -83,36 +87,50 @@ plot.calibrate_thresholds <- function(x,
         Power = prop_pos_alt,
         `Average N under the null` = mean_n1_null,
         `Average N under the alternative` = mean_n1_alt,
-        `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
-        `Distance to (0, 1)` = ab_dist_metric
+        `Distance to optimal efficiency` = n_dist_metric,
+        `Distance to optimal accuracy` = ab_dist_metric
       )
   }
   
   plot_ab <- 
-    dplyr::slice(
-      dplyr::group_by(
-        dplyr::arrange(plot_x, 
-                       `Distance to (0, 1)`, -pp_threshold, -ppp_threshold),
-        `Distance to (0, 1)`), 
-      1)
+    dplyr::mutate(
[…]

### 2021-06-04 R/print.R

@@ -55,6 +55,14 @@ print.calibrate_thresholds <- function(x,
   if(!is.numeric(type1_range) & !is.null(type1_range))
     stop("type1_range must be a numeric vector of length 2 or NULL")
   
+  if(minimum_power < 0 | minimum_power > 1)
+    stop("minimum_power must be a numeric value between 0 and 1")
+  
+  if(type1_range[1] < 0 | type1_range[1] > 1 | 
+     type1_range[2] < 0 | type1_range[2] > 1 |
+     type1_range[1] > type1_range[2])
+    stop("type1_range must be a vector of numeric values, both between 0 and 1, with the first value less than the second value")
+  
   if (is.null(type1_range) & is.null(minimum_power)) {
     print(x$res_summary)
   } else if(is.null(type1_range) & !is.null(minimum_power)) {

### 2021-06-14 README.Rmd
Added actions.

@@
 [![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
+[![R-CMD-check](https://github.com/zabore/ppseq/workflows/R-CMD-check/badge.svg)](https://github.com/zabore/ppseq/actions)
 <!-- badges: end -->

### 2021-06-22 R/calibrate_posterior_threshold.R
changed a couple of arguments and updated everything downstream

@@ -8,7 +8,7 @@
 #' threshold. In an alternative case, this will result in the power at a given
 #' threshold.
 #'
-#' @param prob vector of length two containing the probability of event in
+#' @param p vector of length two containing the probability of event in
 #' the standard of care and experimental arm c(p0, p1) for the two-sample case;
 #' integer of event probability for one-sample case
 #' @param N vector of length two containing the total sample size c(N0, N1)
@@ -35,13 +35,13 @@
 #'
 #' # One-sample case
 #' calibrate_posterior_threshold(
-#'   prob = 0.1, N = 50, direction = "greater",
+#'   p = 0.1, N = 50, direction = "greater",
 #'   p0 = 0.1, delta = NULL, prior = c(0.5, 0.5), S = 5000, theta = c(0.9, 0.95)
 #' )
 #'
 #' # # Two-sample case (not run)
 #' # calibrate_posterior_threshold(
-#' #   prob = c(0.1, 0.1), N = c(50, 50),
+#' #   p = c(0.1, 0.1), N = c(50, 50),
 #' #   direction = "greater", p0 = NULL, delta = 0, prior = c(0.5, 0.5),
 #' #   S = 5000, theta = c(0.9, 0.95)
 #' # )
@@ -50,18 +50,18 @@
 #' @importFrom tibble tibble
 #' @export
 
-calibrate_posterior_threshold <- function(prob, N,
+calibrate_posterior_threshold <- function(p, N,
                                           direction = "greater", p0 = NULL,
                                           delta = 0, prior = c(0.5, 0.5),
                                           S = 5000, theta) {
   if (length(N) == 2) {
     y0 <- map_dbl(
       seq_len(S),
-      ~ rbinom(n = 1, size = N[1], prob = prob[1])
+      ~ rbinom(n = 1, size = N[1], prob = p[1])
     )
     y1 <- map_dbl(
       seq_len(S),
-      ~ rbinom(n = 1, size = N[2], prob = prob[2])
+      ~ rbinom(n = 1, size = N[2], prob = p[2])
     )
 
     posts <-
@@ -76,7 +76,7 @@ calibrate_posterior_threshold <- function(prob, N,
   } else if (length(N) == 1) {
     y1 <- map_dbl(
       seq_len(S),
-      ~ rbinom(n = 1, size = N, prob = prob)
+      ~ rbinom(n = 1, size = N, prob = p)
     )
 
     posts <-

### 2021-06-22 R/calibrate_thresholds.R

@@ -5,7 +5,7 @@
 #' generates a single dataset of n and response count at each look based on the
 #' response probability(ies)
 #'
-#' @param prob vector of length two containing the probability of event in
+#' @param p vector of length two containing the probability of event in
 #' the standard of care and experimental arm c(p0, p1) for the two-sample case;
 #' integer of event probability for one-sample case
 #' @param n matrix containing the total number of patients accrued so far at
@@ -26,31 +26,31 @@
 #' set.seed(123)
 #'
 #' # One-sample case
-#' sim_dat1(prob = 0.1, n = seq(5, 25, 5))
+#' sim_dat1(p = 0.1, n = seq(5, 25, 5))
 #'
 #' # Two-sample case
-#' sim_dat1(prob = c(0.1, 0.3), n = cbind(seq(5, 25, 5), seq(5, 25, 5)))
+#' sim_dat1(p = c(0.1, 0.3), n = cbind(seq(5, 25, 5), seq(5, 25, 5)))
 #' }
 #' @importFrom stats rbinom
 #' @importFrom tibble tibble
-sim_dat1 <- function(prob, n) {
-  if (length(prob) == 2) {
+sim_dat1 <- function(p, n) {
+  if (length(p) == 2) {
     if (length(n) == 2 & is.matrix(n) == FALSE) {
       n <- matrix(n, nrow = 1)
     }
-    y0 <- rbinom(n = 1, size = n[1, 1], prob = prob[1])
-    y1 <- rbinom(n = 1, size = n[1, 2], prob = prob[2])
+    y0 <- rbinom(n = 1, size = n[1, 1], prob = p[1])
+    y1 <- rbinom(n = 1, size = n[1, 2], prob = p[2])
     if (length(n) > 2) {
       for (i in seq_len(nrow(n))[-1]) {
         y0 <- c(y0, y0[length(y0)] + rbinom(
           n = 1,
           size = n[i, 1] - n[i - 1, 1],
-          prob = prob[1]
+          prob = p[1]
         ))
         y1 <- c(y1, y1[length(y1)] + rbinom(
           n = 1,
           size = n[i, 2] - n[i - 1, 2],
-          prob = prob[2]
+          prob = p[2]
         ))
       }
     }
@@ -60,12 +60,12 @@ sim_dat1 <- function(prob, n) {
       y0 = y0,
       y1 = y1
     ))
-  } else if (length(prob) == 1) {
-    y1 <- rbinom(n = 1, size = n[1], prob = prob)
+  } else if (length(p) == 1) {
[…]

### 2021-06-22 R/globals.R

@@ -12,8 +12,5 @@ utils::globalVariables(c(
   "Distance to optimal efficiency",
   "mean_n0_null", "mean_n0_alt", "mean_n_null",
   "mean_n_alt", "ab_dist_metric",
-  "Distance to optimal accuracy",
-  "Distance to (0, 1)",
-  "Distance to min(N under null) and max(N under alt)",
-  "Distance to (0, 1) Distance to min(N under null) and max(N under alt)"
+  "Distance to optimal accuracy"
 ))

### 2021-06-22 R/optimize_design.R

@@ -38,7 +38,7 @@ optimize_design <- function(x,
 #' set.seed(123)
 #'
 #' cal_tbl <- calibrate_thresholds(
-#'   prob_null = 0.1, prob_alt = 0.3,
+#'   p_null = 0.1, p_alt = 0.3,
 #'   n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL,
 #'   prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000,
 #'   pp_threshold = c(0.9, 0.95, 0.96, 0.98),
@@ -76,7 +76,7 @@ optimize_design.calibrate_thresholds <- function(x,
 
   options(tibble.width = Inf)
 
-  if (length(x$inputs$prob_null) == 2) {
+  if (length(x$inputs$p_null) == 2) {
     opt_x <-
       rename(
         mutate(
@@ -97,10 +97,10 @@ optimize_design.calibrate_thresholds <- function(x,
         Power = prop_pos_alt,
         `Average N under the null` = mean_n_null,
         `Average N under the alternative` = mean_n_alt,
-        `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
-        `Distance to (0, 1)` = ab_dist_metric
+        `Distance to optimal efficiency` = n_dist_metric,
+        `Distance to optimal accuracy` = ab_dist_metric
       )
-  } else if (length(x$inputs$prob_null) == 1) {
+  } else if (length(x$inputs$p_null) == 1) {
     opt_x <-
       rename(
         mutate(
@@ -119,8 +119,8 @@ optimize_design.calibrate_thresholds <- function(x,
         Power = prop_pos_alt,
         `Average N under the null` = mean_n1_null,
         `Average N under the alternative` = mean_n1_alt,
-        `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
-        `Distance to (0, 1)` = ab_dist_metric
+        `Distance to optimal efficiency` = n_dist_metric,
+        `Distance to optimal accuracy` = ab_dist_metric
       )
   }
 
@@ -129,9 +129,9 @@ optimize_design.calibrate_thresholds <- function(x,
       group_by(
         arrange(
           opt_x,
-          `Distance to (0, 1)`, -pp_threshold, -ppp_threshold
+          `Distance to optimal accuracy`, -pp_threshold, -ppp_threshold
         ),
-        `Distance to (0, 1)`
+        `Distance to optimal accuracy`
       ),
       1
     )
[…]

### 2021-06-22 R/plot.R

@@ -20,7 +20,6 @@
 #'                   row_number
 #' @importFrom ggplot2 ggplot aes geom_point xlim ylim labs 
 #'                     scale_color_viridis_c theme_bw theme
-#' @importFrom gridExtra grid.arrange
 #' @importFrom plotly ggplotly
 #' @export
 
@@ -44,7 +43,7 @@ plot.calibrate_thresholds <- function(x,
   if (!is.numeric(type1_range) & !is.null(type1_range))
     stop("type1_range must be a numeric vector of length 2 or NULL")
 
-  if (length(x$inputs$prob_null) == 2) {
+  if (length(x$inputs$p_null) == 2) {
     plot_x <-
       rename(
         mutate(
@@ -72,7 +71,7 @@ plot.calibrate_thresholds <- function(x,
         `Distance to min(N under null) and max(N under alt)` = n_dist_metric,
         `Distance to (0, 1)` = ab_dist_metric
       )
-  } else if (length(x$inputs$prob_null) == 1) {
+  } else if (length(x$inputs$p_null) == 1) {
     plot_x <-
       rename(
         mutate(
@@ -144,7 +143,9 @@ plot.calibrate_thresholds <- function(x,
         x = `Type I error`,
         y = Power,
         color = `Distance to optimal accuracy`,
-        Design = Design
+        Design = Design,
+        `Average N under the null` = `Average N under the null`,
+        `Average N under the alternative` = `Average N under the alternative`
       )
     ) +
     geom_point(
@@ -168,7 +169,9 @@ plot.calibrate_thresholds <- function(x,
         x = `Average N under the null`,
         y = `Average N under the alternative`,
         color = `Distance to optimal efficiency`,
-        Design = Design
+        Design = Design,
+        `Type I error` = `Type I error`,
+        Power = Power
       )
     ) +
     geom_point(
@@ -191,8 +194,8 @@ plot.calibrate_thresholds <- function(x,
     theme_bw() +
     theme(legend.position = "bottom")
 
-  ifelse(plotly == TRUE,
-    list(ggplotly(p1), ggplotly(p2)),
-    grid.arrange(p1, p2, nrow = 1, ncol = 2)
-  )
+  if(plotly == TRUE)
+    list(ggplotly(p1), ggplotly(p2)) else
+      p1 + p2
+  
 }

### 2021-06-22 R/print.R

@@ -23,7 +23,7 @@
 #' set.seed(123)
 #'
 #' cal_tbl <- calibrate_thresholds(
-#'   prob_null = 0.1, prob_alt = 0.3,
+#'   p_null = 0.1, p_alt = 0.3,
 #'   n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL,
 #'   prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000,
 #'   pp_threshold = c(0.9, 0.95, 0.96, 0.98),

### 2021-06-22 tests/spelling.R

@@ -1,6 +1,3 @@
-if (requireNamespace('spelling', quietly = TRUE))
-  spelling::spell_check_test(
-    vignettes = TRUE, 
-    error = FALSE,
-    skip_on_cran = TRUE
-  )
+if(requireNamespace('spelling', quietly = TRUE))
+  spelling::spell_check_test(vignettes = TRUE, error = FALSE,
+                             skip_on_cran = TRUE)

### 2021-06-22 tests/testthat/test-calibrate_posterior_threshold.R

@@ -5,7 +5,7 @@ test_that(
     set.seed(123)
     expect_snapshot(
       calibrate_posterior_threshold(
-        prob = c(0.1, 0.1), N = c(50, 50),
+        p = c(0.1, 0.1), N = c(50, 50),
         direction = "greater", p0 = NULL, delta = 0, prior = c(0.5, 0.5),
         S = 50, theta = c(0.9, 0.95)
       )

### 2021-06-22 tests/testthat/test-calibrate_thresholds.R

@@ -3,7 +3,7 @@ test_that(
   "one-sample simulate data",
   {
     set.seed(123)
-    expect_snapshot(sim_dat1(prob = 0.1, n = c(5, 10)))
+    expect_snapshot(sim_dat1(p = 0.1, n = c(5, 10)))
   }
 )
 
@@ -12,7 +12,7 @@ test_that(
   {
     set.seed(123)
     expect_snapshot(
-      sim_dat1(prob = c(0.1, 0.3), n = cbind(c(5, 10), c(5, 10)))
+      sim_dat1(p = c(0.1, 0.3), n = cbind(c(5, 10), c(5, 10)))
     )
   }
 )
@@ -21,7 +21,7 @@ test_that(
   "evaluate threshold one-sample case",
   {
     set.seed(123)
-    dat1 <- sim_dat1(prob = 0.1, n = c(5, 10))
+    dat1 <- sim_dat1(p = 0.1, n = c(5, 10))
     expect_snapshot(
       eval_thresh(dat1, 0.95, 0.3, p0 = 0.1, delta = NULL, S = 500, N = 25)
     )
@@ -32,7 +32,7 @@ test_that(
   "evaluate threshold two-sample case",
   {
     set.seed(123)
-    dat2 <- sim_dat1(prob = c(0.1, 0.3), 
+    dat2 <- sim_dat1(p = c(0.1, 0.3), 
                      n = cbind(c(5, 10), c(5, 10)))
     expect_snapshot(eval_thresh(dat2, 0.95, 0.3, S = 500, N = c(25, 25)))
   }
@@ -44,11 +44,10 @@ test_that(
 
 set.seed(123)
 calibrate_thresholds( 
-  prob_null = 0.01, 
-  prob_alt = 0.9, 
+  p_null = 0.01, 
+  p_alt = 0.9, 
   n = c(5, 25), 
   direction = "greater", 
-  p0 = 0.1, 
   delta = NULL,
   prior = c(0.5, 0.5), 
   S = 200, 
@@ -64,11 +63,10 @@ test_that(
     set.seed(123)
     expect_snapshot(
       calibrate_thresholds( 
-        prob_null = 0.01, 
-        prob_alt = 0.9, 
+        p_null = 0.01, 
+        p_alt = 0.9, 
         n = c(5, 25), 
         direction = "greater", 
-        p0 = 0.1, 
         delta = NULL,
         prior = c(0.5, 0.5), 
         S = 200, 

### 2021-06-22 tests/testthat/test-optimize_design.R

@@ -4,8 +4,8 @@ test_that(
   {
     set.seed(123)
     cal_tbl <- calibrate_thresholds(
-      prob_null = 0.1, prob_alt = 0.3,
-      n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL,
+      p_null = 0.1, p_alt = 0.3,
+      n = seq(5, 25, 5), direction = "greater", delta = NULL,
       prior = c(0.5, 0.5), S = 1000, N = 25, nsim = 10,
       pp_threshold = c(0.9, 0.95, 0.96, 0.98),
       ppp_threshold = c(0.05, 0.1)

### 2021-06-22 tests/testthat/test-sim_single_trial.R

@@ -4,8 +4,8 @@ test_that(
   {
     set.seed(123)
     expect_snapshot(
-      sim_single_trial(prob = 0.3, n = c(5, 10), direction = "greater",
-                       p = 0.1, delta = NULL, prior = c(0.5, 0.5), S = 5000, 
+      sim_single_trial(p = 0.3, n = c(5, 10), direction = "greater",
+                       p0 = 0.1, delta = NULL, prior = c(0.5, 0.5), S = 5000, 
                        N = 25, theta = 0.95)
     )
   }
@@ -16,7 +16,7 @@ test_that(
   {
     set.seed(123)
     expect_snapshot(
-      sim_single_trial(prob = c(0.1, 0.3), n = cbind(c(5, 10), c(5, 10)),
+      sim_single_trial(p = c(0.1, 0.3), n = cbind(c(5, 10), c(5, 10)),
                        direction = "greater", p0 = NULL, delta = 0, 
                        prior = c(0.5, 0.5), S = 5000, N = c(50, 50), 
                        theta = 0.95)

### 2021-06-22 R/calc_decision_rules.R
changing the p0 and delta argument defaults

@@ -14,8 +14,7 @@
 #' @param p0 The target value to compare to in the one-sample case
 #' (i.e. the null response rate)
 #' @param delta clinically meaningful difference between groups.
-#' Typically 0. NULL for one-sample case (default). Placeholder for when
-#' two-sample case is added.
+#' Typically 0 for two-sample case. NULL for one-sample case (default). 
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000

### 2021-06-22 R/calc_posterior.R

@@ -14,12 +14,13 @@
 #' two-sample case; integer of total responses y for one-sample case
 #' @param n vector of length two containing the sample size so far c(n0, n1)
 #' for two-sample case; integer of sample size so far n for one-sample case
+#' @param p0 The target value to compare to in the one-sample case. Set to NULL 
+#' in two-sample case.
 #' @param direction "greater" (default) if interest is in p(p1 > p0) and "less"
 #' if interest is in p(p1 < p0) for two-sample case. For one-sample case,
 #' "greater" if interest is in p(p > p0) and "less" if interest is in p(p < p0).
-#' @param p0 The target value to compare to in the one-sample case
 #' @param delta clinically meaningful difference between groups.
-#' Typically 0 (default).
+#' Typically 0 for two-sample case. NULL for one-sample case (default).
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
@@ -39,8 +40,8 @@
 #' @importFrom stats rbeta
 #' @export
 
-calc_posterior <- function(y, n, direction = "greater", p0 = NULL,
-                           delta = 0, prior = c(0.5, 0.5), S = 5000) {
+calc_posterior <- function(y, n, p0, direction = "greater", delta = NULL, 
+                           prior = c(0.5, 0.5), S = 5000) {
   if (length(y) != length(n))
     stop("y and n must be the same length")
 

### 2021-06-22 R/calibrate_posterior_threshold.R

@@ -13,12 +13,13 @@
 #' integer of event probability for one-sample case
 #' @param N vector of length two containing the total sample size c(N0, N1)
 #' for two-sample case; integer of sample size so far N for one-sample case
+#' @param p0 The target value to compare to in the one-sample case. Set to 
+#' NULL for the two-sample case.
 #' @param direction "greater" (default) if interest is in p(p1 > p0) and "less"
 #' if interest is in p(p1 < p0) for two-sample case. For one-sample case,
 #' "greater" if interest is in p(p > p0) and "less" if interest is in p(p < p0).
-#' @param p0 The target value to compare to in the one-sample case
 #' @param delta clinically meaningful difference between groups.
-#' Typically 0 (default).
+#' Typically 0 for the two-sample case. NULL for the one-sample case (default).
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples drawn from the posterior, and number of simulated
@@ -50,9 +51,9 @@
 #' @importFrom tibble tibble
 #' @export
 
-calibrate_posterior_threshold <- function(p, N,
-                                          direction = "greater", p0 = NULL,
-                                          delta = 0, prior = c(0.5, 0.5),
+calibrate_posterior_threshold <- function(p, N, p0,
+                                          direction = "greater", 
+                                          delta = NULL, prior = c(0.5, 0.5),
                                           S = 5000, theta) {
   if (length(N) == 2) {
     y0 <- map_dbl(

### 2021-06-22 R/calibrate_thresholds.R

@@ -86,18 +86,20 @@ sim_dat1 <- function(p, n) {
 #' @param pp_threshold the posterior probability threshold of interest
 #' @param ppp_threshold the posterior probability threshold of interest for
 #' futility monitoring
-#' @param direction "greater" (default) if interest is in P(p1 > p0) and "less"
-#' if interest is in P(p1 < p0) for two-sample case. For one-sample case,
-#' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
-#' @param p0 The target value to compare to in the one-sample case
-#' @param delta clinically meaningful difference between groups.
-#' Typically 0 (default).
-#' @param prior hyperparameters of prior beta distribution.
-#' Beta(0.5, 0.5) is default
-#' @param S number of samples, default is 5000
+#' @param p0 The target value to compare to in the one-sample case. Set to 
+#' NULL for the two-sample case.
 #' @param N the total planned sample size at the end of the trial, c(N0, N1)
 #' for two-sample case; integer of total planned sample size at end of trial N
 #' for one-sample case
+#' @param direction "greater" (default) if interest is in P(p1 > p0) and "less"
+#' if interest is in P(p1 < p0) for two-sample case. For one-sample case,
+#' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
+#' @param delta clinically meaningful difference between groups.
+#' Typically 0 for the two-sample case. NULL for the one-sample case (default).
+#' @param prior hyperparameters of prior beta distribution.
+#' Beta(0.5, 0.5) is default
+#' @param S number of samples, default is 5000
+
 #'
 #' @return Returns a tibble with the total sample size at the end of the
 #' trial, the number of responses observed at the end of the trial, the
@@ -119,9 +121,9 @@ sim_dat1 <- function(p, n) {
 #' }
 #' @importFrom tibble add_column
 #' @importFrom dplyr mutate case_when
-eval_thresh <- function(data, pp_threshold, ppp_threshold,
-                        direction = "greater", p0 = NULL, delta = 0,
-                        prior = c(0.5, 0.5), S = 5000, N) {
+eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N, 
+                        direction = "greater", delta = NULL,
+                        prior = c(0.5, 0.5), S = 5000) {
   decision <- NULL
   ppp <- NULL
[…]

### 2021-06-22 R/optimize_design.R

@@ -39,10 +39,11 @@ optimize_design <- function(x,
 #'
 #' cal_tbl <- calibrate_thresholds(
 #'   p_null = 0.1, p_alt = 0.3,
-#'   n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL,
-#'   prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000,
+#'   n = seq(5, 25, 5), N = 25, 
 #'   pp_threshold = c(0.9, 0.95, 0.96, 0.98),
-#'   ppp_threshold = seq(0.05, 0.2, 0.05)
+#'   ppp_threshold = seq(0.05, 0.2, 0.05),
+#'   direction = "greater", delta = NULL,
+#'   prior = c(0.5, 0.5), S = 5000, nsim = 1000
 #' )
 #'
 #' optimize_design(cal_tbl)

### 2021-06-22 R/print.R

@@ -24,10 +24,11 @@
 #'
 #' cal_tbl <- calibrate_thresholds(
 #'   p_null = 0.1, p_alt = 0.3,
-#'   n = seq(5, 25, 5), direction = "greater", p0 = 0.1, delta = NULL,
-#'   prior = c(0.5, 0.5), S = 5000, N = 25, nsim = 1000,
+#'   n = seq(5, 25, 5), N = 25, 
 #'   pp_threshold = c(0.9, 0.95, 0.96, 0.98),
-#'   ppp_threshold = seq(0.05, 0.2, 0.05)
+#'   ppp_threshold = seq(0.05, 0.2, 0.05),
+#'   direction = "greater", delta = NULL,
+#'   prior = c(0.5, 0.5), S = 5000, nsim = 1000
 #' )
 #'
 #' print(cal_tbl)

### 2021-06-22 R/sim_single_trial.R

@@ -19,18 +19,18 @@
 #' at the end of the trial, this can be a vector specifying the total sample
 #' size c(N0, N1) for the two-sample case or an integer specifying the total
 #' sample size N for the one-sample case.
-#' @param direction "greater" (default) if interest is in P(p1 > p0) and "less"
-#' if interest is in P(p1 < p0) for two-sample case. For one-sample case,
-#' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
 #' @param p0 The target value to compare to in the one-sample case
-#' @param delta clinically meaningful difference between groups.
-#' Typically 0 (default).
-#' @param prior hyperparameters of prior beta distribution.
-#' Beta(0.5, 0.5) is default
-#' @param S number of samples, default is 5000
 #' @param N the total planned sample size at the end of the trial, c(N0, N1)
 #' for two-sample case; integer of total planned sample size at end of trial N
 #' for one-sample case
+#' @param direction "greater" (default) if interest is in P(p1 > p0) and "less"
+#' if interest is in P(p1 < p0) for two-sample case. For one-sample case,
+#' "greater" if interest is in P(p > p0) and "less" if interest is in P(p < p0).
+#' @param delta clinically meaningful difference between groups.
+#' Typically 0 for the two-sample case. NULL for the one-sample case (default).
+#' @param prior hyperparameters of prior beta distribution.
+#' Beta(0.5, 0.5) is default
+#' @param S number of samples, default is 5000
 #' @param theta The target posterior probability. e.g. Efficacy decision if
 #' P(p1 > p0) > theta for the two-sample case with greater direction.
 #' Default is 0.95. Can be a vector if interest is in selecting from among a
@@ -66,9 +66,10 @@
 #' #  )
 #' @export
 
-sim_single_trial <- function(p, n, direction = "greater", p0 = NULL,
-                             delta = 0, prior = c(0.5, 0.5), S = 5000,
-                             N, theta = 0.95) {
+sim_single_trial <- function(p, n, p0, N, 
+                             direction = "greater", delta = NULL, 
+                             prior = c(0.5, 0.5), S = 5000,
+                             theta = 0.95) {
   if ((is.null(p0) & is.null(delta)) | (!is.null(p0) & !is.null(delta)))
     stop("Exactly one of delta or p0 must be specified for the two-sample and
          one-sample case, respectively")

### 2021-06-22 R/two_sample_cal_tbl-data.R

@@ -1,28 +1,0 @@
-#' Output from a two-sample call to \code{calibrate_threshold}
-#'
-#' This .rda file contains output from a two-sample call to
-#' \code{calibrate_threshold}.
-#' See the vignette titled "Two-sample randomized phase 2" for a description of
-#' the input parameters used, or run
-#' \code{two_sample_cal_tbl$calibrate_thresholds_inputs}
-#' to see a list of the original function inputs. For use in testing global
-#' functions and in vignettes.
-#'
-#' @docType data
-#'
-#' @usage data(two_sample_cal_tbl)
-#'
-#' @format A list containing a
-#' 1) a tibble 'res_summary' containing the posterior probability threshold
-#' (pp_threshold), the predictive probability threshold (ppp_threshold),
-#' the ceiling of the mean sample size (mean_n0 and mean_n1 for two-sample case;
-#' mean_n1 for one-sample case), and the proportion of positive trials.
-#' 2) 'call_list' containing the original function call
-#' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the
-#' original function call
-#' 4) 'simon_res' gives results from a Simon's two-stage minimax design with
-#' unacceptable rate 0.1, acceptable rate 0.3, and theoretical type I error of
-#' 0.05 and power of 0.8
-#'
-#' @keywords datasets
-"two_sample_cal_tbl"

### 2021-06-23 inst/ppseq-package-paper/zabor-ppseq.Rmd
adding R Journal paper to inst folder

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
author:
  # see ?rjournal_article for more information
  - name: Emily C. Zabor
    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
    address:
    - 9500 Euclid Ave. CA-60
    - Cleveland, OH 44195 USA
    url: http://www.emilyzabor.com/
    orcid: 0000-0002-1402-4498
    email:  zabore2@ccf.org
  - name: Brian P. Hobbs
    email:  brian.hobbs@austin.utexas.edu
    orcid: 0000-0003-2189-5846
    affiliation: Dell Medical School, The University of Texas at Austin
    address:
    - 1501 Red River Street MC: Z0100
    - Austin, TX 78712
  - name: Alexander M. Kaizer
    email: alex.kaizer@cuanschutz.edu
    orcid: 0000-0003-2334-5514
    affiliation: Department of Biostatistics & Informatics, University of Colorado-Anschutz Medical Campus,  
    address:
    - 13001 E. 17th Place Campus Box B119
    - Aurora, CO 80045
  - name: Michael J. Kane
    email: michael.kane@yale.edu
    orcid: 0000-0003-1899-6662
    affiliation: Department of Biostatistics, Yale University
    address:
    - 22 Mill Pond Drive
    - Guilford, CT 06437
abstract: >
  Clinical trials in oncology are making increasing use of larger phase 1 and phase 2 sample sizes with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements. With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power. The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability. Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
preamble: |
  % Any extra LaTeX you need in the preamble

# per R journal requirement, the bib filename should be the same as the output 
# tex file. Don't forget to rename the bib file and change this example value.
bibliography: zabor-ppseq.bib

output: rticles::rjournal_article
---

## Introduction

In the context of cytotoxic treatments, phase 1 trials in oncology traditionally have the primary aim of identifying the maximum tolerated dose (MTD), defined as the highest dose that still maintains a certain pre-specified rate of toxicity, most commonly 30\%, in a dose-escalation phase. Designs for dose-escalation trials include the rule-based 3+3 design and the model-based continual reassessment method, among others. But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs. As a result, it is becoming increasingly common to include dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete. In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes that all share a common biomarker that the treatment is targeting, or to further characterize the pharmacokinetics and/or pharmacodynamics. The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].

But dose-expansion cohorts are not always planned in advance, and therefore may be subject to on-the-fly decision making that can lead to large sample sizes and poor statistical properties. For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 495 non-small-cell lung cancer patients across six disease-specific expansion cohorts [@Garon2015]. In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 95 patients, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol but was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].

Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error, and has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; Lee2008; Saville2014]. However, in order to be useful to investigators designing such trials, software must be made available that streamlines design. This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring. Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} assist investigators in comparing designs based on different thresholds for decision making, and we suggest optimization criteria for selecting the ideal design.

## Predictive probability monitoring

Consider the setting of a binary outcome, such as tumor response as measured by the RECIST criteria in the setting of a study in patients with solid tumors, where each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$. Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total enrolled patients. Let $p$ represent the probability of response, where $p_0$ represents the null response rate under no treatment or the standard of care treatment and $p_1$ represents the alternative response rate under the experimental treatment. Most dose-expansion studies with an efficacy aim will wish to test the null hypothesis $H_0: p \leq p_0$ versus the alternative hypothesis $H_1: p \geq p_1$.

The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution. Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$. Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$. Using the posterior probability, which represents the probability of success based only on the data accrued so far, we would declare a treatment efficacious if $\Pr(p>p_0 | X) > \theta$, where $\theta$ represents a pre-specified posterior decision threshold. The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$. Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$. The posterior predictive probability represents the probability that the treatment will be declared efficacious at the end of the trial when full enrollment is reached. We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e. $PPP<\theta^*$. Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success. Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.

## Package overview

The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring. The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error. The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds. Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative. The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power. The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.

### Optimization

After obtaining results for the all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options. The \pkg{ppseq} package introduces two optimization criteria to assist users in making a selection. The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power. The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power. The \code{optimize\_design()} function will return a list that contains the details of each of the two optimal designs.

### Visualizations

To assist users in comparing the results of the various design options, a \code{plot()} option is also available, that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020]. Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design. The motivation for including an interactive graphics option was the utility of the additional information available when hovering over each point. Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot.

### Decision rules

Finally, to ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function. The function takes a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0}, the posterior threshold of the selected design \code{theta}, the predictive threshold of the selected design \code{ppp}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}. The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design. The function results in a \code{tibble} with the sample size at each interim analysis and the decision point $r$. The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$. At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.

## Case study

To demonstrate the functionality of the \pkg{ppseq} package, we focus on a re-design of a dose-expansion cohort for the study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring. Atezolizumab is an anti-PD-L1 treatment that was originally tested in the phase 1 setting in a basket trial across a variety of cancer sites harboring PD-L1 mutations. The atezolizumab expansion study in mUC had the primary aim of further evaluating safety, pharmacodynamics and pharmacokinetics and therefore was not designed to meet any specific criteria for type I error or power. An expansion cohort in mUC was not part of the original protocol design, but was rather added later through a protocol amendment. The expansion cohort in mUC ultimately enrolled a total of 95 participants [@Powles2014]. Other expansion cohorts that were included in the original protocol, including in renal-cell carcinoma, non-small-cell lung cancer, and melanoma, were planned to have a sample size of 40. These pre-planned expansion cohorts were designed with a single interim analysis for futility that would stop the trial if 0 responses were seen in the first 14 patients enrolled. According to the trial protocol, this futility rule is associated with at most a 4.4\% chance of observing no responses in 14 patients if the true response rate is 20\% or higher. The protocol also states the widths of the 90\% confidence intervals for a sample size of 40 if the observed response rate is 30\%. There was no stated decision rule for efficacy since efficacy was not an explicit aim of the expansion cohorts. In the re-design we assume a null, or unacceptable, response rate of 0.1 and an alternative, or acceptable, response rate of 0.2. We plan a study with up to a total of 95 participants. In our sequential predictive probability design we will check for futility after every 5 patients are enrolled. We consider posterior thresholds of 0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, and 1, and predictive thresholds of 0.05, 0.1, 0.15, and 0.2.

First we install and load the \pkg{ppseq} package.

```{r eval = FALSE}
install.packages("ppseq")
```

```{r}
library(ppseq)
```

We use the \code{calibrate\_thresholds()} function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds. Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations. The user will be responsible for setting up a call to \code{future::plan()} that is appropriate to their operating environment and simulation setting. The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid. Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \pkg{ppseq} package.

```{r eval = FALSE}
library(future)

set.seed(123)

plan(multicore(workers = 76))

one_sample_cal_tbl <- 
  calibrate_thresholds(
    p_null = 0.1, 
    p_alt = 0.2, 
    n = seq(5, 95, 5),
    N = 95, 
    pp_threshold = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 
                     0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1),
    ppp_threshold = seq(0.05, 0.2, 0.05),
    direction = "greater", 
    delta = NULL, 
    prior = c(0.5, 0.5), 
    S = 5000, 
    nsim = 1000
    )
```

Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.01 and 0.2, and a minimum power of 0.7. We find that 35 of the 76 designs meet these criteria for type I error and power.

```{r}
print(
  one_sample_cal_tbl,
  type1_range = c(0.01, 0.2),
  minimum_power = 0.7
  )
```

We use the \code{optimize\_designs()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power. We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05. It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91. The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1. It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82. For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.

```{r}
optimize_design(
  one_sample_cal_tbl, 
  type1_range = c(0.05, 0.1), 
  minimum_power = 0.7
  )
```

To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations.

```{r eval = FALSE}
plot(
  one_sample_cal_tbl, 
  type1_range = c(0.01, 0.2), 
  minimum_power = 0.7,
  plotly = TRUE
  )
```

```{r echo = FALSE, fig.width = 9, fig.height = 6}
# Note that the above code does not render properly in R Markdown, though it does produce the plots correctly in an interactive session
p <- plot(
  one_sample_cal_tbl, 
  type1_range = c(0.01, 0.2), 
  minimum_power = 0.7,
  plotly = TRUE
  )

p[[1]]
p[[2]]
```

In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796. This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment. Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial. Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package. In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial. After the first 10 patients we would stop the trial if there were 0 responses, and so on. At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were >=14 responses.

```{r eval = FALSE}
set.seed(123)

one_sample_decision_tbl <- 
  calc_decision_rules(
    n = seq(5, 95, 5),
    N = 95, 
    theta = 0.92, 
    ppp = 0.1, 
    p0 = 0.1, 
    direction = "greater",
    delta = NULL, 
    prior = c(0.5, 0.5), 
    S = 5000
    )
```

```{r}
one_sample_decision_tbl
```

## Summary

With the focus of early stage clinical trial research in oncology shifting away from the study of cytotoxic treatments and toward immunotherapies and other non-cytotoxic treatments, new approaches to clinical trial design are needed that move beyond the traditional search for the maximum tolerated dose. Bayesian sequential predictive probability monitoring provides a natural and flexible way to expand the number of patients studied in phase 1 or to design phase 2 trials that allow for efficient early stopping for futility while maintaining control of type I error and power. The \pkg{ppseq} package implements functionality to consider a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e. type I error and power) or efficiency (i.e. average sample sizes under the null and alternative). Interactive visualization options are provided to ease comparison of the resulting design options. Once an ideal design is selected, a table of decision rules can be obtained to make trial conduct simple and straightforward.

### 2021-06-23 inst/ppseq-package-paper/zabor-ppseq.tex
adding R Journal paper to inst folder

--- 2021-06-23 inst/ppseq-package-paper/zabor-ppseq.Rmd
@@
----
-title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
-author:
-  # see ?rjournal_article for more information
-  - name: Emily C. Zabor
-    affiliation: Department of Quantitative Health Sciences & Taussig Cancer Institute, Cleveland Clinic
-    address:
-    - 9500 Euclid Ave. CA-60
-    - Cleveland, OH 44195 USA
-    url: http://www.emilyzabor.com/
-    orcid: 0000-0002-1402-4498
-    email:  zabore2@ccf.org
-  - name: Brian P. Hobbs
-    email:  brian.hobbs@austin.utexas.edu
-    orcid: 0000-0003-2189-5846
-    affiliation: Dell Medical School, The University of Texas at Austin
-    address:
-    - 1501 Red River Street MC: Z0100
-    - Austin, TX 78712
-  - name: Alexander M. Kaizer
-    email: alex.kaizer@cuanschutz.edu
-    orcid: 0000-0003-2334-5514
-    affiliation: Department of Biostatistics & Informatics, University of Colorado-Anschutz Medical Campus,  
-    address:
-    - 13001 E. 17th Place Campus Box B119
-    - Aurora, CO 80045
-  - name: Michael J. Kane
-    email: michael.kane@yale.edu
-    orcid: 0000-0003-1899-6662
-    affiliation: Department of Biostatistics, Yale University
-    address:
-    - 22 Mill Pond Drive
-    - Guilford, CT 06437
-abstract: >
-  Clinical trials in oncology are making increasing use of larger phase 1 and phase 2 sample sizes with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
-With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
-The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability.
-Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
-preamble: |
-  % Any extra LaTeX you need in the preamble
+% !TeX root = RJwrapper.tex
+\title{ppseq: An R Package for Sequential Predictive Probability Monitoring}
+\author{by Emily C. Zabor, Brian P. Hobbs, Alexander M. Kaizer, and Michael J. Kane}
 
-# per R journal requirement, the bib filename should be the same as the output 
-# tex file. Don't forget to rename the bib file and change this example value.
-bibliography: zabor-ppseq.bib
+\maketitle
 
-output: rticles::rjournal_article
----
+\abstract{%
+Clinical trials in oncology are making increasing use of larger phase 1
+and phase 2 sample sizes with the addition of baskets for different
+disease sub-types or multiple dosing groups, expansion cohorts to
+further study safety and obtain preliminary efficacy information, or
+randomization to further define efficacy or compare across doses, among
+other design elements. With the enlarged sample sizes come increasing
+ethical concerns regarding patients who enroll in clinical trials, and
+the need for rigorous statistical designs to ensure that trials can stop
+early for futility while maintaining traditional control of type I error
+and power. The R package \pkg{ppseq} provides a framework for designing
+early phase clinical trials using sequential monitoring based on the
+Bayesian predictive probability. Trial designs can be compared using
+interactive plots and optimized based on efficiency or accuracy.
+}
 
-## Introduction
+\hypertarget{introduction}{%
+\subsection{Introduction}\label{introduction}}
 
-In the context of cytotoxic treatments, phase 1 trials in oncology traditionally have the primary aim of identifying the maximum tolerated dose (MTD), defined as the highest dose that still maintains a certain pre-specified rate of toxicity, most commonly 30\%, in a dose-escalation phase.
-Designs for dose-escalation trials include the rule-based 3+3 design and the model-based continual reassessment method, among others.
-But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs.
-As a result, it is becoming increasingly common to include dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete.
-In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes that all share a common biomarker that the treatment is targeting, or to further characterize the pharmacokinetics and/or pharmacodynamics.
-The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].
+In the context of cytotoxic treatments, phase 1 trials in oncology
+traditionally have the primary aim of identifying the maximum tolerated
+dose (MTD), defined as the highest dose that still maintains a certain
+pre-specified rate of toxicity, most commonly 30\%, in a dose-escalation
+phase. Designs for dose-escalation trials include the rule-based 3+3
+design and the model-based continual reassessment method, among others.
+But with increasing study focused on non-cytotoxic treatments, such as
+immunotherapies, the MTD either may not exist or may not be relevant,
+and toxicities may either develop much later or even be chronic, making
+these treatments difficult to study with traditional dose-escalation
+designs. As a result, it is becoming increasingly common to include
+dose-expansion cohorts, in which additional patients are enrolled in
+phase 1 after the dose-escalation phase is complete. In this setup, the
+dose-escalation phase is considered phase 1a and used to assess the
+initial safety of multiple doses, then the dose-expansion phase is
+considered phase 1b and can have a variety of aims including to further
+refine the safety of one or more doses, to assess preliminary efficacy,
+to explore the treatment in various disease-specific subtypes that all
+share a common biomarker that the treatment is targeting, or to further
+characterize the pharmacokinetics and/or pharmacodynamics. The use of
+dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011
+\citep{Manji2013} and trials with dose-expansion cohorts led to higher
+response rates and more frequent success in phase 2 trials
+\citep{Bugano2017}.
 
-But dose-expansion cohorts are not always planned in advance, and therefore may be subject to on-the-fly decision making that can lead to large sample sizes and poor statistical properties.
-For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 495 non-small-cell lung cancer patients across six disease-specific expansion cohorts [@Garon2015].
-In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 95 patients, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol but was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
+But dose-expansion cohorts are not always planned in advance, and
+therefore may be subject to on-the-fly decision making that can lead to
+large sample sizes and poor statistical properties. For example, the
+KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3
+dose-escalation trial, included multiple protocol amendments and
+ultimately enrolled a total of 495 non-small-cell lung cancer patients
+across six disease-specific expansion cohorts \citep{Garon2015}. In a
+basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a
+variety of cancers and both with and without PD-L1 expression, an
+expansion cohort in metastatic urothelial bladder cancer ultimately
+enrolled 95 patients, despite the fact that no expansion cohort in this
+disease subtype was originally planned in the trial protocol but was
+rather added later in a protocol amendment in which the sample size was
+increased from what was initially planned
+\citep{Petrylak2018, Powles2014}.
 
-Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error, and has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; Lee2008; Saville2014].
-However, in order to be useful to investigators designing such trials, software must be made available that streamlines design.
-This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring.
-Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} assist investigators in comparing designs based on different thresholds for decision making, and we suggest optimization criteria for selecting the ideal design.
+Bayesian sequential predictive probability monitoring provides a natural
+framework for early phase oncology trial design, allowing flexibility to
+stop early for futility or safety concerns while also maintaining
+traditional levels of power and control of type I error, and has been
+proposed as an approach to interim monitoring in clinical trials
+previously {[}\citet{Dmitrienko2006}; Lee2008; Saville2014{]}. However,
+in order to be useful to investigators designing such trials, software
+must be made available that streamlines design. This paper introduces
+the \pkg{ppseq} package for the R software language \citep{RCT2020},
+which provides functions to design one-sample and two-sample early phase
+clinical trials using sequential predictive probability monitoring.
+Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly}
+assist investigators in comparing designs based on different thresholds
+for decision making, and we suggest optimization criteria for selecting
+the ideal design.
 
-## Predictive probability monitoring
+\hypertarget{predictive-probability-monitoring}{%
+\subsection{Predictive probability
+monitoring}\label{predictive-probability-monitoring}}
 
-Consider the setting of a binary outcome, such as tumor response as measured by the RECIST criteria in the setting of a study in patients with solid tumors, where each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$.
-Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total enrolled patients.
-Let $p$ represent the probability of response, where $p_0$ represents the null response rate under no treatment or the standard of care treatment and $p_1$ represents the alternative response rate under the experimental treatment.
-Most dose-expansion studies with an efficacy aim will wish to test the null hypothesis $H_0: p \leq p_0$ versus the alternative hypothesis $H_1: p \geq p_1$.
+Consider the setting of a binary outcome, such as tumor response as
+measured by the RECIST criteria in the setting of a study in patients
+with solid tumors, where each patient, denoted by \(i\), enrolled in the
+trial either has a response such that \(x_i = 1\) or does not have a
+response such that \(x_i = 0\). Then \(X = \sum_{i=1}^n x_i\) represents
+the number of responses out of \(n\) currently observed patients up to a
+maximum of \(N\) total enrolled patients. Let \(p\) represent the
+probability of response, where \(p_0\) represents the null response rate
+under no treatment or the standard of care treatment and \(p_1\)
+represents the alternative response rate under the experimental
+treatment. Most dose-expansion studies with an efficacy aim will wish to
+test the null hypothesis \(H_0: p \leq p_0\) versus the alternative
+hypothesis \(H_1: p \geq p_1\).
 
-The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution.
-Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$.
-Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$.
-Using the posterior probability, which represents the probability of success based only on the data accrued so far, we would declare a treatment efficacious if $\Pr(p>p_0 | X) > \theta$, where $\theta$ represents a pre-specified posterior decision threshold.
-The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$.
-Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$.
-The posterior predictive probability represents the probability that the treatment will be declared efficacious at the end of the trial when full enrollment is reached.
-We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e.
-$PPP<\theta^*$.
-Predictive probability thresholds closer to 0 lead to less frequent stopping for futility whereas thresholds near 1 lead to frequent stopping unless there is almost certain probability of success.
-Predictive probability provides an intuitive interim monitoring strategy for clinical trials that tells the investigator what the chances are of declaring the treatment efficacious at the end of the trial if we were to continue enrolling to the maximum planned sample size, based on the data observed in the trial to date.
+The Bayesian paradigm of statistics is founded on Bayes rule, which is a
+mathematical theory that specifies how to combine the prior
+distributions that define prior beliefs about parameters, such as the
+response rate \(p\), with the observed data, such as the total number of
+responses \(X\), yielding a posterior distribution. Here the prior
+distribution of the response rate \(\pi(p)\) has a beta distribution
+\(Beta(a_0, b_0)\) and our data \(X\) have a binomial distribution
+\(bin(n, p)\). Combining the likelihood function for the observed data
+\(L_x(p) \propto p^x (1-p)^{n-x}\) with the prior, we obtain the
+posterior distribution of the response rate, which follows the beta
+distribution \(p|x \sim Beta(a_0 + x, b_0 + n - x)\). Using the
+posterior probability, which represents the probability of success based
+only on the data accrued so far, we would declare a treatment
+efficacious if \(\Pr(p>p_0 | X) > \theta\), where \(\theta\) represents
+a pre-specified posterior decision threshold. The posterior predictive
+distribution of the number of future responses \(X^*\) in the remaining
+\(n^*=N-n\) future patients follows a beta-binomial distribution
+\(Beta-binomial(n^*, a_0 + x, b_0 + n - x)\). Then the posterior
+predictive probability (PPP), is calculated as
+\(PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)\).
+The posterior predictive probability represents the probability that the
+treatment will be declared efficacious at the end of the trial when full
+enrollment is reached. We would stop the trial early for futility if the
+posterior predictive probability dropped below a pre-specified threshold
+\(\theta^*\), i.e.~\(PPP<\theta^*\). Predictive probability thresholds
+closer to 0 lead to less frequent stopping for futility whereas
+thresholds near 1 lead to frequent stopping unless there is almost
+certain probability of success. Predictive probability provides an
+intuitive interim monitoring strategy for clinical trials that tells the
+investigator what the chances are of declaring the treatment efficacious
+at the end of the trial if we were to continue enrolling to the maximum
+planned sample size, based on the data observed in the trial to date.
 
-## Package overview
+\hypertarget{package-overview}{%
+\subsection{Package overview}\label{package-overview}}
 
-The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring.
-The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error.
-The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds.
-Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}.
-The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
-The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative.
-The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power.
-The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.
+The \pkg{ppseq} package facilitates the design of clinical trials
+utilizing sequential predictive probability monitoring. The main
+challenge in designing such a trial is joint calibration of the
+posterior probability and posterior predictive probability thresholds to
+be used in the trial in order to maintain the desired level of power and
+type I error. The \code{calibrate\_thresholds()} function will evaluate
+a grid of posterior and predictive thresholds provided by the user as
+vector inputs through the argument \code{pp\_threshold} for posterior
+thresholds and the argument \code{ppp\_threshold} for predictive
+thresholds. Other required arguments include the null response rate
+\code{p\_null}, the alternative response rate \code{p\_alt}, a vector of
+sample sizes at which interim analyses are to be performed \code{n} as
+well as the maximum total sample size \code{N}, the direction of the
+alternative hypothesis \code{direction}, a vector of the two
+hyperparameters of the prior beta distribution \code{prior}, and the
+number of posterior samples \code{S} and the number of simulated trial
+datasets \code{nsim}. The additional argument \code{delta} can specify
+the clinically meaningful difference between groups in the case of a
+two-sample trial design. The function returns a list, the first element
+of which is a \code{tibble} containing the posterior threshold,
+predictive threshold, the mean sample size under the null and the
+alternative, the proportion of positive trials under the null and
+alternative, and the proportion of trials stopped early under the null
+and alternative. The proportion of positive trials under the null
+represents the type I error and the proportion of positive trials under
+the alternative represents the power. The \code{print()} option will
+print the results summary for each combination of thresholds, filtered
+by an acceptable range of type I error and minimum power, if desired.
 
-### Optimization
+\hypertarget{optimization}{%
+\subsubsection{Optimization}\label{optimization}}
 
-After obtaining results for the all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options.
-The \pkg{ppseq} package introduces two optimization criteria to assist users in making a selection.
-The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power.
-The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power.
-The \code{optimize\_design()} function will return a list that contains the details of each of the two optimal designs.
+After obtaining results for the all combinations of evaluated posterior
+and predictive thresholds, the next step is to select the ideal design
+from among the various options. The \pkg{ppseq} package introduces two
+optimization criteria to assist users in making a selection. The first,
+called the ``optimal accuracy'' design, identifies the design that
+minimizes the Euclidean distance to the top left point on a plot of the
+type I error by the power. The second, called the ``optimal efficiency''
+design, identifies the design that minimizes the Euclidean distance to
+the top left point on a plot of the average sample size under the null
+by the average sample size under the alternative, subject to constraints
+on the type I error and power. The \code{optimize\_design()} function
+will return a list that contains the details of each of the two optimal
+designs.
 
-### Visualizations
+\hypertarget{visualizations}{%
+\subsubsection{Visualizations}\label{visualizations}}
 
-To assist users in comparing the results of the various design options, a \code{plot()} option is also available, that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020].
-Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design.
-The motivation for including an interactive graphics option was the utility of the additional information available when hovering over each point.
-Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot.
+To assist users in comparing the results of the various design options,
+a \code{plot()} option is also available, that allows creation of static
+plots using the \CRANpkg{ggplot2} package \citep{Wickham2016} or
+interactive plots using the \CRANpkg{plotly} package
+\citep{Sievert2020}. Two plots are produced, one plotting type I error
+by power and indicating the optimal accuracy design, and one plotting
+the average sample size under the null by the average sample size under
+the alternative and indicating the optimal efficiency design. The
+motivation for including an interactive graphics option was the utility
+of the additional information available when hovering over each point.
+Instead of simply eyeballing where points fall along the axes, users can
+see the specific type I error, power, average sample size under the
+null, average sample size under the alternative, the posterior and
+predictive thresholds associated with the design, as well as the
+distance to the upper left point on the plot.
 
-### Decision rules
+\hypertarget{decision-rules}{%
+\subsubsection{Decision rules}\label{decision-rules}}
 
-Finally, to ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function.
-The function takes a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0}, the posterior threshold of the selected design \code{theta}, the predictive threshold of the selected design \code{ppp}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}.
-The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
-The function results in a \code{tibble} with the sample size at each interim analysis and the decision point $r$.
-The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$.
-At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.
+Finally, to ease the implementation of clinical trials designed with
+sequential predictive probability monitoring, once a design has been
+selected, a table of decision rules can be produced using the
+\code{calc\_decision\_rules()} function. The function takes a vector of
+sample sizes at which interim analyses are to be performed \code{n} as
+well as the maximum total sample size \code{N}, the null value to
+compare to in the one-sample case \code{p0}, the posterior threshold of
+the selected design \code{theta}, the predictive threshold of the
+selected design \code{ppp}, the direction of the alternative hypothesis
+\code{direction}, a vector of the two hyperparameters of the prior beta
+distribution \code{prior}, and the number of posterior samples \code{S}.
+The additional argument \code{delta} can specify the clinically
+meaningful difference between groups in the case of a two-sample trial
+design. The function results in a \code{tibble} with the sample size at
+each interim analysis and the decision point \(r\). The trial would stop
+at a given look if the number of observed responses is \(\leq r\)
+otherwise the trial would continue enrolling if the number of observed
+responses is \(>r\). At the end of the trial when the maximum planned
+sample size is reached, the treatment would be considered promising if
+the number of observed responses is \(>r\).
 
-## Case study
+\hypertarget{case-study}{%
+\subsection{Case study}\label{case-study}}
 
-To demonstrate the functionality of the \pkg{ppseq} package, we focus on a re-design of a dose-expansion cohort for the study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring.
-Atezolizumab is an anti-PD-L1 treatment that was originally tested in the phase 1 setting in a basket trial across a variety of cancer sites harboring PD-L1 mutations.
-The atezolizumab expansion study in mUC had the primary aim of further evaluating safety, pharmacodynamics and pharmacokinetics and therefore was not designed to meet any specific criteria for type I error or power.
-An expansion cohort in mUC was not part of the original protocol design, but was rather added later through a protocol amendment.
-The expansion cohort in mUC ultimately enrolled a total of 95 participants [@Powles2014].
-Other expansion cohorts that were included in the original protocol, including in renal-cell carcinoma, non-small-cell lung cancer, and melanoma, were planned to have a sample size of 40.
-These pre-planned expansion cohorts were designed with a single interim analysis for futility that would stop the trial if 0 responses were seen in the first 14 patients enrolled.
-According to the trial protocol, this futility rule is associated with at most a 4.4\% chance of observing no responses in 14 patients if the true response rate is 20\% or higher.
-The protocol also states the widths of the 90\% confidence intervals for a sample size of 40 if the observed response rate is 30\%.
-There was no stated decision rule for efficacy since efficacy was not an explicit aim of the expansion cohorts.
-In the re-design we assume a null, or unacceptable, response rate of 0.1 and an alternative, or acceptable, response rate of 0.2.
-We plan a study with up to a total of 95 participants.
-In our sequential predictive probability design we will check for futility after every 5 patients are enrolled.
-We consider posterior thresholds of 0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, and 1, and predictive thresholds of 0.05, 0.1, 0.15, and 0.2.
+To demonstrate the functionality of the \pkg{ppseq} package, we focus on
+a re-design of a dose-expansion cohort for the study of atezolizumab in
+metastatic urothelial carcinoma patients (mUC) using sequential
+predictive probability monitoring. Atezolizumab is an anti-PD-L1
+treatment that was originally tested in the phase 1 setting in a basket
+trial across a variety of cancer sites harboring PD-L1 mutations. The
+atezolizumab expansion study in mUC had the primary aim of further
+evaluating safety, pharmacodynamics and pharmacokinetics and therefore
+was not designed to meet any specific criteria for type I error or
+power. An expansion cohort in mUC was not part of the original protocol
+design, but was rather added later through a protocol amendment. The
+expansion cohort in mUC ultimately enrolled a total of 95 participants
+\citep{Powles2014}. Other expansion cohorts that were included in the
+original protocol, including in renal-cell carcinoma, non-small-cell
+lung cancer, and melanoma, were planned to have a sample size of 40.
+These pre-planned expansion cohorts were designed with a single interim
+analysis for futility that would stop the trial if 0 responses were seen
+in the first 14 patients enrolled. According to the trial protocol, this
+futility rule is associated with at most a 4.4\% chance of observing no
+responses in 14 patients if the true response rate is 20\% or higher.
+The protocol also states the widths of the 90\% confidence intervals for
+a sample size of 40 if the observed response rate is 30\%. There was no
+stated decision rule for efficacy since efficacy was not an explicit aim
+of the expansion cohorts. In the re-design we assume a null, or
+unacceptable, response rate of 0.1 and an alternative, or acceptable,
+response rate of 0.2. We plan a study with up to a total of 95
+participants. In our sequential predictive probability design we will
+check for futility after every 5 patients are enrolled. We consider
+posterior thresholds of 0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93,
+0.94, 0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, and 1, and
+predictive thresholds of 0.05, 0.1, 0.15, and 0.2.
 
@@
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 install.packages("ppseq")
-```
+\end{Sinput}
+\end{Schunk}
 
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 library(ppseq)
-```
+\end{Sinput}
+\end{Schunk}
 
-We use the \code{calibrate\_thresholds()} function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
-Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
-The user will be responsible for setting up a call to \code{future::plan()} that is appropriate to their operating environment and simulation setting.
-The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid.
-Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \pkg{ppseq} package.
+We use the \code{calibrate\_thresholds()} function to obtain the
+operating characteristics of designs based on on each combination of
+posterior and predictive thresholds. Because of the inherent computation
+intensity in these calculations, this function relies on the
+\CRANpkg{future} \citep{Bengtsson2020} and \CRANpkg{furrr}
+\citep{Vaughan2021} packages to parallelize computations. The user will
+be responsible for setting up a call to \code{future::plan()} that is
+appropriate to their operating environment and simulation setting. The
+example in this case study was run on a Unix server with 192 cores, and
+we wished to use 76 cores to accommodate the 76 distinct designs that
+result from the 19 posterior by 4 predictive threshold grid. Because the
+code takes some time to run, the results of the below example code are
+available as a dataset called \code{one\_sample\_cal\_tbl} included in
+the \pkg{ppseq} package.
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 library(future)
@@
     )
-```
+\end{Sinput}
+\end{Schunk}
 
-Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.01 and 0.2, and a minimum power of 0.7.
-We find that 35 of the 76 designs meet these criteria for type I error and power.
+Next we print the results table using the \code{print()} option, and
+limited to designs with type I error between 0.01 and 0.2, and a minimum
+power of 0.7. We find that 35 of the 76 designs meet these criteria for
+type I error and power.
 
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 print(
@@
   )
-```
+\end{Sinput}
+\begin{Soutput}
+#> # A tibble: 35 x 8
+#>    pp_threshold ppp_threshold mean_n1_null prop_pos_null prop_stopped_null
+#>           <dbl>         <dbl>        <dbl>         <dbl>             <dbl>
+#>  1         0.7           0.1          18.2         0.199             0.543
+#>  2         0.7           0.15         16.7         0.182             0.728
+#>  3         0.7           0.2          11.6         0.139             0.801
+#>  4         0.74          0.1          18.1         0.199             0.544
+#>  5         0.74          0.15         16.7         0.182             0.728
+#>  6         0.74          0.2          11.6         0.139             0.801
+#>  7         0.78          0.1          18.1         0.200             0.544
+#>  8         0.78          0.15         16.7         0.182             0.728
+#>  9         0.78          0.2          11.6         0.139             0.801
+#> 10         0.82          0.1          18.1         0.200             0.544
+#> # ... with 25 more rows, and 3 more variables: mean_n1_alt <dbl>,
+#> #   prop_pos_alt <dbl>, prop_stopped_alt <dbl>
+\end{Soutput}
+\end{Schunk}
 
-We use the \code{optimize\_designs()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
-We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.
-It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91.
-The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1.
-It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82.
-For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.
+We use the \code{optimize\_designs()} function to obtain the details of
+the optimal accuracy and optimal efficiency designs, limited to our
+desired range of type I error and minimum power. We find that the
+optimal accuracy design is the one with posterior threshold 0.9 and
+predictive threshold 0.05. It has a type I error of 0.072, power of
+0.883, average sample size under the null of 51, and average sample size
+under the alternative of 91. The optimal efficiency design is the one
+with posterior threshold of 0.92 and predictive threshold of 0.1. It has
+a type I error of 0.06, power of 0.796, average sample size under the
+null of 39, and average sample size under the alternative of 82. For
+comparison, the original design of the atezolizumab expansion cohort in
+mUC, with a single look for futility after the first 14 patients, has a
+type I error of 0.005, power of 0.528, average sample size under the
+null of 76, and average sample size under the alternative of 92.
 
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 optimize_design(
@@
   )
-```
+\end{Sinput}
+\begin{Soutput}
+#> $`Optimal accuracy design:`
+#> # A tibble: 1 x 6
+#>   pp_threshold ppp_threshold `Type I error` Power `Average N under the null`
+#>          <dbl>         <dbl>          <dbl> <dbl>                      <dbl>
+#> 1         0.93           0.1         0.0872 0.890                       16.7
+#>   `Average N under the alternative`
+#>                               <dbl>
+#> 1                              24.3
+#> 
+#> $`Optimal efficiency design:`
+#> # A tibble: 1 x 6
+#>   pp_threshold ppp_threshold `Type I error` Power `Average N under the null`
+#>          <dbl>         <dbl>          <dbl> <dbl>                      <dbl>
+#> 1         0.93          0.15         0.0692 0.780                       11.6
+#>   `Average N under the alternative`
+#>                               <dbl>
+#> 1                              21.5
+\end{Soutput}
+\end{Schunk}
 
-To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations.
+To compare these optimal designs with all other designs, we can use the
+\code{plot()} function with the \code{plotly = TRUE} option to obtain
+interactive visualizations.
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 plot(
@@
   )
-```
+\end{Sinput}
+\end{Schunk}
 
-```{r echo = FALSE, fig.width = 9, fig.height = 6}
-# Note that the above code does not render properly in R Markdown, though it does produce the plots correctly in an interactive session
-p <- plot(
-  one_sample_cal_tbl, 
-  type1_range = c(0.01, 0.2), 
-  minimum_power = 0.7,
-  plotly = TRUE
-  )
+\begin{Schunk}
 
-p[[1]]
-p[[2]]
-```
+\includegraphics{zabor-ppseq_files/figure-latex/unnamed-chunk-7-1} 
+\includegraphics{zabor-ppseq_files/figure-latex/unnamed-chunk-7-2} \end{Schunk}
 
-In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796.
-This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment.
-Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial.
-Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package.
-In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial.
-After the first 10 patients we would stop the trial if there were 0 responses, and so on.
-At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were >=14 responses.
+In this case we may choose to use the optimal efficiency design, which
+has the desirable trait of a very small average sample size under the
+null of just 39 patients, while still maintaining reasonable type I
+error of 0.06 and power of 0.796. This design would allow us to stop
+early if the treatment were inefficacious, thus preserving valuable
+financial resources for use in studying more promising treatments and
+preventing our human subjects from continuing an ineffective treatment.
+Finally, we generate the decision table associated with the selected
+design for use in making decision at each interim analysis during the
+conduct of the trial. Because of the computational time involved, the
+results of the below example code are available as a dataset called
+\code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package.
+In the results table, we see that at the first interim futility look
+after just 5 patients, we would not stop the trial. After the first 10
+patients we would stop the trial if there were 0 responses, and so on.
+At the end of the trial when all 95 patients have accrued, we would
+declare the treatment promising of further study if there were
+\textgreater=14 responses.
 
-```{r eval = FALSE}
+\begin{Schunk}
+\begin{Sinput}
 set.seed(123)
@@
     )
-```
+\end{Sinput}
+\end{Schunk}
 
-```{r}
+\begin{Schunk}
+\begin{Sinput}
 one_sample_decision_tbl
-```
+\end{Sinput}
+\begin{Soutput}
+#> # A tibble: 19 x 2
+#>        n     r
+#>    <dbl> <int>
+#>  1     5    NA
+#>  2    10     0
+#>  3    15     0
+#>  4    20     1
+#>  5    25     1
+#>  6    30     2
+#>  7    35     2
+#>  8    40     3
+#>  9    45     4
+#> 10    50     4
+#> 11    55     5
+#> 12    60     6
+#> 13    65     7
+#> 14    70     8
+#> 15    75     8
+#> 16    80     9
+#> 17    85    10
+#> 18    90    11
+#> 19    95    13
+\end{Soutput}
+\end{Schunk}
 
-## Summary
+\hypertarget{summary}{%
+\subsection{Summary}\label{summary}}
 
-With the focus of early stage clinical trial research in oncology shifting away from the study of cytotoxic treatments and toward immunotherapies and other non-cytotoxic treatments, new approaches to clinical trial design are needed that move beyond the traditional search for the maximum tolerated dose.
-Bayesian sequential predictive probability monitoring provides a natural and flexible way to expand the number of patients studied in phase 1 or to design phase 2 trials that allow for efficient early stopping for futility while maintaining control of type I error and power.
-The \pkg{ppseq} package implements functionality to consider a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e.
-type I error and power) or efficiency (i.e.
-average sample sizes under the null and alternative).
-Interactive visualization options are provided to ease comparison of the resulting design options.
-Once an ideal design is selected, a table of decision rules can be obtained to make trial conduct simple and straightforward.
+With the focus of early stage clinical trial research in oncology
+shifting away from the study of cytotoxic treatments and toward
+immunotherapies and other non-cytotoxic treatments, new approaches to
+clinical trial design are needed that move beyond the traditional search
+for the maximum tolerated dose. Bayesian sequential predictive
+probability monitoring provides a natural and flexible way to expand the
+number of patients studied in phase 1 or to design phase 2 trials that
+allow for efficient early stopping for futility while maintaining
+control of type I error and power. The \pkg{ppseq} package implements
+functionality to consider a range of posterior and predictive thresholds
+for a given study design and identify the optimal design based on
+accuracy (i.e.~type I error and power) or efficiency (i.e.~average
+sample sizes under the null and alternative). Interactive visualization
+options are provided to ease comparison of the resulting design options.
+Once an ideal design is selected, a table of decision rules can be
+obtained to make trial conduct simple and straightforward.
+
+\bibliography{zabor-ppseq.bib}
+
+\address{%
+Emily C. Zabor\\
+Department of Quantitative Health Sciences \& Taussig Cancer Institute,
+Cleveland Clinic\\%
+9500 Euclid Ave. CA-60\\ Cleveland, OH 44195 USA\\
+%
+\url{http://www.emilyzabor.com/}\\%
+\textit{ORCiD: \href{https://orcid.org/0000-0002-1402-4498}{0000-0002-1402-4498}}\\%
+\href{mailto:zabore2@ccf.org}{\nolinkurl{zabore2@ccf.org}}%
+}
+
+\address{%
+Brian P. Hobbs\\
+Dell Medical School, The University of Texas at Austin\\%
+true\\ Austin, TX 78712\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0003-2189-5846}{0000-0003-2189-5846}}\\%
+\href{mailto:brian.hobbs@austin.utexas.edu}{\nolinkurl{brian.hobbs@austin.utexas.edu}}%
+}
+
+\address{%
+Alexander M. Kaizer\\
+Department of Biostatistics \& Informatics, University of
+Colorado-Anschutz Medical Campus,\\%
+13001 E. 17th Place Campus Box B119\\ Aurora, CO 80045\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0003-2334-5514}{0000-0003-2334-5514}}\\%
+\href{mailto:alex.kaizer@cuanschutz.edu}{\nolinkurl{alex.kaizer@cuanschutz.edu}}%
+}
+
+\address{%
+Michael J. Kane\\
+Department of Biostatistics, Yale University\\%
+22 Mill Pond Drive\\ Guilford, CT 06437\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0003-1899-6662}{0000-0003-1899-6662}}\\%
+\href{mailto:michael.kane@yale.edu}{\nolinkurl{michael.kane@yale.edu}}%
+}

### 2021-06-23 R/calc_decision_rules.R

@@ -9,21 +9,21 @@
 #' \code{optimize_design}.
 #'
 #' @param n integer of sample size so far for one-sample case
-#' @param direction For one-sample case, "greater" if interest is in P(p > p0)
-#' and "less" if interest is in P(p < p0).
+#' @param N tinteger of total planned sample size at end of trial N
+#' for one-sample case
 #' @param p0 The target value to compare to in the one-sample case
 #' (i.e. the null response rate)
+#' @param theta The target posterior probability. e.g. Efficacy decision if
+#' P(p > p0) > theta for the one-sample case with greater direction.
+#' @param ppp The target predictive probability. e.g. Stop the trial if the
+#' predictive probability falls below this target.
+#' @param direction For one-sample case, "greater" if interest is in P(p > p0)
+#' and "less" if interest is in P(p < p0).
 #' @param delta clinically meaningful difference between groups.
 #' Typically 0 for two-sample case. NULL for one-sample case (default). 
 #' @param prior hyperparameters of prior beta distribution.
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
-#' @param N tinteger of total planned sample size at end of trial N
-#' for one-sample case
-#' @param theta The target posterior probability. e.g. Efficacy decision if
-#' P(p > p0) > theta for the one-sample case with greater direction.
-#' @param ppp The target predictive probability. e.g. Stop the trial if the
-#' predictive probability falls below this target.
 #'
 #' @return Returns a tibble with n at each look and r. Stop the trial at that
 #' look if the number of observed responses is <=r. At the end of the trial,
@@ -41,9 +41,9 @@
 #' @importFrom tibble tibble
 #' @export
 
-calc_decision_rules <- function(n, direction = "greater", p0,
-                                delta = NULL, prior = c(0.5, 0.5), S = 5000,
-                                N, theta, ppp) {
+calc_decision_rules <- function(n, N, theta, ppp, p0,
+                                direction = "greater", delta = NULL, 
+                                prior = c(0.5, 0.5), S = 5000) {
   if ((is.null(p0) & is.null(delta)) | (!is.null(p0) & !is.null(delta)))
     stop("Exactly one of delta or p0 must be specified for the ", 
          "two-sample and one-sample case, respectively")

### 2021-06-23 inst/ppseq-package-paper/zabor-ppseq.R

@@ -1,0 +1,90 @@
+## ----eval = FALSE-------------------------------------------------------------
+#> install.packages("ppseq")
+
+
+## -----------------------------------------------------------------------------
+library(ppseq)
+
+
+## ----eval = FALSE-------------------------------------------------------------
+#> library(future)
+#> 
+#> set.seed(123)
+#> 
+#> plan(multicore(workers = 76))
+#> 
+#> one_sample_cal_tbl <-
+#>   calibrate_thresholds(
+#>     p_null = 0.1,
+#>     p_alt = 0.2,
+#>     n = seq(5, 95, 5),
+#>     N = 95,
+#>     pp_threshold = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94,
+#>                      0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1),
+#>     ppp_threshold = seq(0.05, 0.2, 0.05),
+#>     direction = "greater",
+#>     delta = NULL,
+#>     prior = c(0.5, 0.5),
+#>     S = 5000,
+#>     nsim = 1000
+#>     )
+
+
+## -----------------------------------------------------------------------------
+print(
+  one_sample_cal_tbl,
+  type1_range = c(0.01, 0.2),
+  minimum_power = 0.7
+  )
+
+
+## -----------------------------------------------------------------------------
+optimize_design(
+  one_sample_cal_tbl, 
+  type1_range = c(0.05, 0.1), 
+  minimum_power = 0.7
+  )
+
+
+## ----eval = FALSE-------------------------------------------------------------
+#> plot(
+#>   one_sample_cal_tbl,
+#>   type1_range = c(0.01, 0.2),
+#>   minimum_power = 0.7,
+#>   plotly = TRUE
+#>   )
+
+
+## ----echo = FALSE, fig.width = 9, fig.height = 6------------------------------
+# Note that the above code does not render properly in R Markdown, though it does produce the plots correctly in an interactive session
+p <- plot(
+  one_sample_cal_tbl, 
+  type1_range = c(0.01, 0.2), 
+  minimum_power = 0.7,
+  plotly = TRUE
+  )
+
+p[[1]]
+p[[2]]
+
+
+## ----eval = FALSE-------------------------------------------------------------
+#> set.seed(123)
+#> 
+#> one_sample_decision_tbl <-
[…]

### 2021-08-05 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
adding a new version of the paper

--- 2021-06-23 inst/ppseq-package-paper/zabor-ppseq.Rmd
@@
 title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
+abstract: >
+  Clinical trials in oncology are making increasing use of larger phase 1 and phase 2 sample sizes with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
+With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
+The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability.
+Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
+draft: true
 author:  
@@
     - Austin, TX 78712
-  - name: Alexander M. Kaizer
-    email: alex.kaizer@cuanschutz.edu
-    orcid: 0000-0003-2334-5514
-    affiliation: Department of Biostatistics & Informatics, University of Colorado-Anschutz Medical Campus,  
-    address:
-    - 13001 E. 17th Place Campus Box B119
-    - Aurora, CO 80045
   - name: Michael J. Kane
@@
     - Guilford, CT 06437
-abstract: >
-  Clinical trials in oncology are making increasing use of larger phase 1 and phase 2 sample sizes with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
-With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
-The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability.
-Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
-preamble: |
-  % Any extra LaTeX you need in the preamble
-
-# per R journal requirement, the bib filename should be the same as the output 
-# tex file. Don't forget to rename the bib file and change this example value.
-bibliography: zabor-ppseq.bib
-
-output: rticles::rjournal_article
+date: "`r Sys.Date()`"
+creative_commons: CC BY
+output: 
+  - rjtools::rjournal_pdf_article
+  - rjtools::rjournal_web_article
+bibliography: zabor-hobbs-kane.bib
+always_allow_html: true
 ---
 
-## Introduction
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
+library(plotly)
+library(ggplot2)
+```
+
+# Introduction
 
@@
 
-## Predictive probability monitoring
+# Predictive probability monitoring
 
@@
 
-## Package overview
+# Package overview
 
@@
 
-### Optimization
+## Optimization
 
@@
 
-### Visualizations
+## Decision rules
 
-To assist users in comparing the results of the various design options, a \code{plot()} option is also available, that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020].
+To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function.
+The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, the predictive threshold of the selected design \code{ppp}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}.
+The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case).
+The function results in a \code{tibble}.
+In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}.
+The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$.
+At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.
+In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively, columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively,and a column for the posterior predictive probability associated with that decision \code{ppp}.
+
+## Visualizations
+
+Finally, to assist users in comparing the results of the various design options, a \code{plot()} option is available for the results of \code{calibrate\_thresholds} that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020].
 Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design.
@@
 Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot.
+A \code{plot()} option is also available for the results of \code{calc\_decision\_rules}.
+In the one-sample case it produces a single plot showing the sample size at each interim analysis on the x-axis and the possible number of responses at each interim analysis on the y-axis.
+In the two-sample case a grid of plots is produced, with one plot for each interim analysis.
+The x-axis shows the number of possible responses in the control group and the y-axis shows the number of possible responses in the experimental group.
+In both cases, the boxes are colored green for a "proceed" decision and red for a "stop" decision for each combination and the hover box produced by \CRANpkg{plotly} provides the details.
 
-### Decision rules
-
-Finally, to ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function.
-The function takes a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0}, the posterior threshold of the selected design \code{theta}, the predictive threshold of the selected design \code{ppp}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}.
-The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
-The function results in a \code{tibble} with the sample size at each interim analysis and the decision point $r$.
-The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$.
-At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.
-
-## Case study
+# Case study
 
@@
 
-## Summary
+```{r message = FALSE, fig.height = 10, fig.width = 8}
+plot(one_sample_decision_tbl)
+```
+
+# Summary
 

### 2021-08-05 R/calc_decision_rules.R

@@ -35,10 +35,18 @@
 #' Beta(0.5, 0.5) is default
 #' @param S number of samples, default is 5000
 #'
-#' @return Returns a tibble with n at each look and r. Stop the trial at that
-#' look if the number of observed responses is <=r. At the end of the trial,
-#' the treatment is considered promising if the number of observed responses 
-#' is >r.
+#' @return In the one-sample case, returns a tibble with n at each look, r at
+#' each look, and ppp, the associated posterior predictive probability. 
+#' Stop the trial at that look if the number of observed responses is <=r. 
+#' At the end of the trial, the treatment is considered promising if the number 
+#' of observed responses is >r. 
+#' In the two-sample case, returns a tibble with n0 and n1, the number 
+#' enrolled subjects in the control and experimental arms at each look, 
+#' respectively, r0 and r1, the number of possible responses in the control and 
+#' experimental arms at each look, respectively, and ppp, the associated 
+#' posterior predictive probability. Any NA value in either table represents an
+#' interim look where there is no number of responses that would lead to 
+#' stopping the trial.
 #'
 #' @examples
 #' \dontrun{

### 2021-08-05 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -1,3 +1,9 @@
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
+library(plotly)
+library(ggplot2)
+
+
 ## ----eval = FALSE-------------------------------------------------------------
 #> install.packages("ppseq")
 
@@ -87,4 +93,8 @@ p[[2]]
 
 ## -----------------------------------------------------------------------------
 one_sample_decision_tbl
+
+
+## ----message = FALSE, fig.height = 10, fig.width = 8--------------------------
+plot(one_sample_decision_tbl)
 

### 2021-08-10 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
Plot testing.

@@
     address:
-    - 22 Mill Pond Drive
-    - Guilford, CT 06437
+    - 60 College Street
+    - New Haven, CT 06511
 date: "`r Sys.Date()`"

### 2021-08-10 R/plot.R

@@ -21,6 +21,7 @@
 #' @importFrom ggplot2 ggplot aes geom_point xlim ylim labs 
 #'                     scale_color_viridis_c theme_bw theme
 #' @importFrom plotly ggplotly
+#' @import patchwork
 #' @export
 plot.calibrate_thresholds <- function(x,
                                       type1_range = c(0.05, 0.1),
@@ -208,6 +209,7 @@ plot.calibrate_thresholds <- function(x,
 #' 
 #' @param x an object of class 'calc_decision_rules', usually returned by the
 #' \code{calc_decision_rules} function
+#' @param plotly should the plot be rendered in plotly? (Default is TRUE)
 #' @param ... unused
 #' 
 #' @importFrom dplyr mutate filter select rename ungroup group_by 
@@ -221,7 +223,7 @@ plot.calibrate_thresholds <- function(x,
 #' @importFrom plotly ggplotly
 #' 
 #' @export
-plot.calc_decision_rules <- function(x, ...) {
+plot.calc_decision_rules <- function(x, plotly = TRUE, ...) {
   
   if (any(class(x) == "calc_decision_rules") == FALSE)
     stop("x must be class 'calc_decision_rules', usually an object returned from a call to the function ppseq::calc_decision_rules()")
@@ -342,7 +344,9 @@ plot.calc_decision_rules <- function(x, ...) {
       theme_bw() +
       xlab("N at interim analysis")
   }
-  
-  ggplotly(p)
+
+  if (plotly) {  
+    ggplotly(p)
+  }
   
 }

### 2021-08-10 tests/testthat/test-plot.R

@@ -1,0 +1,112 @@
+library(vdiffr)
+
+test_that(
+  "one-sample calibrate threshold plotting",
+  {
+    expect_error(plot.calibrate_thresholds(3))
+
+    set.seed(123)
+    one_sample_cal_tbl <- 
+      calibrate_thresholds(
+          p_null = 0.1,
+          p_alt = 0.3,
+          n = c(5, 25),
+          N = 25,
+          pp_threshold = c(0.2, 0.9),
+          ppp_threshold = c(0.05, 0.2),
+          direction = "greater",
+          delta = NULL,
+          prior = c(0.5, 0.5),
+          S = 200,
+          nsim = 100
+        )
+
+    expect_error(plot(one_sample_cal_tbl, type1_range = 3))
+
+    expect_error(
+      plot(one_sample_cal_tbl, type1_range = c("a", "b"))
+    )
+
+    expect_error(
+      plot(one_sample_cal_tbl, minimum_power = c(1, 2))
+    )
+
+    expect_error(plot(one_sample_cal_tbl, minimum_power = "a"))
+
+    expect_doppelganger('plot_one_sample_cal_tbl', plot(one_sample_cal_tbl))
+  }
+)
+
+# Emily, can you provide an example where this works and plots?
+#test_that(
+#  "two-sample calibrate threshold plotting",
+#  {
+#    set.seed(123)
+#    two_sample_cal_tbl <- 
+#      calibrate_thresholds(
+#          p_null = c(0.1, 0.1),
+#          p_alt = c(0.5, 0.6),
+#          n = cbind(c(50, 75), c(50, 75)),
+#          N = c(100, 100),
+#          pp_threshold = 0.4,
+#          ppp_threshold = 0.2,
+#          direction = "greater",
+#          delta = 0,
+#          prior = c(0.5, 0.5),
+#          S = 200,
+#          nsim = 200
+#      )
+#
+#    expect_doppelganger('plot_two_sample_cal_tbl', plot(two_sample_cal_tbl))
+#
+#  }
+#)
+
+test_that(
+  'plot one-sample decision rules works',
+  {
+    set.seed(123)
+    cd1 <- 
+      calc_decision_rules(
+        n = c(10, 20),
+        N = 25,
+        theta = 0.86,
+        ppp = 0.2,
+        p0 = 0.1,
+        direction = "greater",
+        delta = NULL,
+        prior = c(0.5, 0.5),
+        S = 100
+      )
[…]

### 2021-08-18 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
updates to the paper Rmd

@@
 But dose-expansion cohorts are not always planned in advance, and therefore may be subject to on-the-fly decision making that can lead to large sample sizes and poor statistical properties.
-For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 495 non-small-cell lung cancer patients across six disease-specific expansion cohorts [@Garon2015].
+For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015].
 In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 95 patients, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol but was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
 
-Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error, and has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; Lee2008; Saville2014].
+Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error, and has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014].
 However, in order to be useful to investigators designing such trials, software must be made available that streamlines design.
@@
 
-We use the \code{optimize\_designs()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
+We use the \code{optimize\_design()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
 We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.

### 2021-08-20 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
updates to the RJournal paper

@@
 abstract: >
-  Clinical trials in oncology are making increasing use of larger phase 1 and phase 2 sample sizes with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
+  Advances in drug discovery have produced numerous biomarker-guided therapeutic strategies for treating cancer.
+Yet the promise of precision medicine comes with the cost of increased complexity.
+Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
 With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
@@
     - New Haven, CT 06511
-date: "`r Sys.Date()`"
+date: "August 20, 2021"
+type: package
 creative_commons: CC BY
 output: 
-  - rjtools::rjournal_pdf_article
-  - rjtools::rjournal_web_article
+  rjtools::rjournal_web_article
 bibliography: zabor-hobbs-kane.bib
-always_allow_html: true
+
 ---
@@
 Designs for dose-escalation trials include the rule-based 3+3 design and the model-based continual reassessment method, among others.
-But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs.
+But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs [@Pestana2020].
 As a result, it is becoming increasingly common to include dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete.
@@
 For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015].
-In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 95 patients, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol but was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
+In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 97 patients and evaluated 95, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol.
+The expansion cohort in metastatic urothelial carcinoma was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
 
-Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error, and has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014].
-However, in order to be useful to investigators designing such trials, software must be made available that streamlines design.
+Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error.
+As a result, predictive probability has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014; @Hobbs2018].
+However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design.
 This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring.
-Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} assist investigators in comparing designs based on different thresholds for decision making, and we suggest optimization criteria for selecting the ideal design.
+Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} compare designs based on different thresholds for decision making.
+Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability design using the \pkg{ppseq} package.
+While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential clinical trial design.
 
@@
 Consider the setting of a binary outcome, such as tumor response as measured by the RECIST criteria in the setting of a study in patients with solid tumors, where each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$.
-Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total enrolled patients.
+Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients.
 Let $p$ represent the probability of response, where $p_0$ represents the null response rate under no treatment or the standard of care treatment and $p_1$ represents the alternative response rate under the experimental treatment.
@@
 The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds.
-Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}.
+Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}.
+Default arguments are available for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}, but customization is also possible.
 The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
@@
 
-After obtaining results for the all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options.
+After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options.
 The \pkg{ppseq} package introduces two optimization criteria to assist users in making a selection.
@@
 To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function.
-The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, the predictive threshold of the selected design \code{ppp}, the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}.
+The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, and the predictive threshold of the selected design \code{ppp}.
+Default arguments are provided for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}, though customization is also possible.
 The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case).
@@
 At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.
-In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively, columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively,and a column for the posterior predictive probability associated with that decision \code{ppp}.
+In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively.
+There are also columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively.
+Finally, there is a column for the posterior predictive probability associated with that decision \code{ppp}.
 
@@
 An expansion cohort in mUC was not part of the original protocol design, but was rather added later through a protocol amendment.
-The expansion cohort in mUC ultimately enrolled a total of 95 participants [@Powles2014].
+The expansion cohort in mUC ultimately evaluated a total of 95 participants [@Powles2014].
 Other expansion cohorts that were included in the original protocol, including in renal-cell carcinoma, non-small-cell lung cancer, and melanoma, were planned to have a sample size of 40.
@@
 
-```{r echo = FALSE, fig.width = 9, fig.height = 6}
-# Note that the above code does not render properly in R Markdown, though it does produce the plots correctly in an interactive session
+```{r echo = FALSE, fig.width = 6, fig.height = 4, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot.
+The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
+p <- plot(
+  one_sample_cal_tbl,
+  type1_range = c(0.01, 0.2),
+  minimum_power = 0.7,
+  plotly = FALSE
+  )
+p
+```
+
+```{r echo = FALSE, fig.width = 6, fig.height = 4, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of design options made with plotly.
+The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
 p <- plot(
@@
   )
-
 p[[1]]
@@
 
-```{r message = FALSE, fig.height = 10, fig.width = 8}
+```{r message = FALSE, fig.height = 8, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
+The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
+plot(one_sample_decision_tbl, plotly = FALSE)
+```
+
+```{r message = FALSE, fig.height = 8, fig.width = 6, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
+The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
 plot(one_sample_decision_tbl)
@@
 
-With the focus of early stage clinical trial research in oncology shifting away from the study of cytotoxic treatments and toward immunotherapies and other non-cytotoxic treatments, new approaches to clinical trial design are needed that move beyond the traditional search for the maximum tolerated dose.
+With the focus of early stage clinical trial research in oncology shifting away from the study of cytotoxic treatments and toward immunotherapies and other non-cytotoxic treatments, new approaches to clinical trial design are needed that move beyond the traditional search for the maximum tolerated dose [@Hobbs2019].
 Bayesian sequential predictive probability monitoring provides a natural and flexible way to expand the number of patients studied in phase 1 or to design phase 2 trials that allow for efficient early stopping for futility while maintaining control of type I error and power.
-The \pkg{ppseq} package implements functionality to consider a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e.
+The \pkg{ppseq} package implements functionality to evaluate a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e.
 type I error and power) or efficiency (i.e.

### 2021-08-20 R/plot.R

@@ -325,7 +325,8 @@ plot.calc_decision_rules <- function(x, plotly = TRUE, ...) {
               r <= stop_criteria ~ "Stop",
               r > stop_criteria ~ "Proceed",
               is.na(stop_criteria) ~ "Proceed"
-              )
+              ), 
+            n = as.factor(n)
             )
           ),
         "# responses" = r,
@@ -333,7 +334,7 @@ plot.calc_decision_rules <- function(x, plotly = TRUE, ...) {
       )
     
     p <- 
-      ggplot(plotdat, aes(x = as.factor(`N at interim analysis`), 
+      ggplot(plotdat, aes(x = `N at interim analysis`, 
                            y = `# responses`, 
                            fill = Decision)) + 
       geom_tile(color = "black") + 
@@ -347,6 +348,6 @@ plot.calc_decision_rules <- function(x, plotly = TRUE, ...) {
 
   if (plotly) {  
     ggplotly(p)
-  }
+  } else p
   
 }

### 2021-08-20 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -61,17 +61,25 @@ optimize_design(
 #>   )
 
 
-## ----echo = FALSE, fig.width = 9, fig.height = 6------------------------------
-# Note that the above code does not render properly in R Markdown, though it does produce the plots correctly in an interactive session
+## ----echo = FALSE, fig.width = 7, fig.height = 5, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
 p <- plot(
-  one_sample_cal_tbl, 
-  type1_range = c(0.01, 0.2), 
+  one_sample_cal_tbl,
+  type1_range = c(0.01, 0.2),
   minimum_power = 0.7,
-  plotly = TRUE
+  plotly = FALSE
   )
+p
 
-p[[1]]
-p[[2]]
+
+## ----echo = FALSE, fig.width = 6, fig.height = 4, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of design options made with plotly. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
+#> p <- plot(
+#>   one_sample_cal_tbl,
+#>   type1_range = c(0.01, 0.2),
+#>   minimum_power = 0.7,
+#>   plotly = TRUE
+#>   )
+#> p[[1]]
+#> p[[2]]
 
 
 ## ----eval = FALSE-------------------------------------------------------------
@@ -95,6 +103,12 @@ p[[2]]
 one_sample_decision_tbl
 
 
-## ----message = FALSE, fig.height = 10, fig.width = 8--------------------------
-plot(one_sample_decision_tbl)
+## ----message = FALSE, fig.height = 8, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."----
+plot(one_sample_decision_tbl, plotly = FALSE)
 
+
+## ----message = FALSE, fig.height = 8, fig.width = 6, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."----
+#> plot(one_sample_decision_tbl)
+
+```{.r .distill-force-highlighting-css}
+```

### 2021-08-27 R/one_sample_cal_tbl-data.R
adding updated data files

@@ -21,6 +21,9 @@
 #' 2) 'call_list' containing the original function call
 #' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the
 #' original function call
+#' 4) 'protocol_res' gives results from the protocol-specified design of the 
+#' atezolizumab case study, as detailed in the vignette titled "One-sample 
+#' expansion cohort"
 #'
 #' @keywords datasets
 "one_sample_cal_tbl"

### 2021-08-27 data-raw/one_sample_cal_tbl.R
updating the decision table dataset

@@ -1,0 +1,331 @@
+## code to prepare `one_sample_cal_tbl` dataset goes here
+
+# Note that this takes a long time to run and needs to be run on the server
+
+library(ppseq)
+library(tibble)
+library(dplyr)
+library(purrr)
+library(furrr)
+
+eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N, 
+                        direction = "greater", delta = NULL,
+                        prior = c(0.5, 0.5), S = 5000) {
+  decision <- NULL
+  ppp <- NULL
+  for (i in 1:nrow(data)) {
+    if (ncol(data) == 4) {
+      ppp[i] <- calc_predictive(
+        y = c(data$y0[i], data$y1[i]),
+        n = c(data$n0[i], data$n1[i]),
+        direction = direction, 
+        p0 = p0, 
+        delta = delta,
+        prior = prior, 
+        S = S, 
+        N = N,
+        theta = pp_threshold
+      )
+    } else if (ncol(data) == 2) {
+      ppp[i] <- calc_predictive(
+        y = data$y1[i], 
+        n = data$n1[i],
+        direction = direction, 
+        p0 = p0, 
+        delta = delta,
+        prior = prior, 
+        S = S, 
+        N = N,
+        theta = pp_threshold
+      )
+    }
+    decision[i] <- ppp[i] < ppp_threshold
+    if (decision[i] == TRUE) break
+  }
+  res0 <- add_column(
+    data[ifelse(any(decision == TRUE),
+                which(decision == TRUE),
+                length(decision)
+    ), ],
+    pp_threshold = pp_threshold,
+    ppp_threshold = ppp_threshold,
+    ppp = ppp[ifelse(any(decision == TRUE),
+                     which(decision == TRUE),
+                     length(decision)
+    )]
+  )
+  
+  if (ncol(data) == 4) {
+    res <- mutate(
+      res0,
+      positive = case_when(
+        sum(n0, n1) == sum(N) & ppp > pp_threshold ~ TRUE,
+        sum(n0, n1) == sum(N) & ppp <= pp_threshold ~ FALSE,
+        sum(n0, n1) != sum(N) ~ FALSE
+      )
+    )
+  } else if (ncol(data) == 2) {
+    res <- mutate(
+      res0,
+      positive = case_when(
+        n1 == N & ppp > pp_threshold ~ TRUE,
+        n1 == N & ppp <= pp_threshold ~ FALSE,
+        n1 != N ~ FALSE
+      )
[…]

### 2021-08-27 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
updating R Journal paper with new datasets

@@
 
-Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.01 and 0.2, and a minimum power of 0.7.
+Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7.
 We find that 35 of the 76 designs meet these criteria for type I error and power.
@@
   one_sample_cal_tbl,
-  type1_range = c(0.01, 0.2),
+  type1_range = c(0.05, 0.1),
   minimum_power = 0.7
@@
   one_sample_cal_tbl, 
-  type1_range = c(0.01, 0.2), 
+  type1_range = c(0.05, 0.1), 
   minimum_power = 0.7,
@@
   one_sample_cal_tbl,
-  type1_range = c(0.01, 0.2),
+  type1_range = c(0.05, 0.1),
   minimum_power = 0.7,
@@
 
-```{r message = FALSE, fig.height = 8, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
+```{r message = FALSE, fig.height = 9, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
@@
 
-```{r message = FALSE, fig.height = 8, fig.width = 6, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
+```{r message = FALSE, fig.height = 9, fig.width = 6, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}

### 2021-08-27 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -39,7 +39,7 @@ library(ppseq)
 ## -----------------------------------------------------------------------------
 print(
   one_sample_cal_tbl,
-  type1_range = c(0.01, 0.2),
+  type1_range = c(0.05, 0.1),
   minimum_power = 0.7
   )
 
@@ -55,16 +55,16 @@ optimize_design(
 ## ----eval = FALSE-------------------------------------------------------------
 #> plot(
 #>   one_sample_cal_tbl,
-#>   type1_range = c(0.01, 0.2),
+#>   type1_range = c(0.05, 0.1),
 #>   minimum_power = 0.7,
 #>   plotly = TRUE
 #>   )
 
 
-## ----echo = FALSE, fig.width = 7, fig.height = 5, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
+## ----echo = FALSE, fig.width = 6, fig.height = 4, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
 p <- plot(
   one_sample_cal_tbl,
-  type1_range = c(0.01, 0.2),
+  type1_range = c(0.05, 0.1),
   minimum_power = 0.7,
   plotly = FALSE
   )
@@ -74,7 +74,7 @@ p
 ## ----echo = FALSE, fig.width = 6, fig.height = 4, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of design options made with plotly. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
 #> p <- plot(
 #>   one_sample_cal_tbl,
-#>   type1_range = c(0.01, 0.2),
+#>   type1_range = c(0.05, 0.1),
 #>   minimum_power = 0.7,
 #>   plotly = TRUE
 #>   )
@@ -103,11 +103,11 @@ p
 one_sample_decision_tbl
 
 
-## ----message = FALSE, fig.height = 8, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."----
[…]

### 2022-01-13 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
paper edits and some man fixes

@@
 With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
-The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential monitoring based on the Bayesian predictive probability.
+The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability.
 Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
@@
 However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design.
-This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring.
-Interactive plots produced using \CRANpkg{ggplot2} and \CRANpkg{plotly} compare designs based on different thresholds for decision making.
-Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability design using the \pkg{ppseq} package.
-While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential clinical trial design.
+This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility.
+Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making.
+Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \pkg{ppseq} package.
+While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.
 
@@
 Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$.
-Using the posterior probability, which represents the probability of success based only on the data accrued so far, we would declare a treatment efficacious if $\Pr(p>p_0 | X) > \theta$, where $\theta$ represents a pre-specified posterior decision threshold.
+A posterior probability threshold $\theta$ would be pre-specified during the trial design stage.
+At the end of the trial if the posterior probability exceeded the pre-specified threshold, i.e.
+if $\Pr(p>p_0 | X) > \theta$, the trial would be declared a success.
+
 The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$.
 Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$.
-The posterior predictive probability represents the probability that the treatment will be declared efficacious at the end of the trial when full enrollment is reached.
+The posterior predictive probability represents the probability that, at any given interim monitoring point, the treatment will be declared efficacious at the end of the trial when full enrollment is reached, conditional on the currently observed data and the specified priors.
 We would stop the trial early for futility if the posterior predictive probability dropped below a pre-specified threshold $\theta^*$, i.e.
@@
 
-The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring.
+The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility.
+The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial.
 The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error.
@@
 In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}.
-The trial would stop at a given look if the number of observed responses is $\leq r$ otherwise the trial would continue enrolling if the number of observed responses is $>r$.
-At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is $>r$.
+The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$.
+At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$.
 In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively.
@@
 After the first 10 patients we would stop the trial if there were 0 responses, and so on.
-At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were >=14 responses.
+At the end of the trial when all 95 patients have accrued, we would declare the treatment promising of further study if there were greater than or equal to 14 responses.
 

### 2022-01-13 R/calibrate_thresholds.R

@@ -175,12 +175,11 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #' probability threshold with interim futility monitoring
 #'
 #' @description This function is meant to be used in the context of a
-#' clinical trial with a binary endpoint. For a vector of possible posterior
-#' decision thresholds, the function simulates many trials and then calculates
-#' the average number of times the posterior probability exceeds a given
-#' threshold. In a null case, this will result in the type I error at a given
-#' threshold. In an alternative case, this will result in the power at a given
-#' threshold.
+#' clinical trial with a binary endpoint. For every combination of the provided
+#' posterior thresholds and predictive thresholds, the function simulates many 
+#' trials and then calculates the average number of times a trial was positive. 
+#' In the null case, this is the type I error for the given thresholds. 
+#' In the alternative case, this is the power for the given thresholds.
 #'
 #' @param p_null vector of length two containing the probability of event in
 #' the standard of care and experimental arm c(p0, p1) for the two-sample case
@@ -203,8 +202,8 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #' for two-sample case; integer of total planned sample size at end of trial N
 #' for one-sample case
 #' @param pp_threshold the posterior probability threshold of interest
-#' @param ppp_threshold the posterior probability threshold of interest for
-#' futility monitoring
+#' @param ppp_threshold the posterior predictive probability threshold of 
+#' interest for futility monitoring
 #' @param direction "greater" (default) if interest is in p(p1 > p0) and "less"
 #' if interest is in p(p1 < p0) for two-sample case. For one-sample case,
 #' "greater" if interest is in p(p > p0) and "less" if interest is in p(p < p0).
@@ -295,10 +294,15 @@ calibrate_thresholds <- function(p_null, p_alt, n, N,
       function(x) 
         pmap_dfr(cross_threshold,
                  function(pp_threshold, ppp_threshold) 
-                   eval_thresh(x, pp_threshold, ppp_threshold,
-                 direction = direction, p0 = p0, 
-                 delta = delta, prior = prior, 
-                 S = S, N = N)), 
+                   eval_thresh(x, 
+                               pp_threshold, 
+                               ppp_threshold,
+                               direction = direction, 
+                               p0 = p0, 
+                               delta = delta, 
+                               prior = prior, 
+                               S = S, 
+                               N = N)), 
       .options = furrr_options(seed = TRUE)
     )
   
[…]

### 2022-01-13 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -33,7 +33,7 @@ library(ppseq)
 #>     prior = c(0.5, 0.5),
 #>     S = 5000,
 #>     nsim = 1000
-#>     )
+#>   )
 
 
 ## -----------------------------------------------------------------------------
@@ -41,7 +41,7 @@ print(
   one_sample_cal_tbl,
   type1_range = c(0.05, 0.1),
   minimum_power = 0.7
-  )
+)
 
 
 ## -----------------------------------------------------------------------------
@@ -49,7 +49,7 @@ optimize_design(
   one_sample_cal_tbl, 
   type1_range = c(0.05, 0.1), 
   minimum_power = 0.7
-  )
+)
 
 
 ## ----eval = FALSE-------------------------------------------------------------
@@ -58,7 +58,7 @@ optimize_design(
 #>   type1_range = c(0.05, 0.1),
 #>   minimum_power = 0.7,
 #>   plotly = TRUE
-#>   )
+#> )
 
 
 ## ----echo = FALSE, fig.width = 6, fig.height = 4, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
@@ -96,7 +96,7 @@ p
 #>     delta = NULL,
 #>     prior = c(0.5, 0.5),
 #>     S = 5000
-#>     )
+#>   )
 
 
 ## -----------------------------------------------------------------------------

### 2022-01-24 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
close to final paper draft for submission

@@
 However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design.
-This paper introduces the \pkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility.
+This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility.
 Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making.
-Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \pkg{ppseq} package.
-While the \pkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.
+Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package.
+While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.
 
@@
 
-The \pkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility.
+The \CRANpkg{ppseq} package facilitates the design of clinical trials utilizing sequential predictive probability monitoring for futility.
 The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial.
@@
 After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options.
-The \pkg{ppseq} package introduces two optimization criteria to assist users in making a selection.
+The \CRANpkg{ppseq} package introduces two optimization criteria to assist users in making a selection.
 The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power.
@@
 
-To demonstrate the functionality of the \pkg{ppseq} package, we focus on a re-design of a dose-expansion cohort for the study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring.
+To demonstrate the functionality of the \CRANpkg{ppseq} package, we focus on a re-design of a dose-expansion cohort for the study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring.
 Atezolizumab is an anti-PD-L1 treatment that was originally tested in the phase 1 setting in a basket trial across a variety of cancer sites harboring PD-L1 mutations.
@@
 
-First we install and load the \pkg{ppseq} package.
+First we install and load the \CRANpkg{ppseq} package.
 
@@
 The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid.
-Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \pkg{ppseq} package.
+Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \CRANpkg{ppseq} package.
 
@@
 
-To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations.
+To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations, or the \code{plotly = FALSE} option to obtain static visualizations.
 
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 4, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot.
+```{r echo = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot2.
+The accuracy designs are in the plot on the left and the efficiency designs are on the right.
 The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
@@
 
-```{r echo = FALSE, fig.width = 6, fig.height = 4, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of design options made with plotly.
-The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
+```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with plotly.
+The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond.
+A similar plot is available for the accuracy design options."}
+
+# Right now only including the efficiency design plot in the html paper version
 p <- plot(
@@
   )
-p[[1]]
+# p[[1]]
 p[[2]]
+```
+
+```{r, echo = FALSE, include=knitr::is_html_output(), eval=FALSE, fig.cap = "Plot of design options made with plotly.
+The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
+
+# Not currently evaluating this code chunk. I was trying to further manipulate the interactive plots to get them to be placed side-by-side, but I just don't think it's possible with two legends
+
+# Above instead I think it is best to only include one of the plots as an example in the html paper version
+library(dplyr)
+x <- one_sample_cal_tbl
+type1_range <- c(0.05, 0.1)
+minimum_power <- 0.7
+
+plot_x <-
+  rename(
+    mutate(
+      filter(
+        x$res_summary,
+        prop_pos_null >= type1_range[1] &
+          prop_pos_null <= type1_range[2] &
+          prop_pos_alt >= minimum_power
+      ),
+      Design = paste0(
+        "Posterior threshold = ", pp_threshold,
+        " and predictive threshold = ", ppp_threshold
+      ),
+      ab_dist_metric = ((prop_pos_null - 0)^2 +
+        (prop_pos_alt - 1)^2)^(1 / 2),
+      n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 +
+        (mean_n1_alt - max(mean_n1_alt))^2)^(1 / 2)
+    ),
+    `Type I error` = prop_pos_null,
+    Power = prop_pos_alt,
+    `Average N under the null` = mean_n1_null,
+    `Average N under the alternative` = mean_n1_alt,
+    `Distance to optimal efficiency` = n_dist_metric,
+    `Distance to optimal accuracy` = ab_dist_metric
+  )
+
+plot_ab <-
+  mutate(
+    ungroup(
+      slice(
+        group_by(
+          arrange(
+            plot_x,
+            `Distance to optimal accuracy`,
+            -pp_threshold,
+            -ppp_threshold
+          ),
+          `Distance to optimal accuracy`
+        ),
+        1
+      )
+    ),
+    optimal_accuracy = ifelse(row_number() == 1, TRUE, FALSE)
+  )
+
+plot_nn <-
+  mutate(
+    ungroup(
+      slice(
+        group_by(
+          arrange(
+            plot_x,
+            `Distance to optimal efficiency`,
+            -pp_threshold, -ppp_threshold
+          ),
+          `Distance to optimal efficiency`
+        ),
+        1
+      )
+    ),
+    optimal_efficiency = ifelse(row_number() == 1, TRUE, FALSE)
+  )
+
+p1 <-
+  ggplot(
+    plot_ab,
+    aes(
+      x = `Type I error`,
+      y = Power,
+      color = `Distance to optimal accuracy`,
+      Design = Design,
+      `Average N under the null` = `Average N under the null`,
+      `Average N under the alternative` = `Average N under the alternative`
+    )
+  ) +
+  geom_point(
+    shape = ifelse(plot_ab$optimal_accuracy == TRUE, 18, 19),
+    size = ifelse(plot_ab$optimal_accuracy == TRUE, 4, 2)
+  ) +
+  ylim(0, 1) +
+  xlim(0, 1) +
+  labs(
+    x = "Type I error",
+    y = "Power"
+  ) +
+  scale_color_viridis_c() +
+  theme_bw() +
+  theme(legend.position = "bottom")
+
+p2 <-
+  ggplot(
+    plot_nn,
+    aes(
+      x = `Average N under the null`,
+      y = `Average N under the alternative`,
+      color = `Distance to optimal efficiency`,
+      Design = Design,
+      `Type I error` = `Type I error`,
+      Power = Power
+    )
+  ) +
+  geom_point(
+    shape = ifelse(plot_nn$optimal_efficiency == TRUE, 18, 19),
+    size = ifelse(plot_nn$optimal_efficiency == TRUE, 4, 2)
+  ) +
+  ylim(
+    min(plot_x$`Average N under the null`),
+    max(plot_x$`Average N under the alternative`)
+  ) +
+  xlim(
+    min(plot_x$`Average N under the null`),
+    max(plot_x$`Average N under the alternative`)
+  ) +
+  labs(
+    x = "Average N under the null",
+    y = "Average N under the alternative"
+  ) +
+  scale_color_viridis_c() +
+  theme_bw() +
+  theme(legend.position = "bottom")
+
+ggplotly(p1, width = 4, height = 4)
+# ggplotly(p2, width = 4, height = 4)
 ```
@@
 Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial.
-Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \pkg{ppseq} package.
+Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \CRANpkg{ppseq} package.
 In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial.
@@
 Bayesian sequential predictive probability monitoring provides a natural and flexible way to expand the number of patients studied in phase 1 or to design phase 2 trials that allow for efficient early stopping for futility while maintaining control of type I error and power.
-The \pkg{ppseq} package implements functionality to evaluate a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e.
+The \CRANpkg{ppseq} package implements functionality to evaluate a range of posterior and predictive thresholds for a given study design and identify the optimal design based on accuracy (i.e.
 type I error and power) or efficiency (i.e.

### 2022-01-25 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
fixing issues with html paper

@@
 bibliography: zabor-hobbs-kane.bib
-
 ---
@@
 The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error.
-The \code{calibrate\_thresholds()} function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument \code{pp\_threshold} for posterior thresholds and the argument \code{ppp\_threshold} for predictive thresholds.
-Other required arguments include the null response rate \code{p\_null}, the alternative response rate \code{p\_alt}, a vector of sample sizes at which interim analyses are to be performed \code{n} as well as the maximum total sample size \code{N}.
-Default arguments are available for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S} and the number of simulated trial datasets \code{nsim}, but customization is also possible.
-The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
-The function returns a list, the first element of which is a \code{tibble} containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative.
+The `calibrate_thresholds()` function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument `pp_threshold` for posterior thresholds and the argument `ppp_threshold` for predictive thresholds.
+Other required arguments include the null response rate `p_null`, the alternative response rate `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n` as well as the maximum total sample size `N`.
+Default arguments are available for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S` and the number of simulated trial datasets `nsim`, but customization is also possible.
+The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
+The function returns a list, the first element of which is a `tibble` containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative.
 The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power.
-The \code{print()} option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.
+The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.
 
@@
 The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power.
-The \code{optimize\_design()} function will return a list that contains the details of each of the two optimal designs.
+The `optimize_design()` function will return a list that contains the details of each of the two optimal designs.
 
@@
 
-To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the \code{calc\_decision\_rules()} function.
-The function takes the sample sizes \code{n} at which interim analyses are to be performed as well as the maximum total sample size \code{N}, the null value to compare to in the one-sample case \code{p0} (set to \code{NULL} in the two-sample case), the posterior threshold of the selected design \code{theta}, and the predictive threshold of the selected design \code{ppp}.
-Default arguments are provided for the direction of the alternative hypothesis \code{direction}, a vector of the two hyperparameters of the prior beta distribution \code{prior}, and the number of posterior samples \code{S}, though customization is also possible.
-The additional argument \code{delta} can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to \code{NULL} in the one-sample case).
-The function results in a \code{tibble}.
-In the one-sample case, the \code{tibble} includes a column for the sample size \code{n} at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision \code{ppp}.
+To ease the implementation of clinical trials designed with sequential predictive probability monitoring, once a design has been selected, a table of decision rules can be produced using the `calc_decision_rules()` function.
+The function takes the sample sizes `n` at which interim analyses are to be performed as well as the maximum total sample size `N`, the null value to compare to in the one-sample case `p0` (set to `NULL` in the two-sample case), the posterior threshold of the selected design `theta`, and the predictive threshold of the selected design `ppp`.
+Default arguments are provided for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S`, though customization is also possible.
+The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to `NULL` in the one-sample case).
+The function results in a `tibble`.
+In the one-sample case, the `tibble` includes a column for the sample size `n` at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision `ppp`.
 The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$.
 At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$.
-In the two-sample case, the \code{tibble} includes columns for \code{n0} and \code{n1}, the sample size at each interim analysis in the control and experimental arms, respectively.
-There are also columns for \code{r0} and \code{r1}, the number of responses in the control arm and experimental arm leading to the decision, respectively.
-Finally, there is a column for the posterior predictive probability associated with that decision \code{ppp}.
+In the two-sample case, the `tibble` includes columns for `n0` and `n1`, the sample size at each interim analysis in the control and experimental arms, respectively.
+There are also columns for `r0` and `r1`, the number of responses in the control arm and experimental arm leading to the decision, respectively.
+Finally, there is a column for the posterior predictive probability associated with that decision `ppp`.
 
@@
 
-Finally, to assist users in comparing the results of the various design options, a \code{plot()} option is available for the results of \code{calibrate\_thresholds} that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020].
+Finally, to assist users in comparing the results of the various design options, a `plot()` option is available for the results of `calibrate_thresholds` that allows creation of static plots using the \CRANpkg{ggplot2} package [@Wickham2016] or interactive plots using the \CRANpkg{plotly} package [@Sievert2020].
 Two plots are produced, one plotting type I error by power and indicating the optimal accuracy design, and one plotting the average sample size under the null by the average sample size under the alternative and indicating the optimal efficiency design.
@@
 Instead of simply eyeballing where points fall along the axes, users can see the specific type I error, power, average sample size under the null, average sample size under the alternative, the posterior and predictive thresholds associated with the design, as well as the distance to the upper left point on the plot.
-A \code{plot()} option is also available for the results of \code{calc\_decision\_rules}.
+A `plot()` option is also available for the results of `calc_decision_rules`.
 In the one-sample case it produces a single plot showing the sample size at each interim analysis on the x-axis and the possible number of responses at each interim analysis on the y-axis.
@@
 
-We use the \code{calibrate\_thresholds()} function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
+We use the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
 Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
-The user will be responsible for setting up a call to \code{future::plan()} that is appropriate to their operating environment and simulation setting.
+The user will be responsible for setting up a call to `future::plan()` that is appropriate to their operating environment and simulation setting.
 The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid.
-Because the code takes some time to run, the results of the below example code are available as a dataset called \code{one\_sample\_cal\_tbl} included in the \CRANpkg{ppseq} package.
+Because the code takes some time to run, the results of the below example code are available as a dataset called `one_sample_cal_tbl` included in the \CRANpkg{ppseq} package.
 
@@
 
-Next we print the results table using the \code{print()} option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7.
+Next we print the results table using the `print()` option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7.
 We find that 35 of the 76 designs meet these criteria for type I error and power.
@@
 
-We use the \code{optimize\_design()} function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
+We use the `optimize_design()` function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
 We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.
@@
 
-To compare these optimal designs with all other designs, we can use the \code{plot()} function with the \code{plotly = TRUE} option to obtain interactive visualizations, or the \code{plotly = FALSE} option to obtain static visualizations.
+To compare these optimal designs with all other designs, we can use the `plot()` function with the `plotly = TRUE` option to obtain interactive visualizations, or the `plotly = FALSE` option to obtain static visualizations.
 
@@
 Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial.
-Because of the computational time involved, the results of the below example code are available as a dataset called \code{one\_sample\_decision\_tbl} included in the \CRANpkg{ppseq} package.
+Because of the computational time involved, the results of the below example code are available as a dataset called `one_sample_decision_tbl` included in the \CRANpkg{ppseq} package.
 In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial.

### 2022-03-28 README.Rmd
Actions.
add cran installation to readme

--- 2021-04-13 README.Rmd
@@
 <!-- badges: start -->
-[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
-[![Build Status](https://travis-ci.com/zabore/ppseq.svg?branch=main)](https://travis-ci.com/zabore/ppseq) [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
+[![R-CMD-check](https://github.com/zabore/ppseq/workflows/R-CMD-check/badge.svg)](https://github.com/zabore/ppseq/actions)
 [![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
@@
 
-The goal of ppseq is to provide functions that aid in designing clinical trials with sequential posterior probability monitoring incorporated.
+The goal of ppseq is to provide functions to design clinical trials using sequential predictive probability monitoring for futility stopping.
 
@@
 
-You can install the development version of ppseq from GitHub with:
+You can install the production version of ppseq from CRAN with:
+
+```r
+install.packages("ppseq")
+```
+
+Or you can install the development version of ppseq from GitHub with:
 

### 2022-07-21 data-raw/one_sample_cal_tbl.R
vignette updates - pending server results

@@ -1,331 +1,355 @@
 ## code to prepare `one_sample_cal_tbl` dataset goes here
-
-# Note that this takes a long time to run and needs to be run on the server
-
 library(ppseq)
-library(tibble)
-library(dplyr)
-library(purrr)
-library(furrr)
-
-eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N, 
-                        direction = "greater", delta = NULL,
-                        prior = c(0.5, 0.5), S = 5000) {
-  decision <- NULL
-  ppp <- NULL
-  for (i in 1:nrow(data)) {
-    if (ncol(data) == 4) {
-      ppp[i] <- calc_predictive(
-        y = c(data$y0[i], data$y1[i]),
-        n = c(data$n0[i], data$n1[i]),
-        direction = direction, 
-        p0 = p0, 
-        delta = delta,
-        prior = prior, 
-        S = S, 
-        N = N,
-        theta = pp_threshold
-      )
-    } else if (ncol(data) == 2) {
-      ppp[i] <- calc_predictive(
-        y = data$y1[i], 
-        n = data$n1[i],
-        direction = direction, 
-        p0 = p0, 
-        delta = delta,
-        prior = prior, 
-        S = S, 
-        N = N,
-        theta = pp_threshold
-      )
-    }
-    decision[i] <- ppp[i] < ppp_threshold
-    if (decision[i] == TRUE) break
-  }
-  res0 <- add_column(
-    data[ifelse(any(decision == TRUE),
-                which(decision == TRUE),
-                length(decision)
-    ), ],
-    pp_threshold = pp_threshold,
-    ppp_threshold = ppp_threshold,
-    ppp = ppp[ifelse(any(decision == TRUE),
-                     which(decision == TRUE),
-                     length(decision)
-    )]
-  )
-  
-  if (ncol(data) == 4) {
-    res <- mutate(
-      res0,
-      positive = case_when(
-        sum(n0, n1) == sum(N) & ppp > pp_threshold ~ TRUE,
-        sum(n0, n1) == sum(N) & ppp <= pp_threshold ~ FALSE,
-        sum(n0, n1) != sum(N) ~ FALSE
-      )
-    )
-  } else if (ncol(data) == 2) {
-    res <- mutate(
-      res0,
-      positive = case_when(
-        n1 == N & ppp > pp_threshold ~ TRUE,
-        n1 == N & ppp <= pp_threshold ~ FALSE,
-        n1 != N ~ FALSE
-      )
-    )
-  }
[…]

### 2022-08-01 R/calc_decision_rules.R
updating vignettes and website

@@ -47,7 +47,7 @@
 #' posterior predictive probability. For a given value of r0, stop the trial if 
 #' the number of observed responses in the experimental arm is <=r1. At the end
 #' of the trial, the treatment is considered promising if the number of 
-#' observed responses in the experimental arm is >=r1 for a given r0.
+#' observed responses in the experimental arm is >r1 for a given r0.
 #' Any NA value in either table represents an
 #' interim look where there is no number of responses that would lead to 
 #' stopping the trial.

### 2022-08-01 R/one_sample_cal_tbl-data.R

@@ -4,7 +4,7 @@
 #' \code{calibrate_thresholds()}.
 #' See the vignette titled "One-sample expansion cohort" for a description of
 #' the input parameters used, or run
-#' \code{one_sample_cal_tbl$calibrate_thresholds_inputs}
+#' \code{one_sample_cal_tbl$inputs}
 #' to see a list of the original function inputs. For use in testing
 #' functions and in vignettes.
 #'
@@ -14,16 +14,15 @@
 #'
 #' @format A list containing a
 #' 1) a tibble 'res_summary' containing the posterior probability threshold
-#' (pp_threshold), the predictive probability threshold (ppp_threshold),
-#' the mean sample size (mean_n0 and mean_n1 for two-sample case;
-#' mean_n1 for one-sample case), and the proportion of positive trials under
-#' the null and alternative response rates.
+#' (pp_threshold); the predictive probability threshold (ppp_threshold);
+#' the mean sample size under the null (mean_n1_null) and alternative 
+#' (mean_n1_alt) response rates; the proportion of positive trials under
+#' the null (prop_pos_null) and alternative (prop_pos_alt) response rates; and 
+#' the proportion of trials stopped under the null (prop_stopped_null) and
+#' alternative (prop_stopped_alt) response rates
 #' 2) 'call_list' containing the original function call
-#' 3) 'calibrate_thresholds_inputs' a list containing the inputs to the
+#' 3) 'inputs' a list containing the inputs to the
 #' original function call
-#' 4) 'protocol_res' gives results from the protocol-specified design of the 
-#' atezolizumab case study, as detailed in the vignette titled "One-sample 
-#' expansion cohort"
 #'
 #' @keywords datasets
 "one_sample_cal_tbl"

### 2022-08-01 R/two_sample_decision_tbl-data.R

@@ -1,0 +1,20 @@
+#' Output from a two-sample call to \code{calc_decision_rules}
+#'
+#' This .rda file contains output from a two-sample call to
+#' \code{calc_decision_rules()}.
+#' See the vignette titled "Two-sample randomized trail" for a description of
+#' the input parameters used.
+#'
+#' @docType data
+#'
+#' @usage data(two_sample_decision_tbl)
+#'
+#' @format A tibble containing n, the number of patients enrolled at each
+#' futility monitoring point; r, the number of responses at which we would
+#' stop the trial at a given look if the number of observed responses is <=r,
+#' or at the end of the trial the treatment is considered promising if the
+#' number of observed responses is >r; and ppp, the predictive probability at
+#' each given look 
+#'
+#' @keywords datasets
+"two_sample_decision_tbl"

### 2022-08-01 data-raw/one_sample_cal_tbl.R

@@ -1,10 +1,11 @@
 ## code to prepare `one_sample_cal_tbl` dataset goes here
+# Note that this was actually run on the server using file H:/ppseq-papers/one_sample_cal_tbl-server.R
 library(ppseq)
 library(future)
 
 set.seed(123)
 
-future::plan(future::multicore(workers = 76))
+future::plan(future::multicore(workers = 56))
 
 one_sample_cal_tbl <- 
   calibrate_thresholds(p_null = 0.1, 
@@ -22,334 +23,4 @@ one_sample_cal_tbl <-
                        nsim = 1000
   )
 
-save(one_sample_cal_tbl, "one_sample_cal_tbl.rda")
-
-# # Note that this takes a long time to run and needs to be run on the server
-# 
-# library(ppseq)
-# library(tibble)
-# library(dplyr)
-# library(purrr)
-# library(furrr)
-# 
-# eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N, 
-#                         direction = "greater", delta = NULL,
-#                         prior = c(0.5, 0.5), S = 5000) {
-#   decision <- NULL
-#   ppp <- NULL
-#   for (i in 1:nrow(data)) {
-#     if (ncol(data) == 4) {
-#       ppp[i] <- calc_predictive(
-#         y = c(data$y0[i], data$y1[i]),
-#         n = c(data$n0[i], data$n1[i]),
-#         direction = direction, 
-#         p0 = p0, 
-#         delta = delta,
-#         prior = prior, 
-#         S = S, 
-#         N = N,
-#         theta = pp_threshold
-#       )
-#     } else if (ncol(data) == 2) {
-#       ppp[i] <- calc_predictive(
-#         y = data$y1[i], 
-#         n = data$n1[i],
-#         direction = direction, 
-#         p0 = p0, 
-#         delta = delta,
-#         prior = prior, 
-#         S = S, 
-#         N = N,
-#         theta = pp_threshold
-#       )
-#     }
-#     decision[i] <- ppp[i] < ppp_threshold
-#     if (decision[i] == TRUE) break
-#   }
-#   res0 <- add_column(
-#     data[ifelse(any(decision == TRUE),
-#                 which(decision == TRUE),
-#                 length(decision)
-#     ), ],
-#     pp_threshold = pp_threshold,
-#     ppp_threshold = ppp_threshold,
-#     ppp = ppp[ifelse(any(decision == TRUE),
[…]

### 2022-08-01 R/one_sample_decision_tbl-data.R
Adding two-sample vignette (#17)
* initial commit creating two-sample vignette

* vignette updates - pending server results

* updating website

* a bunch of website updates

* updating vignettes and website

* Increment version number to 0.1.2.9002

* Update NEWS.md

@@ -10,10 +10,11 @@
 #' @usage data(one_sample_decision_tbl)
 #'
 #' @format A tibble containing n, the number of patients enrolled at each
-#' futility monitoring point, and r, the number of responses at which we would
+#' futility monitoring point; r, the number of responses at which we would
 #' stop the trial at a given look if the number of observed responses is <=r,
 #' or at the end of the trial the treatment is considered promising if the
-#' number of observed responses is >r.
+#' number of observed responses is >r; and ppp, the predictive probability at
+#' each given look 
 #'
 #' @keywords datasets
 "one_sample_decision_tbl"

### 2022-08-01 data-raw/one_sample_decision_tbl.R

@@ -1,4 +1,6 @@
 ## code to prepare `one_sample_decision_tbl` dataset goes here
+# Note that this was actually run on the server using file H:/ppseq-papers/one_sample_decision_tbl-server.R
+library(ppseq)
 
 set.seed(123)
 

### 2022-08-01 data-raw/two_sample_cal_tbl.R

@@ -1,0 +1,26 @@
+## code to prepare `two_sample_cal_tbl` dataset goes here
+# Note that this was actually run on the server using file H:/ppseq-papers/two_sample_cal_tbl-server.R
+library(ppseq)
+library(future)
+
+set.seed(123)
+
+future::plan(future::multicore(workers = 56))
+
+two_sample_cal_tbl <- 
+  calibrate_thresholds(p_null = c(0.1, 0.1), 
+                       p_alt = c(0.1, 0.25), 
+                       n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+                       N = c(50, 50), 
+                       pp_threshold = c(0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 
+                                        0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 
+                                        0.98, 0.99),
+                       ppp_threshold = seq(0.05, 0.2, 0.05),
+                       direction = "greater", 
+                       delta = 0, 
+                       prior = c(0.5, 0.5), 
+                       S = 5000, 
+                       nsim = 1000
+  )
+
+usethis::use_data(two_sample_cal_tbl, overwrite = TRUE)

### 2022-08-01 data-raw/two_sample_decision_tbl.R

@@ -1,0 +1,20 @@
+## code to prepare `two_sample_decision_tbl` dataset goes here
+# Note that this was actually run on the server using file H:/ppseq-papers/two_sample_decision_tbl-server.R
+library(ppseq)
+
+set.seed(123)
+
+two_sample_decision_tbl <- 
+  calc_decision_rules(
+    n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+    N = c(50, 50),
+    theta = 0.92, 
+    ppp = 0.05, 
+    p0 = NULL, 
+    direction = "greater", 
+    delta = 0, 
+    prior = c(0.5, 0.5), 
+    S = 5000
+  )
+
+usethis::use_data(two_sample_decision_tbl, overwrite = TRUE)

### 2022-08-03 R/calc_next.R
unequal weights and informative messaging in `optimize_design()` (#18)
* add informative message to optimize_design

* added weighting option to optimize_design

* updating website

* Increment version number to 0.1.2.9003

* updating news

@@ -32,13 +32,15 @@
 #' # One-sample case
 #' calc_next(
 #'   y = 27, 
-#'   n = 100
+#'   n = 100,
+#'   S = 100
 #'   )
 #'
 #' # Two-sample case
 #' calc_next(
 #'   y = c(14, 23), 
-#'   n = c(100, 100)
+#'   n = c(100, 100),
+#'   S = 100
 #'   )
 #' 
 #' @importFrom stats rbeta quantile

### 2022-08-03 R/optimize_design.R

@@ -35,6 +35,14 @@ optimize_design <- function(x,
 #' @param minimum_power a numeric between 0 and 1 specifying the minimum
 #' acceptable power. Specify NULL to return the full range of resulting power.
 #' Defaults to 0 to return all results.
+#' @param w_type1 a user-specified weight on the type 1 error.
+#' Defaults to 1 for no weighting.
+#' @param w_power a user-specified weight on the power. 
+#' Defaults to 1 for no weighting.
+#' @param w_Nnull a user-specified weight on the average sample size under the 
+#' null. Defaults to 1 for no weighting.
+#' @param w_Nalt a user-specified weight on the average sample size under the
+#' alternative. Defaults to 1 for no weighting.
 #' @param ... ignored
 #' 
 #' @return A list of length two containing details of the optimal efficiency
@@ -84,7 +92,12 @@ optimize_design <- function(x,
 
 optimize_design.calibrate_thresholds <- function(x,
                                                  type1_range = c(0, 1),
-                                                 minimum_power = 0, ...) {
+                                                 minimum_power = 0, 
+                                                 w_type1 = 1, 
+                                                 w_power = 1,
+                                                 w_Nnull = 1,
+                                                 w_Nalt = 1,
+                                                 ...) {
   if (any(class(x) == "calibrate_thresholds") == FALSE)
     stop("x must be class 'calibrate_thresholds', usually an object ",
          "returned from a call to the function ppseq::calibrate_thresholds()")
@@ -106,23 +119,31 @@ optimize_design.calibrate_thresholds <- function(x,
     stop("type1_range must be a numeric vector of length 2 or NULL")
 
   options(tibble.width = Inf)
+  
+  filter_x <- 
+    filter(
+      x$res_summary,
+      prop_pos_null >= type1_range[1] &
+        prop_pos_null <= type1_range[2] &
+        prop_pos_alt >= minimum_power
+    )
+  
+  if (nrow(filter_x) == 0 )
+    stop("No results returned for the given combination of type I error and power. Please try a larger upper bound for type I error and/or a lower minimum power.")
 
   if (length(x$inputs$p_null) == 2) {
[…]

### 2022-08-03 README.Rmd
updating vignettes and website
a few README updates

@@
 
+```{r setup, include = FALSE}
+knitr::opts_chunk$set(
+  collapse = TRUE,
+  warning = FALSE,
+  comment = "#>",
+  fig.path = "man/figures/README-",
+  out.width = "100%"
+)
+```
+
 <!-- badges: start -->
@@
 
-The goal of ppseq is to provide functions to design clinical trials using sequential predictive probability monitoring for futility stopping.
+<br>
+
+## ppseq
+
+The {ppseq} package provides functions to design clinical trials using Bayesian sequential predictive probability monitoring.
+Functionality is available to design [one-arm](https://www.emilyzabor.com/ppseq/articles/one_sample_expansion.html) or [two-arm](https://www.emilyzabor.com/ppseq/articles/two_sample_randomized.html) trials by searching over a grid of combinations of posterior and predictive thresholds and identifying the optimal design according to two criteria: accuracy and efficiency.
+Interactive plotting allows easy comparison of the various design options and easy trial implementation through decision rule plots.
 
@@
 ```
+
+## Basic usage

### 2022-08-04 README.Rmd
trying to fix pkgdown badges

@@
 [![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
+[![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
+[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
 <!-- badges: end -->
 
+<br>
 <br>
@@
 ## Basic usage
+
+The primary function to search over a grid of combinations of posterior and predictive thresholds for a certain trial design is `calibrate_thresholds()`.
+This function is computationally intensive to varying degrees depending on the number of looks and the number of threshold combinations, and is best run on a server and/or with parallelization.
+
+```{r eval = FALSE}
+set.seed(12345)
+calthresh <-
+  calibrate_thresholds(
+    p_null = c(0.2, 0.2),
+    p_alt = c(0.2, 0.5),
+    n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+    N = c(50, 50),
+    pp_threshold = seq(0.9, 0.95, 0.01),
+    ppp_threshold = seq(0.05, 0.2, 0.05),
+    delta = 0
+    )
+```
+
+The resulting design options can be interactively compared by passing the results to `plot()` with the option `plotly = TRUE`:
+
+```{r eval = FALSE}
+plot(calthresh, plotly = TRUE)
+```

### 2022-08-08 README.Rmd
README updates

@@
 
+```{r eval = FALSE}
+library(ppseq)
+```
+
 The primary function to search over a grid of combinations of posterior and predictive thresholds for a certain trial design is `calibrate_thresholds()`.
@@
 set.seed(12345)
+
 calthresh <-
@@
 
-The resulting design options can be interactively compared by passing the results to `plot()` with the option `plotly = TRUE`:
+The resulting design options can be compared interactively compared by passing the results to `plot()` with the option `plotly = TRUE`.
+Static versions of the plots are also available using the default option `plotly = FALSE`, which produces `ggplot` results.
+Plot output can optionally be filtered by desired range of type 1 error and minimum power.
+The default plots all design options.
+The results can also be viewed in tabular form by passing the results to `print()` with filtering options for the desired range of type 1 error and minimum power.
 
 ```{r eval = FALSE}
-plot(calthresh, plotly = TRUE)
+plot(calthresh)
 ```
+
+```{r echo = FALSE}
+knitr::include_graphics("man/figures/README-calthresh-plot.png")
+```
+
+The optimal accuracy and optimal efficiency designs can be obtained by passing the results to the `optimize_design()` function, with filtering applied for the desired range of type 1 error and minimum power.
+
+```{r eval = FALSE}
+optimize_design(calthresh, type1_range = c(0.025, 0.05), minimum_power = 0.8)
+```
+
+```{r echo = FALSE}
+knitr::include_graphics("man/figures/README-opt-acc-table.png")
+knitr::include_graphics("man/figures/README-opt-eff-table.png")
+```
+
+After selecting a design, we can obtain a set of decision rules to implement the trial, so that no calculations will be needed during the course of the trial.
+The `calc_decision_rules()` function will generate the decision rules to stop or continue at each interim look of the trial.
+
+```{r eval = FALSE}
+set.seed(123456)
+
+opteffrules <- 
+  calc_decision_rules(
+    n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+    N = c(50, 50),
+    theta = 0.94,
+    ppp = 0.2,
+    p0 = NULL, 
+    delta = 0
+  )
+```
+
+The results can be displayed with interactive graphics by passing the results to `plot()` with the default option `plotly = TRUE`.
+Below are the static `ggplot` versions created with the option `plotly = FALSE` for demonstration purposes.
+Tabular results can be obtained by passing the results to `print()`.
+
+```{r eval = FALSE}
+plot(opteffrules, plotly = FALSE)
+```
+
+```{r echo = FALSE}
+knitr::include_graphics("D:/ppseq/man/figures/README-opteffrules-plot.png")
+```
+
+See the vignettes for the [one-sample](https://www.emilyzabor.com/ppseq/articles/one_sample_expansion.html) and [two-sample](https://www.emilyzabor.com/ppseq/articles/two_sample_randomized.html) cases for additional details about available features and options.

### 2022-08-08 README.Rmd
getting release-ready

@@
 [![R-CMD-check](https://github.com/zabore/ppseq/workflows/R-CMD-check/badge.svg)](https://github.com/zabore/ppseq/actions)
-[![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
+[![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zabore/ppseq?branch=main)
 [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)

### 2022-08-08 R/calc_decision_rules.R

@@ -55,8 +55,6 @@
 #' @examples
 #' set.seed(123)
 #' 
-#' # Setting S = 100 for speed, in practice you would want a much larger sample
-#'
 #' # One-sample case
 #' calc_decision_rules(
 #'   n = seq(5, 25, 5), 
@@ -64,7 +62,7 @@
 #'   theta = 0.86, 
 #'   ppp = 0.2, 
 #'   p0 = 0.1, 
-#'   S = 100
+#'   S = 50
 #'   )
 #' 
 #' # Two-sample case
@@ -76,7 +74,7 @@
 #'   p0 = NULL, 
 #'   direction = "greater", 
 #'   delta = 0, 
-#'   S = 100
+#'   S = 50
 #'   )
 #'
 #' @importFrom tibble tibble add_row

### 2022-08-08 R/calibrate_thresholds.R

@@ -228,9 +228,6 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #'
 #' @examples
 #' 
-#' # Setting S = 50 and nsim = 50 for speed
-#' # In practice you would want a much larger sample and more simulations
-#' 
 #' # One-sample case
 #' set.seed(123)
 #' 
@@ -241,8 +238,8 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #'   N = 15,
 #'   pp_threshold = c(0.85, 0.9),
 #'   ppp_threshold = c(0.1, 0.2),
-#'   S = 50, 
-#'   nsim = 50
+#'   S = 10, 
+#'   nsim = 10
 #'   )
 #' 
 #' # Two-sample case
@@ -256,8 +253,8 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #'   pp_threshold = c(0.8, 0.85),
 #'   ppp_threshold = c(0.2, 0.3),
 #'   delta = 0,
-#'   S = 50, 
-#'   nsim = 50
+#'   S = 10, 
+#'   nsim = 10
 #'   )
 #' 
 #' @importFrom purrr pmap_dfr map cross_df

### 2022-08-08 R/optimize_design.R

@@ -63,8 +63,8 @@ optimize_design <- function(x,
 #'   N = 15,
 #'   pp_threshold = c(0.85, 0.9),
 #'   ppp_threshold = c(0.1, 0.2),
-#'   S = 50, 
-#'   nsim = 50
+#'   S = 10, 
+#'   nsim = 10
 #'   )
 #' 
 #' optimize_design(cal_tbl1)
@@ -81,8 +81,8 @@ optimize_design <- function(x,
 #'   pp_threshold = c(0.8, 0.85),
 #'   ppp_threshold = c(0.2, 0.3),
 #'   delta = 0,
-#'   S = 50, 
-#'   nsim = 50
+#'   S = 10, 
+#'   nsim = 10
 #'   )
 #' 
 #' optimize_design(cal_tbl2)

### 2022-08-08 R/plot.R

@@ -37,8 +37,8 @@
 #'   N = 15,
 #'   pp_threshold = c(0.85, 0.9),
 #'   ppp_threshold = c(0.1, 0.2),
-#'   S = 50,
-#'   nsim = 50
+#'   S = 10,
+#'   nsim = 10
 #'   )
 #' 
 #' plot(cal_tbl1, type1_range = c(0.01, 0.2), minimum_power = 0.7)
@@ -252,8 +252,6 @@ plot.calibrate_thresholds <- function(x,
 #' 
 #' set.seed(123)
 #' 
-#' # Setting S = 100 for speed, in practice you would want a much larger sample
-#' 
 #' # Two-sample case
 #' dec_tbl <- calc_decision_rules(
 #'   n = cbind(seq(5, 15, 5), seq(5, 15, 5)),
@@ -263,10 +261,10 @@ plot.calibrate_thresholds <- function(x,
 #'   p0 = NULL,
 #'   direction = "greater",
 #'   delta = 0,
-#'   S = 100
+#'   S = 50
 #' )
 #' 
-#' plot(dec_tbl)
+#' plot(dec_tbl, plotly = FALSE)
 #' 
 #' @importFrom dplyr mutate filter select rename ungroup group_by 
 #' full_join case_when

### 2022-08-08 R/print.R

@@ -20,9 +20,6 @@
 #'
 #' @examples
 #' 
-#' # Setting S = 50 and nsim = 50 for speed
-#' # In practice you would want a much larger sample and more simulations
-#' 
 #' set.seed(123)
 #' 
 #' cal_tbl1 <- calibrate_thresholds(
@@ -32,8 +29,8 @@
 #'   N = 15,
 #'   pp_threshold = c(0.85, 0.9),
 #'   ppp_threshold = c(0.1, 0.2),
-#'   S = 50,
-#'   nsim = 50
+#'   S = 10,
+#'   nsim = 10
 #'   )
 #' 
 #' print(cal_tbl1)

### 2022-09-28 inst/ppseq-package-paper/zabor-hobbs-kane-1-review-response.Rmd
drafting revision 1 response

---
title: "ppseq: An R Package for Sequential Predictive Probability Monitoring"
subtitle: "Review 1 Responses"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> **Reviewer 1**

> Overview:
> The authors present an R package to design early phase clinical trials with binary outcomes leveraging a Bayesian framework. I find the package potentially applicable to many settings. However, the paper includes some statements that are not sufficiently clear, or debatable.  Some details should be clarified. I also think that a running toy-example that can be implemented without the use of a server would be a good tutorial to familiarize with the package before large scale implementation. Some comments to follow:

Thank you for this thorough review. We agree that there were parts of the paper that were unclear or lacked sufficient detail, and have strived to address all of your comments in an effort to improve the clarify and accuracy throughout.

> Major comments:

> - The package and paper only consider binary outcomes. This should be clearly stated in the abstract/introduction.

Thank you for pointing this out. We agree, and have added this detail to both the abstract and the last paragraph of the introduction.

> - I am not sure what is the point of second paragraph of the introduction. In particular, the authors claim that on-the fly decisions can lead to “large sample size” and “poor statistical properties”. Why? I don’t think that this is necessarily the case. I suggest improving the clarity of this paragraph.

The goal of this paragraph was to introduce readers to the often large sample sizes that have been used in recent expansion cohorts, which is part of the problem addressed by this package. We see how the wording did not emphasize this point properly, though, and have reworded in response to this comment. Thank you.

> - In the third paragraph of the introduction the author claims that Bayesian models are flexible “while also maintaining the control of type-I error”. I disagree with this statement. Bayesian models do not naturally account for type-I error and power (frequentist power?) constraints , which is the whole point of the calibration exercise that motivate this package.

We agree that the way this sentence was worded was not accurate. We have reworded this section to improve clarity and accuracy.

> - In the introduction and later in the paper, the authors introduce “one-sample” and “two-sample” phase-I design. The meaning should be explained somewhere.

We removed this language from the introduction, and added a definition of one-sample and two-sample to the beginning of the "Predictive probability monitoring" section. In this context, one-sample refers to a trial in which all patients are enrolled onto a single arm and given the experimental treatment and two-sample refers to a trial in which patients are enrolled onto two different treatment arms, or a treatment and control arm, for comparative purposes.

> - Page 3: “The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power.” I don’t think this is correct. Specifically, I am not sure what “represent” means in this context.  The proportion of “positive trial” (maybe better defined as the proportion of simulated trials for which we reject the null hypothesis of inferiority, or something along this line) is a Monte Carlo estimate of the type-I error. Correct? Same goes for power. The number of trials replications should also be discussed somewhere in the paper. I think this is very important aspect   when selecting an optimal strategy, i.e., we don’t want the monte Carlo error to dominate the difference between alternative strategies. This aspect should also be discussed (see also following point).

Thank you for pointing out the lack of specificity in our wording. We meant to convey that the proportion of trials simulated under the null hypothesis for which we reject the null hypothesis is an estimate of the type I error. And similarly the proportion of trials simulated under the alternative hypothesis for which we reject the null hypothesis is an estimate of the power. We have revised the sentence accordingly.

And you are correct that the number of trial replications is an important aspect of this. Our default arguments are 1000 datasets with 5000 posterior draws in each, which we believe are reasonable values. We have added a note of caution related to these values in the Package Overview section.

> - I find the section “Optimization” that describe the function “optimal_design()” to be misleading. First, I think that the authors should explicitly state that this function does not perform an optimization, but rather select a trial among the ones that have been simulated. There is no guarantee that the selected design is in fact optimal with respect to any criteria (e.g., if we select a coarse grid of thresholds). Second, I think that with a minimal effort this function can be generalized to take as input any user specified utility that take as impute type-I error and power, and/or average sample sizes. I think this could be helpful for a broader use of the package.

This is a very good point, and we apologize for the poor word choice here. We have renamed this section "Design selection" and have removed the word optimization to avoid any confusion. Since the paper was originally drafted, we have added the option of weighting the various criteria, to allow more user-specified flexibility in terms of the value of type I error versus power, and average sample size under the null versus average sample size under the alternative. We have added a description of this new functionality to the Design selection section.

> - In the case study: what is the rational to select the considered posterior thresholds and not, say, an equi-spaced grid. I think this is an important point to discuss, and can have a large impact on the result.

Thank you, we have written the case study a bit to be a two-sample example in order to differentiate from the added basic example, and have also added some information justifying the selected thresholds.

> - In the code snipped at page 4 the parameters `S` and `nsim` of the `calibrate_thresholds()` function have not been discussed in the text. I think this are important parameters.

We have added more discussion of the importance of selecting sufficiently large values of `S` and `nsim` throughout the paper, thank you.

> - Figure 1: I think the x-axis should be rescaled in a (say 0-0.25 for the left panel and 35-55 for the right panel). The plot would be easier to read.

Thank you for this suggestion, we have rescaled the axes.

> - Page 7: In the first line of the tibble `one_sample_decision_tbl` there is NA,NA. If I am not mistaken this is interpreted as “never stop”. I will explicitly indicate this.
> - There are some typos in the bibliography. For example, “r” in place of “R”, “phase i” in place of “phase I”. Please double check.
> - I think that the paper would benefit of an additional toy example that can be run on a personal computer rather than on a server. This would allow to familiarize with the functions before running large scale simulations.

We agree, and have added a toy example that can easily be run interactively prior to presenting the more realistic case study.

> Minor comments:

> - The third sentence in the abstract “Recent successes […]” is very long  and convoluted. I suggest rephrasing.

We have taken this suggestion and revised the abstract to improve clarity.

> - Page 2: Define the acronym RECIST.

In parentheses we have added the definition of the acronym RECIST: Response Evaluation Criteria in Solid Tumors.

> - Page 2: Bayes rule -> Bayes’s theorem.

Thank you, we have made this correction.

> - Page2: when naming distribution (e.g., Beta-Binomial) I suggest to use $\mbox{Beta-Binomial}()$ or similar, to avoid wrong spacing/font.

Thank you for this suggestion, we have used it for all distributions to achieve the proper font/spacing.

> - In the section decision rules, what is a decision point “r”? Please define this.

We have reworded this for clarity, thank you.

> -Figure 1: in the legend (#fig:unnamed-chunk-7). I suppose this is a compilation error.

This is indeed a compilation error. I am not sure why this is being generated or how to suppress it, so hopefully it can be fixed when the article is finalized for publication,

> **Reviewer 5**

> The article describes the ppseq package for R, which provides several functions to facilitate clinical trial design using sequential predictive probability monitoring. A user looking to design an early phase trial can use Bayesian methods to ask what the probability of early stopping is and what expected sample sizes would be under different assumptions about outcomes in treatment and control groups and with different stopping rules, while maintaining desired type I error rates. The package is small, providing only three user-facing functions reported in the manuscript, along with plot and print methods. However, the functions are user friendly with obvious options, and the article provides a clear use case and justification for the methods. However, the article itself is quite difficult to follow, which is not helped by the example not being feasible to run for a user with a standard desktop or laptop computer (the example is reportedly run here on a 192 core computing cluster!). I have provided some queries, recommendations below.

We are sorry that the article in it's original form was difficult to follow. We hope the changes have made it more understandable, and that all of the needed detail is now included. Thank you for taking the time to conduct such a thorough review of our paper, we do believe that it has been improved as a result.

> 1. I found the Package Overview section difficult to follow, since it provides only a narrative description of the inputs of the functions and options. It would be useful to link the description to the statistical discussion in the previous section, explaining which option relates to which parameter, and what each output reports in terms of the statistical notation. It may help to integrate the case study into the package description so that the reader can see what the text means in terms of the example. However, I would also strongly recommend providing an example(s) that a user can implement themselves in a reasonable amount of time, as this often helps when using a package for the first time. Indeed, is the package useful if it requires such significant computational resources (see also point 3)?

Thank you for this comment, we agree that much detail was overlooked in the Package Overview section, and we have expanded this section to include much more detail of the underlying algorithm, and to link the notation to the preceding section. We believe this is much easier to follow now. We have also added a section for a toy example that can easily be run interactively prior to presenting the more detailed case study.

We do think this package is useful despite the potentially long computational time. This is meant to be used at the design stage of a clinical trial, when there is ample time to allow for computation. Once the design has been selected and decided, no additional computations are required throughout the conduct of the trial, as the decision table resulting from `calc_decision_rules()` is available for decisions at each interim look, making trial conduct fast at seamless at each interim analysis and at the end of the trial.

> 2. I was a little confused by the notation, some more detail would make it clearer I believe. Is the setting a randomized trial with treatment and control groups, or a dose expansion cohort? We are told that $p$ is the probability of response – is this in the dose expansion cohort? And we have $p_0$ and $p_1$ as the probabilities of response in the control and treatment groups, respectively. But are these random variables we intend to estimate, or are they pre-defined thresholds of interest (e.g. “high” and “low” efficacy)?

Thank you for pointing out the lack of clarity here. We have revised the wording to clarify that $p$ represents the true response probability, $p_0$ is a fixed threshold of unacceptable response rate, $p_1$ is a fixed threshold of acceptable response rate. We have also clarified that our example focuses on the setting of a single arm dose-expansion cohort.

> 3. It was not clear to me what each function was doing computationally. In particular, the method implemented by the `calibrate_thresholds()` function was not explicated. I wanted to understand why the function took so long to run, and why it might be drawing posterior samples. My assumption was that the algorithm was something roughly like the following. For nsim iterations:
i. Simulate a realization of $p^r$ from the prior $\Beta(a_0, b_0)$
ii. Simulate a realization of $x^r | n, p^r ~ \Binomial(n, p^r)$
iii. Calculate the probability of success in that trial $$PP^r=\Pr(p^r>p_0|x^r, N, n)=\int_{p_0}^1 f_p(q|x^r, N, n)\,dq=\sum_{x^*}^{N-n}\int_{p_0}^1\Pr(X^*=x^*|q,N-n)f_p(q|x^r,N,X^*=x^*)\,dq

> Then the PPP is the the proportion of $PP^r>\theta^*$. If this is the case, then the integral term has a closed form and posterior simulations do not seem necessary, in which case the function should run much quicker than it currently does. Running time does seem like a key issue for the functions in this package. I may of course be completely wrong about what the function is doing, but it would help to have some clarification so we know the function is doing what the user expects.

Thank you for drawing our attention to this oversight. We have added much more detail about the underlying computational algorithm for the `calibrate_thresholds()` function in the Package Overview section, and hope that this clarifies the functionality.

>4. The section “Optimisation” mentions minimizing Euclidean distance on different plots, however it isn’t clear what this means without notation or until the example is provided.

Thank you for pointing out the lack of clarity in this section. We have added much more detail about the computations being done, with notation, to this section, and hope it is now more clear. Additionally, in response to another reviewer's comment we have renamed this section "Design selection" and removed reference to optimization, since no true mathematical optimization was being conducted.

### 2022-09-28 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
drafting revision 1 response

@@
 Yet the promise of precision medicine comes with the cost of increased complexity.
-Recent successes for non-cytotoxic treatments have relied on expansive early phase clinical trials that far exceeded the size of trials typically implemented for chemotherapies, with the addition of baskets for different disease sub-types or multiple dosing groups, expansion cohorts to further study safety and obtain preliminary efficacy information, or randomization to further define efficacy or compare across doses, among other design elements.
-With the enlarged sample sizes come increasing ethical concerns regarding patients who enroll in clinical trials, and the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
-The R package \pkg{ppseq} provides a framework for designing early phase clinical trials using sequential futility monitoring based on the Bayesian predictive probability.
-Trial designs can be compared using interactive plots and optimized based on efficiency or accuracy.
+Recent trials of targeted treatments have included expansion cohorts with sample sizes far exceeding those in traditional early phase trials of chemotherapeutic agents.
+The enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
+The R package \pkg{ppseq} provides a framework for designing early phase clinical trials of binary endpoints using sequential futility monitoring based on Bayesian predictive probability.
+Trial designs can be compared using interactive plots and selected based on measures of efficiency or accuracy.
 draft: true
@@
 
-In the context of cytotoxic treatments, phase 1 trials in oncology traditionally have the primary aim of identifying the maximum tolerated dose (MTD), defined as the highest dose that still maintains a certain pre-specified rate of toxicity, most commonly 30\%, in a dose-escalation phase.
-Designs for dose-escalation trials include the rule-based 3+3 design and the model-based continual reassessment method, among others.
-But with increasing study focused on non-cytotoxic treatments, such as immunotherapies, the MTD either may not exist or may not be relevant, and toxicities may either develop much later or even be chronic, making these treatments difficult to study with traditional dose-escalation designs [@Pestana2020].
-As a result, it is becoming increasingly common to include dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete.
-In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes that all share a common biomarker that the treatment is targeting, or to further characterize the pharmacokinetics and/or pharmacodynamics.
+Statistical methods for early phase oncology trials were developed in the context of cytotoxic treatments.
+Most cytotoxic treatments exhibit increases in both efficacy and toxicity with increasing dose.
+As a result, phase I trials were centered on identifying the maximum tolerated dose (MTD), defined as the highest dose that did not exceed a pre-specified toxicity threshold.
+Phase I dose-escalation trials were designed using the rule-based 3+3 method or the model-based continual reassessment method, among others.
+But advances in drug discovery have produced numerous biomarker-guided non-cytotoxic therapies such as small molecule inhibitors, antibody drug conjugates, immune checkpoint inhibitors, and monoclonal antibodies.
+These therapies typically do not exhibit the same pattern of increases in both efficacy and toxicity with increasing dose, so the MTD as traditionally defined may not exist.
+Instead, lower doses may have equivalent efficacy but lower toxicity as compared to higher doses.
+As a result, these therapies can be difficult to study with traditional dose-escalation designs [@Pestana2020].
+To address this issue, recent phase I trials have included large dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete.
+In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes, or to further characterize the pharmacokinetics and/or pharmacodynamics.
 The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].
 
-But dose-expansion cohorts are not always planned in advance, and therefore may be subject to on-the-fly decision making that can lead to large sample sizes and poor statistical properties.
+But despite these successes, recent dose-expansion cohorts have not always been planned in advance, leading to uncertain statistical properties, and have at times included samples sizes that far exceed those typically seen in early phase trials of cytotoxic treatments.
 For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015].
@@
 The expansion cohort in metastatic urothelial carcinoma was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
+These enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
 
-Bayesian sequential predictive probability monitoring provides a natural framework for early phase oncology trial design, allowing flexibility to stop early for futility or safety concerns while also maintaining traditional levels of power and control of type I error.
-As a result, predictive probability has been proposed as an approach to interim monitoring in clinical trials previously [@Dmitrienko2006; @Lee2008; @Saville2014; @Hobbs2018].
-However, in order to be useful to investigators designing such trials, software must be made available to select an optimal design.
-This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design one-sample and two-sample early phase clinical trials using sequential predictive probability monitoring for futility.
+Bayesian predictive probability has been proposed as an approach for sequential monitoring in early phase oncology trials [@Dmitrienko2006; @Lee2008; @Hobbs2018; @Saville2014].
+However, in order to be useful to investigators designing such trials, software must be made available to calibrate the design for the desired statistical properties.
+To our knowledge no such software currently exists in the R programming language.
+This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility.
 Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making.
-Moreover, we demonstrate optimization criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package.
+Moreover, we demonstrate criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package.
 While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.
@@
 
-Consider the setting of a binary outcome, such as tumor response as measured by the RECIST criteria in the setting of a study in patients with solid tumors, where each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$.
+Consider the setting of an expansion cohort with a binary outcome, such as tumor response as measured by the RECIST (Response Evaluation Criteria in Solid Tumors) criteria.
+Here we will focus on the one-sample setting, in which all patients in the trial are enrolled onto a single arm and given the experimental treatment of interest.
+Functionality is also available for the two-sample setting, in which patients are enrolled onto two different treatment arms, or a treatment and a control arm, for comparative purposes.
+Each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$.
 Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients.
-Let $p$ represent the probability of response, where $p_0$ represents the null response rate under no treatment or the standard of care treatment and $p_1$ represents the alternative response rate under the experimental treatment.
+Let $p$ represent the true probability of response.
+Fix $p_0$ as the threshold for unacceptable response rate and $p_1$ as the threshold for acceptable response rate.
 Most dose-expansion studies with an efficacy aim will wish to test the null hypothesis $H_0: p \leq p_0$ versus the alternative hypothesis $H_1: p \geq p_1$.
 
-The Bayesian paradigm of statistics is founded on Bayes rule, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution.
-Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $Beta(a_0, b_0)$ and our data $X$ have a binomial distribution $bin(n, p)$.
-Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim Beta(a_0 + x, b_0 + n - x)$.
+The Bayesian paradigm of statistics is founded on Bayes' theorem, which is a mathematical theory that specifies how to combine the prior distributions that define prior beliefs about parameters, such as the true response rate $p$, with the observed data, such as the total number of responses $X$, yielding a posterior distribution.
+Here the prior distribution of the response rate $\pi(p)$ has a beta distribution $\mbox{Beta}(a_0, b_0)$ and our data $X$ have a binomial distribution $\mbox{Bin}(n, p)$.
+Combining the likelihood function for the observed data $L_x(p) \propto p^x (1-p)^{n-x}$ with the prior, we obtain the posterior distribution of the response rate, which follows the beta distribution $p|x \sim \mbox{Beta}(a_0 + x, b_0 + n - x)$.
 A posterior probability threshold $\theta$ would be pre-specified during the trial design stage.
@@
 
-The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $Beta-binomial(n^*, a_0 + x, b_0 + n - x)$.
+The posterior predictive distribution of the number of future responses $X^*$ in the remaining $n^*=N-n$ future patients follows a beta-binomial distribution $\mbox{Beta-binomial}(n^*, a_0 + x, b_0 + n - x)$.
 Then the posterior predictive probability (PPP), is calculated as $PPP = \sum_{{x^*}=0}^{n^*} \Pr(X^*=x^*|x) \times I(\Pr(p>p_0 | X, X^*=x^*) > \theta)$.
@@
 The goal is to establish a set of decision rules at the trial planning phase that would be used for interim monitoring during the course of the trial.
-The main challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to maintain the desired level of power and type I error.
-The `calibrate_thresholds()` function will evaluate a grid of posterior and predictive thresholds provided by the user as vector inputs through the argument `pp_threshold` for posterior thresholds and the argument `ppp_threshold` for predictive thresholds.
-Other required arguments include the null response rate `p_null`, the alternative response rate `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n` as well as the maximum total sample size `N`.
-Default arguments are available for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S` and the number of simulated trial datasets `nsim`, but customization is also possible.
-The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design.
-The function returns a list, the first element of which is a `tibble` containing the posterior threshold, predictive threshold, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative.
-The proportion of positive trials under the null represents the type I error and the proportion of positive trials under the alternative represents the power.
+The main computational challenge in designing such a trial is joint calibration of the posterior probability and posterior predictive probability thresholds to be used in the trial in order to achieve the desired levels of frequentist type I error and power.
+The main function for achieving this aim is the `calibrate_thresholds()` function, which will evaluate a grid of posterior thresholds $\theta$ and predictive thresholds $\theta^*$ provided by the user as vector inputs specified with the arguments `pp_threshold` and `ppp_threshold`, respectively.
+Other required arguments include the unacceptable response rate $p_0$ specified by `p_null`, the acceptable response rate $p_1$ specified by `p_alt`, a vector of sample sizes at which interim analyses are to be performed `n`, and the maximum total sample size `N`.
+The direction of the alternative hypothesis is specified with the argument `direction` and defaults to `"greater"`, which corresponds to the alternative hypothesis $H_1: p \geq p_1$.
+The hyperparameters of the prior beta distribution are specified with the argument `prior` and default to `c(0.5, 0.5)`, which denotes a $\mbox{Beta}(0.5, 0.5)$ distribution.
+The number of posterior samples are specified with the argument `S`, which defaults to `5000` samples, and the number of simulated trial datasets is specified with the argument `nsim`, which defaults to `1000`.
+The additional argument `delta`, which defaults to `NULL` for the one-sample setting, can specify a clinically meaningful difference between groups $\delta = p_1 - p_0$ in the case of a two-sample trial design.
+
+The `calibrate_thresholds()` function conducts the following algorithm, given the default arguments:
+
+1.
+Generate `nsim` datasets, denoted by $j$, containing the cumulative number of responses $x$ at each interim sample size $n$ from a binomial distribution under the unacceptable (i.e.
+null) response rate $p_0$, specified by `p_null`, and under the acceptable (i.e.
+alternative) response rate $p_1$, specified by `p_alt`, for each interim look, denoted by $l$.
+2.
+For dataset $j$ and posterior threshold $\theta_k$, draw `S` samples, denoted by $s$, from the posterior distribution $p|x_{ls} \sim \mbox{Beta}(a_0 + x_l, b_0 + n_l - x_l)$ where $x_l$ is the number of responses at interim look $l$ and $n_l$ is the number of enrolled patients at interim look $l$.
+Use each $p|x_{ls}$ as the response probability in a binomial distribution to generate the number of future responses $X^*_{ls}$ in the remaining $n^*_l=N-n_l$ future patients at interim look $l$.
+
+    a.
+Then for each $X^*_{ls}$, generate `S` posterior probabilities, denoted by $s'$, at the end of the trial: $PP^*_{lss'} \sim \mbox{Beta}(a_0 + X^*_{ls} + x_l, b_0 + n^*_l - (X^*_{ls} + x_l))$ and calculate $PP^*_{ls} = \Pr(p>p_0 | X^*_{ls}) = \frac{\sum_1^{S'} PP^*_{lss'} > p_0}{S'}$.
+    b. Estimate the predictive probability at posterior threshold $k$ as $PPP_{lk} = \frac{\sum_1^S PP^*_{ls}>\theta_k}{S}$.
+    c. Stop the trial for dataset $j$ at interim look $l$ and predictive threshold $m$ if $PPP_{lk} < \theta^*_m$. Otherwise continue enrolling.
+3. Repeat (2) over all combinations of datasets $j$, posterior thresholds $k$, and predictive thresholds $m$.
+4.
+If dataset $j$ was stopped early for futility then we do not reject the null hypothesis.
+If dataset $j$ reached full enrollment, we reject the null hypothesis $H_0: p \leq p_0$ at posterior threshold $k$ if $PPP_{lk} > \theta_k$.
+
+The function returns a list, the first element of which is a `tibble` containing the posterior threshold $\theta$, the predictive threshold $\theta^*$, the mean sample size under the null and the alternative, the proportion of positive trials under the null and alternative, and the proportion of trials stopped early under the null and alternative.
+The proportion of trials simulated under the null hypothesis for which the null hypothesis was rejected is an estimate of the type I error, and the proportion of trials simulated under the alternative hypothesis for which the null hypothesis was rejected is an estimate of the power.
 The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.
+Note that the results will be sensitive to the choice of `nsim` and `S`.
+We have set what we believe are reasonable defaults and would caution users against reducing these values without careful consideration.
 
-## Optimization
+## Design selection
 
 After obtaining results for all combinations of evaluated posterior and predictive thresholds, the next step is to select the ideal design from among the various options.
-The \CRANpkg{ppseq} package introduces two optimization criteria to assist users in making a selection.
-The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the type I error by the power.
-The second, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to the top left point on a plot of the average sample size under the null by the average sample size under the alternative, subject to constraints on the type I error and power.
-The `optimize_design()` function will return a list that contains the details of each of the two optimal designs.
+The \CRANpkg{ppseq} package introduces two criteria to assist users in making a selection.
+The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to 0 type I error probability and a power of 1.
+To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_(1-\beta) * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power.
+The design with the smallest value of $AED_{km}$ is selected as optimal.
+The second criteria, called the "optimal efficiency" design, identifies the design that minimizes the Euclidean distance to minimal average sample size under the null and maximal average sample size under the alternative.
+To accomplish this, the efficiency Euclidean distant (EED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $EED_{km} = w_{\bar{N}_{H_0}} * (\bar{N}_{H_0km} - min(\bar{N}_{H_0}))^2 + w_{\bar{N}_{H_1}} * (\bar{N}_{H_1km}-max(\bar{N}_{H_1}))^2$, where $w_{\bar{N}_{H_0}}$ and $w_{\bar{N}_{H_1}}$ are optional weights on the average sample size under the null and alternative, respectively, $\bar{N}_{H_0km}$ and $\bar{N}_{H_1km}$ denote the average sample sizes under the null and alternative, respectively, and $min(\bar{N}_{H_0})$ and $max(\bar{N}_{H_1})$ denote the minimum average sample size under the null and the maximum average sample size under the alternative alternative, respectively, across all combinations of $k$ and $m$.
+The design with the smallest value of $EED_{km}$ is selected as optimal.
+The `optimize_design()` function returns a list that contains the details of each of the two optimal designs.
 
@@
 The function takes the sample sizes `n` at which interim analyses are to be performed as well as the maximum total sample size `N`, the null value to compare to in the one-sample case `p0` (set to `NULL` in the two-sample case), the posterior threshold of the selected design `theta`, and the predictive threshold of the selected design `ppp`.
-Default arguments are provided for the direction of the alternative hypothesis `direction`, a vector of the two hyperparameters of the prior beta distribution `prior`, and the number of posterior samples `S`, though customization is also possible.
-The additional argument `delta` can specify the clinically meaningful difference between groups in the case of a two-sample trial design (set to `NULL` in the one-sample case).
+Arguments `direction`, `prior`, `S`,  `delta` are as described in the Package Overview section, with the same defaults.
 The function results in a `tibble`.
-In the one-sample case, the `tibble` includes a column for the sample size `n` at each interim analysis, a column for the decision point $r$, and a column for the posterior predictive probability associated with that decision `ppp`.
-The trial would stop at a given look if the number of observed responses is less than or equal to $r$ otherwise the trial would continue enrolling if the number of observed responses is greater than $r$.
+The trial would stop at a given look if the number of observed responses is less than or equal to $r$, otherwise the trial would continue enrolling if the number of observed responses is greater than $r$.
 At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$.
+In the one-sample case, the resulting `tibble` includes a column for the sample size `n` at each interim look, $r$ at each look, and a column for the associated posterior predictive probability `ppp`.
 In the two-sample case, the `tibble` includes columns for `n0` and `n1`, the sample size at each interim analysis in the control and experimental arms, respectively.
-There are also columns for `r0` and `r1`, the number of responses in the control arm and experimental arm leading to the decision, respectively.
+There are also columns for `r0` and `r1`, the number of responses in the control arm and experimental arm, respectively, leading to the decision to stop or continue.
 Finally, there is a column for the posterior predictive probability associated with that decision `ppp`.
@@
 
-# Case study
+# Package demonstration
 
-To demonstrate the functionality of the \CRANpkg{ppseq} package, we focus on a re-design of a dose-expansion cohort for the study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring.
-Atezolizumab is an anti-PD-L1 treatment that was originally tested in the phase 1 setting in a basket trial across a variety of cancer sites harboring PD-L1 mutations.
-The atezolizumab expansion study in mUC had the primary aim of further evaluating safety, pharmacodynamics and pharmacokinetics and therefore was not designed to meet any specific criteria for type I error or power.
-An expansion cohort in mUC was not part of the original protocol design, but was rather added later through a protocol amendment.
-The expansion cohort in mUC ultimately evaluated a total of 95 participants [@Powles2014].
-Other expansion cohorts that were included in the original protocol, including in renal-cell carcinoma, non-small-cell lung cancer, and melanoma, were planned to have a sample size of 40.
-These pre-planned expansion cohorts were designed with a single interim analysis for futility that would stop the trial if 0 responses were seen in the first 14 patients enrolled.
-According to the trial protocol, this futility rule is associated with at most a 4.4\% chance of observing no responses in 14 patients if the true response rate is 20\% or higher.
-The protocol also states the widths of the 90\% confidence intervals for a sample size of 40 if the observed response rate is 30\%.
-There was no stated decision rule for efficacy since efficacy was not an explicit aim of the expansion cohorts.
-In the re-design we assume a null, or unacceptable, response rate of 0.1 and an alternative, or acceptable, response rate of 0.2.
-We plan a study with up to a total of 95 participants.
-In our sequential predictive probability design we will check for futility after every 5 patients are enrolled.
-We consider posterior thresholds of 0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, and 1, and predictive thresholds of 0.05, 0.1, 0.15, and 0.2.
+In this section I will present a basic example to demonstrate the functionality included in the \CRANpkg{ppseq}. The following section will present a more realistic case study.
 
@@
 
-We use the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
-Because of the inherent computation intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
+Consider the case where we are interested in designing a trial to investigate how a new treatment impacts tumor response measured as a binary outcome of response versus no response.
+We know the current standard of care treatment results in a tumor response rate of 10\%, and we wish to improve this by 30\%.
+So we wish to test $H_0: p \leq 0.1$ versus $H_1: p \geq 0.4$, so we set `p_null = 0.1` and `p_alt = 0.4`.
+This is a rare disease so our maximum sample size is 15, so we set `N = 15`, and we will do interim analyses after every 5 patients, so we set `n = seq(5, 15, 5)`.
+We wish to examine designs based on combinations of posterior thresholds $\theta = {0.85, 0.90}$ and predictive thresholds $\theta^*={0.1, 0.2}$, so we set `pp_threshold = c(0.85, 0.9)` and `ppp_threshold = c(0.1, 0.2)`.
+Finally, for computational speed in this basic example, we set `S=50` and `nsim=50`, but in practice we would want to use much larger values, in the thousands.
+
+```{r message = FALSE}
+set.seed(123)
+
+cal_thresh <-
+  calibrate_thresholds(
+    p_null = 0.1, 
+    p_alt = 0.4,
+    n = seq(5, 15, 5), 
+    N = 15,
+    pp_threshold = c(0.85, 0.9),
+    ppp_threshold = c(0.1, 0.2),
+    S = 50, 
+    nsim = 50
+    )
+```
+
+Since there are only four design options in this toy example, we print the entire results table using a call to `print()`:
+
+```{r}
+print(cal_thresh)
+```
+
+We use `optimize_design()` to identify the optimal accuracy and optimal efficiency designs, subject to type I error between 0.025 and 0.1, specified by `type1_range = c(0.025, 0.1)`, and power of at least 0.75, specified by `minimum_power = 0.75`:
+
+```{r}
+optimize_design(cal_thresh, type1_range = c(0.025, 0.1), minimum_power = 0.75)
+```
+
+And to ease interim analysis during the course of the trial, we obtain the decision table using the `calc_decision_rules` function, with the thresholds of our selected optimal efficiency design passed as `theta = 0.9` and `ppp_threshold = 0.2`.
+For computational speed in this basic example, we set `S=50`, but in practice we would want to draw many more posterior samples, and values of 5000 or more should be considered.
+
+```{r}
+calc_decision_rules(
+  n = seq(5, 15, 5), 
+  N = 15,
+  theta = 0.9,
+  ppp = 0.2,
+  p0 = 0.1,
+  S = 50
+)
+```
+
+This table tells us that at the first interim analysis after we have observed the first five patients for response, we would stop the trial if no responses had been observed, and would continue enrolling if at least one response had been observed.
+At the second interim analysis after we have observed the first ten patients for response, we would stop the trial if zero or one responses had been observed, and would continue enrolling if at least two responses had been observed.
+At the end of the trial after all 15 patients have been observed for response, we would declare the new treatment promising if at least three responses had been observed.
+
+# Case study
+
+The case study focuses on a re-design of a phase II study of atezolizumab in metastatic urothelial carcinoma patients (mUC) using sequential predictive probability monitoring.
+Atezolizumab is a programmed death-ligand 1 (PD-L1) blocking monoclonal antibody that was given accelerated approval by the U.S.
+Food and Drug Administration in May 2016 for the treatment of patients with locally advanced or metastatic urothelial carcinoma who had disease progression following platinum-containing chemotherapy.
+The approval was based on the results of a single-arm phase II study in 310 patients [@Rosenberg2016].
+The phase II study used a hierarchical fixed-sequence testing procedure to test increasingly broad subgroups of patients based on PD-L1 status, and found overall response rates of 26\% (95\% CI: 18-36), 18\% (95\% CI: 13-24), and 15\% (95\% CI 11-19) in patients with ≥5\% PD-L1-positive immune cells (IC2/3 subgroup), in patients with ≥1\% PD-L1-positive immune cells (IC1/2/3 subgroup), and in all patients, respectively [@Rosenberg2016].
+All three rates exceeded the historical control rate of 10\%.
+Then, in March 2021, the approval in this indication was voluntarily withdrawn by the sponsor following negative results from a randomized phase III study [@Powles2018].
+In the phase III study, 931 patients were randomly assigned to receive atezolizumab or chemotherapy in a 1:1 ratio, and the same hierarchical fixed-sequence testing procedure as in the phase II study was used.
+The phase III study found that overall survival did not differ significantly between the atezolizumab and chemotherapy groups of the IC2/3 subgroup (median survival 11.1 months [95\% CI: 8.6-15.5] versus 10.6 months [95\% CI: 8.4-12.2]), so no further testing was conducted for the primary endpoint [@Powles2018].
+Further analyses revealed that while the response rates to atezolizumab were comparable to those seen in the phase II study, the response rates to chemotherapy were much higher than the historical control rate of 10\%.
+The overall response rates to chemotherapy were 21.6\% (95\% CI: 14.5-30.2), 14.7\% (95% CI: 10.9-19.2), and 13.4\% (95\% CI: 10.5-16.9) for the IC2/3 subgroup, IC1/2/3 subgroup, and all patients, respectively.
+The overall response rates to atezolizumab were 23\% (95\% CI: 15.6-31.9), 14.1\% (95\% CI: 10.4-18.5), and 13.4\% (95\% CI: 10.5-16.9) for the IC2/3 subgroup, IC1/2/3 subgroup, and all patients, respectively.
+These results indicate that PD-L1 status is a prognostic biomarker for both standard of care chemotherapies that comprised the control arm as well as atezolizumab in this patient population.
+
+We wish to re-design the phase II trial of atezolizumab using a two-arm randomized design with sequential predictive probability monitoring.
+We willuse the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
+We focus here on the main biomarker subgroup of interest, the IC2/3 subgroup.
+We design the study with a null response rate of 0.1 in both arms, and an alternative response rate of 0.25 in the atezolizumab arm.
+So we set `p_null = c(0.1, 0.1)` and `p_alt = c(0.1, 0.25)`.
+We plan the study with 100 participants, assuming that the total sample size available is similar to the 310 used in the actual single-arm phase II trial, and that a third of that patient population fall into our desired biomarker subgroup.
+So we set `N = c(50, 50)` to indicate equal allocation of the 100 patients to each of the two arms.
+We will check for futility after every 10 patients are enrolled on each arm, so we set `n = cbind(seq(10, 50, 10), seq(10, 50, 10))`, a matrix showing the interim analysis sample sizes in each of the two arms.
+To design the study, we need to calibrate the design over a range of posterior probability thresholds and predictive probability thresholds.
+In this example we will consider posterior thresholds of 0.9, 0.91, 0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 0.98, and 0.99, and predictive thresholds of 0.05, 0.1, 0.15, and 0.2.
+So we set `pp_threshold = seq(0.9, 0.99, 0.01)` and `ppp_threshold = seq(0.05, 0.2, 0.05)`.
+We selected these posterior thresholds because we want to have a posterior probability of at least 0.9 that our response rate exceeds the null.
+And we selected these predictive thresholds because we would  not want to stop the trial early for futility if there was more than a 20\% chance of success when full enrollment was reached.
+We use the defaults for all remaining arguments, including the defaults of `S=5000` and `nsim=1000`, which we believe are sufficient to obtain stable results.
+Because of the inherent computational intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
 The user will be responsible for setting up a call to `future::plan()` that is appropriate to their operating environment and simulation setting.
-The example in this case study was run on a Unix server with 192 cores, and we wished to use 76 cores to accommodate the 76 distinct designs that result from the 19 posterior by 4 predictive threshold grid.
-Because the code takes some time to run, the results of the below example code are available as a dataset called `one_sample_cal_tbl` included in the \CRANpkg{ppseq} package.
+Because the code takes some time to run, the results of the below example code are available as a dataset called `two_sample_cal_tbl` included in the \CRANpkg{ppseq} package.
 
@@
 
-plan(multicore(workers = 76))
+future::plan(future::multicore(workers = 40))
 
-one_sample_cal_tbl <- 
+two_sample_cal_tbl <- 
   calibrate_thresholds(
-    p_null = 0.1, 
-    p_alt = 0.2, 
-    n = seq(5, 95, 5),
-    N = 95, 
-    pp_threshold = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94, 
-                     0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1),
+    p_null = c(0.1, 0.1), 
+    p_alt = c(0.1, 0.25),
+    n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+    N = c(50, 50), 
+    pp_threshold = seq(0.9, 0.99, 0.01),
     ppp_threshold = seq(0.05, 0.2, 0.05),
     direction = "greater", 
-    delta = NULL, 
+    delta = 0, 
     prior = c(0.5, 0.5), 
@@
     nsim = 1000
-  )
-```
-
-Next we print the results table using the `print()` option, and limited to designs with type I error between 0.05 and 0.1, and a minimum power of 0.7.
-We find that 35 of the 76 designs meet these criteria for type I error and power.
-
-```{r}
-print(
-  one_sample_cal_tbl,
-  type1_range = c(0.05, 0.1),
-  minimum_power = 0.7
-)
-```
-
-We use the `optimize_design()` function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
-We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.
-It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91.
-The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1.
-It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82.
-For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.
-
-```{r}
-optimize_design(
-  one_sample_cal_tbl, 
-  type1_range = c(0.05, 0.1), 
-  minimum_power = 0.7
   )
@@
 p[[2]]
+```
+
+We use the `optimize_design()` function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
+We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.
+It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91.
+The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1.
+It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82.
+For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.
+
+```{r}
+optimize_design(
+  one_sample_cal_tbl, 
+  type1_range = c(0.05, 0.1), 
+  minimum_power = 0.7
+)
 ```

### 2022-10-03 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
updating datasets with different thresholds

@@
 For computational speed in this basic example, we set `S=50`, but in practice we would want to draw many more posterior samples, and values of 5000 or more should be considered.
+If any `NA` value was returned in this table, it would indicate that at that interim look you would never stop, regardless of the number of responses.
 
@@
 
-To compare these optimal designs with all other designs, we can use the `plot()` function with the `plotly = TRUE` option to obtain interactive visualizations, or the `plotly = FALSE` option to obtain static visualizations.
+We can compare the design options that meet our desired range of type I error and minimum power by passing the results of our call to the `calibrate_thresholds()` function to the `plot()` function.
+The `plotly = TRUE` option produces interactive visualizations or the `plotly = FALSE` option produces static visualizations.
 
@@
 plot(
-  one_sample_cal_tbl, 
+  two_sample_cal_tbl, 
   type1_range = c(0.05, 0.1), 
@@
 
-```{r echo = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with ggplot2.
-The accuracy designs are in the plot on the left and the efficiency designs are on the right.
+```{r echo = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package.
+The accuracy designs are on the left and the efficiency designs are on the right.
 The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
 p <- plot(
-  one_sample_cal_tbl,
+  two_sample_cal_tbl,
   type1_range = c(0.05, 0.1),
@@
 
-```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with plotly.
+```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with the {plotly} package.
 The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond.
@@
 
-We use the `optimize_design()` function to obtain the details of the optimal accuracy and optimal efficiency designs, limited to our desired range of type I error and minimum power.
-We find that the optimal accuracy design is the one with posterior threshold 0.9 and predictive threshold 0.05.
-It has a type I error of 0.072, power of 0.883, average sample size under the null of 51, and average sample size under the alternative of 91.
-The optimal efficiency design is the one with posterior threshold of 0.92 and predictive threshold of 0.1.
-It has a type I error of 0.06, power of 0.796, average sample size under the null of 39, and average sample size under the alternative of 82.
-For comparison, the original design of the atezolizumab expansion cohort in mUC, with a single look for futility after the first 14 patients, has a type I error of 0.005, power of 0.528, average sample size under the null of 76, and average sample size under the alternative of 92.
+We see that the optimal efficiency design is the one with posterior threshold 0.92 and predictive threshold 0.05.
+It has a type I error of 0.07, power of 0.7, average sample size under the null of 29 per arm, and average sample size under the alternative of 46 per arm.
+This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment.
 
@@
 optimize_design(
-  one_sample_cal_tbl, 
+  two_sample_cal_tbl, 
   type1_range = c(0.05, 0.1), 
@@
 
-```{r, echo = FALSE, include=knitr::is_html_output(), eval=FALSE, fig.cap = "Plot of design options made with plotly.
-The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
-
-# Not currently evaluating this code chunk. I was trying to further manipulate the interactive plots to get them to be placed side-by-side, but I just don't think it's possible with two legends
-
-# Above instead I think it is best to only include one of the plots as an example in the html paper version
-library(dplyr)
-x <- one_sample_cal_tbl
-type1_range <- c(0.05, 0.1)
-minimum_power <- 0.7
-
-plot_x <-
-  rename(
-    mutate(
-      filter(
-        x$res_summary,
-        prop_pos_null >= type1_range[1] &
-          prop_pos_null <= type1_range[2] &
-          prop_pos_alt >= minimum_power
-      ),
-      Design = paste0(
-        "Posterior threshold = ", pp_threshold,
-        " and predictive threshold = ", ppp_threshold
-      ),
-      ab_dist_metric = ((prop_pos_null - 0)^2 +
-        (prop_pos_alt - 1)^2)^(1 / 2),
-      n_dist_metric = ((mean_n1_null - min(mean_n1_null))^2 +
-        (mean_n1_alt - max(mean_n1_alt))^2)^(1 / 2)
-    ),
-    `Type I error` = prop_pos_null,
-    Power = prop_pos_alt,
-    `Average N under the null` = mean_n1_null,
-    `Average N under the alternative` = mean_n1_alt,
-    `Distance to optimal efficiency` = n_dist_metric,
-    `Distance to optimal accuracy` = ab_dist_metric
-  )
-
-plot_ab <-
-  mutate(
-    ungroup(
-      slice(
-        group_by(
-          arrange(
-            plot_x,
-            `Distance to optimal accuracy`,
-            -pp_threshold,
-            -ppp_threshold
-          ),
-          `Distance to optimal accuracy`
-        ),
-        1
-      )
-    ),
-    optimal_accuracy = ifelse(row_number() == 1, TRUE, FALSE)
-  )
-
-plot_nn <-
-  mutate(
-    ungroup(
-      slice(
-        group_by(
-          arrange(
-            plot_x,
-            `Distance to optimal efficiency`,
-            -pp_threshold, -ppp_threshold
-          ),
-          `Distance to optimal efficiency`
-        ),
-        1
-      )
-    ),
-    optimal_efficiency = ifelse(row_number() == 1, TRUE, FALSE)
-  )
-
-p1 <-
-  ggplot(
-    plot_ab,
-    aes(
-      x = `Type I error`,
-      y = Power,
-      color = `Distance to optimal accuracy`,
-      Design = Design,
-      `Average N under the null` = `Average N under the null`,
-      `Average N under the alternative` = `Average N under the alternative`
-    )
-  ) +
-  geom_point(
-    shape = ifelse(plot_ab$optimal_accuracy == TRUE, 18, 19),
-    size = ifelse(plot_ab$optimal_accuracy == TRUE, 4, 2)
-  ) +
-  ylim(0, 1) +
-  xlim(0, 1) +
-  labs(
-    x = "Type I error",
-    y = "Power"
-  ) +
-  scale_color_viridis_c() +
-  theme_bw() +
-  theme(legend.position = "bottom")
-
-p2 <-
-  ggplot(
-    plot_nn,
-    aes(
-      x = `Average N under the null`,
-      y = `Average N under the alternative`,
-      color = `Distance to optimal efficiency`,
-      Design = Design,
-      `Type I error` = `Type I error`,
-      Power = Power
-    )
-  ) +
-  geom_point(
-    shape = ifelse(plot_nn$optimal_efficiency == TRUE, 18, 19),
-    size = ifelse(plot_nn$optimal_efficiency == TRUE, 4, 2)
-  ) +
-  ylim(
-    min(plot_x$`Average N under the null`),
-    max(plot_x$`Average N under the alternative`)
-  ) +
-  xlim(
-    min(plot_x$`Average N under the null`),
-    max(plot_x$`Average N under the alternative`)
-  ) +
-  labs(
-    x = "Average N under the null",
-    y = "Average N under the alternative"
-  ) +
-  scale_color_viridis_c() +
-  theme_bw() +
-  theme(legend.position = "bottom")
-
-ggplotly(p1, width = 4, height = 4)
-# ggplotly(p2, width = 4, height = 4)
-```
-
-In this case we may choose to use the optimal efficiency design, which has the desirable trait of a very small average sample size under the null of just 39 patients, while still maintaining reasonable type I error of 0.06 and power of 0.796.
-This design would allow us to stop early if the treatment were inefficacious, thus preserving valuable financial resources for use in studying more promising treatments and preventing our human subjects from continuing an ineffective treatment.
-Finally, we generate the decision table associated with the selected design for use in making decision at each interim analysis during the conduct of the trial.
-Because of the computational time involved, the results of the below example code are available as a dataset called `one_sample_decision_tbl` included in the \CRANpkg{ppseq} package.
+Finally, we generate the decision table associated with the selected design for use in making decisions at each interim analysis during the conduct of the trial.
+Because of the computational time involved, the results of the below example code are available as a dataset called `two_sample_decision_tbl` included in the \CRANpkg{ppseq} package.
 In the results table, we see that at the first interim futility look after just 5 patients, we would not stop the trial.
@@
 
-one_sample_decision_tbl <- 
+two_sample_decision_tbl <- 
   calc_decision_rules(
-    n = seq(5, 95, 5),
-    N = 95, 
+    n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+    N = c(50, 50),
     theta = 0.92, 
-    ppp = 0.1, 
-    p0 = 0.1, 
+    ppp = 0.05, 
+    p0 = NULL, 
     direction = "greater", 
-    delta = NULL, 
+    delta = 0, 
     prior = c(0.5, 0.5), 
@@
 
-```{r}
-one_sample_decision_tbl
-```
+The resulting table contains all possible combinations of responses in the two arms, so is quite long.
+We can visualize the resulting decision rules by passing the results of the call to the `calc_decision_rules()` function to the `plot()` function, which returns a faceted plot according to the interim look.
+The color denotes whether the trial should proceed (green) or stop (red) for a given combination of number of responses in the control arm (x-axis) and number of responses on the experimental arm (y-axis).
+For example, we see that at the second interim look when there are 20 patients enrolled on each arm, if there were 10 responses in the control arm, we would stop the trial if there were 8 or fewer responses on the experimental arm; otherwise we would proceed with enrollment.
 
@@
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
-plot(one_sample_decision_tbl, plotly = FALSE)
+plot(two_sample_decision_tbl, plotly = FALSE)
 ```
@@
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
-plot(one_sample_decision_tbl)
+plot(two_sample_decision_tbl)
 ```

### 2022-10-03 data-raw/one_sample_cal_tbl.R

@@ -5,16 +5,14 @@ library(future)
 
 set.seed(123)
 
-future::plan(future::multicore(workers = 56))
+future::plan(future::multicore(workers = 40))
 
 one_sample_cal_tbl <- 
   calibrate_thresholds(p_null = 0.1, 
                        p_alt = 0.2, 
                        n = seq(5, 95, 5),
                        N = 95, 
-                       pp_threshold = c(0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 
-                                        0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 
-                                        0.98, 0.99),
+                       pp_threshold = seq(0.9, 0.99, 0.01),
                        ppp_threshold = seq(0.05, 0.2, 0.05),
                        direction = "greater", 
                        delta = NULL, 

### 2022-10-03 data-raw/two_sample_cal_tbl.R

@@ -5,16 +5,14 @@ library(future)
 
 set.seed(123)
 
-future::plan(future::multicore(workers = 56))
+future::plan(future::multicore(workers = 40))
 
 two_sample_cal_tbl <- 
   calibrate_thresholds(p_null = c(0.1, 0.1), 
                        p_alt = c(0.1, 0.25), 
                        n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
                        N = c(50, 50), 
-                       pp_threshold = c(0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 
-                                        0.92, 0.93, 0.94, 0.95, 0.96, 0.97, 
-                                        0.98, 0.99),
+                       pp_threshold = seq(0.9, 0.99, 0.01),
                        ppp_threshold = seq(0.05, 0.2, 0.05),
                        direction = "greater", 
                        delta = 0, 

### 2022-10-03 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
first paper revision draft

@@
 The approval was based on the results of a single-arm phase II study in 310 patients [@Rosenberg2016].
-The phase II study used a hierarchical fixed-sequence testing procedure to test increasingly broad subgroups of patients based on PD-L1 status, and found overall response rates of 26\% (95\% CI: 18-36), 18\% (95\% CI: 13-24), and 15\% (95\% CI 11-19) in patients with ≥5\% PD-L1-positive immune cells (IC2/3 subgroup), in patients with ≥1\% PD-L1-positive immune cells (IC1/2/3 subgroup), and in all patients, respectively [@Rosenberg2016].
+The phase II study used a hierarchical fixed-sequence testing procedure to test increasingly broad subgroups of patients based on PD-L1 status, and found overall response rates of 26\% (95\% CI: 18-36), 18\% (95\% CI: 13-24), and 15\% (95\% CI 11-19) in patients with $\geq5\%$ PD-L1-positive immune cells (IC2/3 subgroup), in patients with $\geq1\%$ PD-L1-positive immune cells (IC1/2/3 subgroup), and in all patients, respectively [@Rosenberg2016].
 All three rates exceeded the historical control rate of 10\%.
@@
 
-```{r message = FALSE, fig.height = 9, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
+```{r message = FALSE, fig.height = 6, fig.width = 8, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
@@
 
-```{r message = FALSE, fig.height = 9, fig.width = 6, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
+```{r message = FALSE, fig.height = 6, fig.width = 8, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}

### 2022-10-03 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -1,3 +1,6 @@
+# Generated by `rjournal_article()` using `knitr::purl()`: do not edit by hand
+# Please edit zabor-hobbs-kane.Rmd to modify this file
+
 ## ----setup, include=FALSE-----------------------------------------------------
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 library(plotly)
@@ -12,58 +15,76 @@ library(ggplot2)
 library(ppseq)
 
 
+## ----message = FALSE----------------------------------------------------------
+set.seed(123)
+
+cal_thresh <-
+  calibrate_thresholds(
+    p_null = 0.1, 
+    p_alt = 0.4,
+    n = seq(5, 15, 5), 
+    N = 15,
+    pp_threshold = c(0.85, 0.9),
+    ppp_threshold = c(0.1, 0.2),
+    S = 50, 
+    nsim = 50
+    )
+
+
+## -----------------------------------------------------------------------------
+print(cal_thresh)
+
+
+## -----------------------------------------------------------------------------
+optimize_design(cal_thresh, type1_range = c(0.025, 0.1), minimum_power = 0.75)
+
+
+## -----------------------------------------------------------------------------
+calc_decision_rules(
+  n = seq(5, 15, 5), 
+  N = 15,
+  theta = 0.9,
+  ppp = 0.2,
+  p0 = 0.1,
+  S = 50
+)
+
+
 ## ----eval = FALSE-------------------------------------------------------------
 #> library(future)
 #> 
 #> set.seed(123)
 #> 
-#> plan(multicore(workers = 76))
+#> future::plan(future::multicore(workers = 40))
 #> 
-#> one_sample_cal_tbl <-
+#> two_sample_cal_tbl <-
 #>   calibrate_thresholds(
-#>     p_null = 0.1,
-#>     p_alt = 0.2,
-#>     n = seq(5, 95, 5),
-#>     N = 95,
-#>     pp_threshold = c(0, 0.7, 0.74, 0.78, 0.82, 0.86, 0.9, 0.92, 0.93, 0.94,
-#>                      0.95, 0.96, 0.97, 0.98, 0.99, 0.999, 0.9999, 0.99999, 1),
+#>     p_null = c(0.1, 0.1),
+#>     p_alt = c(0.1, 0.25),
+#>     n = cbind(seq(10, 50, 10), seq(10, 50, 10)),
+#>     N = c(50, 50),
[…]

### 2022-10-06 inst/ppseq-package-paper/zabor-hobbs-kane-1-review-response.Rmd
my draft of revision 1 responses

@@
 ```
+
+<br>
+<br>
 
@@
 We agree that there were parts of the paper that were unclear or lacked sufficient detail, and have strived to address all of your comments in an effort to improve the clarify and accuracy throughout.
+Please see our detailed responses that follow.
 
@@
 
-Thank you for pointing this out. We agree, and have added this detail to both the abstract and the last paragraph of the introduction.
+Thank you for pointing this out. We have added this important detail to both the Abstract and the last paragraph of the Introduction.
 
@@
 The goal of this paragraph was to introduce readers to the often large sample sizes that have been used in recent expansion cohorts, which is part of the problem addressed by this package.
-We see how the wording did not emphasize this point properly, though, and have reworded in response to this comment.
+We see how the wording did not emphasize this point properly, though, and have rewritten the entire paragraph in response to this comment.
 Thank you.
@@
 
-We agree that the way this sentence was worded was not accurate. We have reworded this section to improve clarity and accuracy.
+We agree that the way this sentence was worded was not accurate.
+Our intention was to introduce the reader to the possibility of implementing Bayesian designs with consideration of frequentist type I error and power.
+We have reworded this section to improve clarity and accuracy.
 
@@
 
-Thank you, we have written the case study a bit to be a two-sample example in order to differentiate from the added basic example, and have also added some information justifying the selected thresholds.
+Thank you, we have rewritten the case study a bit to be a two-sample example in order to differentiate from the added basic example, and have also added some information justifying the selected thresholds.
+The previous thresholds were a holdover from another application, and we apologize for not cleaning this up.
+The thresholds now represent an equi-spaced grid as appropriate in this setting.
 
@@
 
-Thank you for this suggestion, we have rescaled the axes.
+Thank you for this suggestion, we have rescaled the axes so that the accuracy plot axes are limited to the ranges of considered values, and the efficiency plot is a close-up square.
 
 > - Page 7: In the first line of the tibble `one_sample_decision_tbl` there is NA,NA. If I am not mistaken this is interpreted as “never stop”. I will explicitly indicate this.
+
+Thank you, we have changed to a basic example that can easily be run so now there is no NA, but we have still made a note that "If any NA value was returned in this table, it would indicate that at that interim look you would never stop, regardless of the number of responses."
+
 > - There are some typos in the bibliography. For example, “r” in place of “R”, “phase i” in place of “phase I”. Please double check.
+
+These corrections have been made, thank you.
+
 > - I think that the paper would benefit of an additional toy example that can be run on a personal computer rather than on a server.
@@
 
-We agree, and have added a toy example that can easily be run interactively prior to presenting the more realistic case study.
+We agree, and have added a basic example to the Package demonstration section that can easily be run interactively, which is followed by an altered Case study section.
 
@@
 
-In parentheses we have added the definition of the acronym RECIST: Response Evaluation Criteria in Solid Tumors.
+In parentheses we have added the definition of the acronym RECIST (Response Evaluation Criteria in Solid Tumors)
 
@@
 
-Thank you, we have made this correction.
+We have made this correction, thank you.
 
@@
 
-This is indeed a compilation error. I am not sure why this is being generated or how to suppress it, so hopefully it can be fixed when the article is finalized for publication,
+This is indeed a compilation error.
+I am not sure why this is being generated or how to suppress it, so we will work with the editorial team to ensure it isn't displayed when the article is finalized for publication.
+
+<br>
+<br>
 
@@
 Thank you for taking the time to conduct such a thorough review of our paper, we do believe that it has been improved as a result.
+Please see our detailed responses below.
 
@@
 
-Thank you for this comment, we agree that much detail was overlooked in the Package Overview section, and we have expanded this section to include much more detail of the underlying algorithm, and to link the notation to the preceding section.
-We believe this is much easier to follow now.
-We have also added a section for a toy example that can easily be run interactively prior to presenting the more detailed case study.
+Thank you for this comment, we agree that much detail was overlooked in the Package Overview section, and we have expanded this section to include much more detail of the underlying algorithm, and to link the notation from the preceding section to the function arguments.
+We believe this section is much easier to follow now.
+We have also added a new section called Package demonstration that includes a basic example that can easily be run interactively, which is followed by an altered Case study section.
 
@@
 This is meant to be used at the design stage of a clinical trial, when there is ample time to allow for computation.
-Once the design has been selected and decided, no additional computations are required throughout the conduct of the trial, as the decision table resulting from `calc_decision_rules()` is available for decisions at each interim look, making trial conduct fast at seamless at each interim analysis and at the end of the trial.
+Once the design has been selected and the decision table generated with the function `calc_decision_rules()`, no additional computations are required throughout the conduct of the trial.
+As a result, the trial can be conducted seamlessly at each interim analysis, and through the end of the trial.
 
@@
 
-> 3.
+>3.
 It was not clear to me what each function was doing computationally.
@@
 For nsim iterations:
-i. Simulate a realization of $p^r$ from the prior $\Beta(a_0, b_0)$
-ii. Simulate a realization of $x^r | n, p^r ~ \Binomial(n, p^r)$
-iii. Calculate the probability of success in that trial $$PP^r=\Pr(p^r>p_0|x^r, N, n)=\int_{p_0}^1 f_p(q|x^r, N, n)\,dq=\sum_{x^*}^{N-n}\int_{p_0}^1\Pr(X^*=x^*|q,N-n)f_p(q|x^r,N,X^*=x^*)\,dq
+    i. Simulate a realization of $p^r$ from the prior $\mbox{Beta}(a_0, b_0)$
+    ii. Simulate a realization of $x^r | n, p^r ~ \mbox{Binomial}(n, p^r)$
+    iii. Calculate the probability of success in that trial 
+    $$PP^r=\Pr(p^r>p_0|x^r, N, n)=\int_{p_0}^1 f_p(q|x^r, N, n)\,dq=\sum_{x^*}^{N-n}\int_{p_0}^1\Pr(X^*=x^*|q,N-n)f_p(q|x^r,N,X^*=x^*)\,dq$$
 
-> Then the PPP is the the proportion of $PP^r>\theta^*$.
+>Then the PPP is the the proportion of $PP^r>\theta^*$.
 If this is the case, then the integral term has a closed form and posterior simulations do not seem necessary, in which case the function should run much quicker than it currently does.
@@
 
-Thank you for drawing our attention to this oversight.
-We have added much more detail about the underlying computational algorithm for the `calibrate_thresholds()` function in the Package Overview section, and hope that this clarifies the functionality.
+We apologize that the description in the Package overview was not clear.
+Please see the expanded section, which now includes all of the details of the implemented algorithm.
+Posterior sampling is required to account for the uncertainty in $p$.
+In the algorithm you describe above, using a single estimate $p^r$ will result in a too-narrow predictive distribution.
+In our scheme and using your notation, we rather draw `nsim` iterations $x^r | n, p^r ~ \mbox{Binomial}(n, p^r)$, then we take `S` posterior samples $p^s|x^r \sim \mbox{Beta}(a_0 + x^r, b_0 + n - x^r)$.
 
@@
 We have added much more detail about the computations being done, with notation, to this section, and hope it is now more clear.
-Additionally, in response to another reviewer's comment we have renamed this section "Design selection" and removed reference to optimization, since no true mathematical optimization was being conducted.
+Additionally, in response to another reviewer's comment we have renamed this section "Design selection" and removed reference to optimization, since no true mathematical optimization is being conducted.

### 2022-10-06 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -5,6 +5,7 @@
 knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 library(plotly)
 library(ggplot2)
+library(patchwork)
 
 
 ## ----eval = FALSE-------------------------------------------------------------
@@ -82,27 +83,41 @@ calc_decision_rules(
 #> )
 
 
-## ----echo = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package. The accuracy designs are on the left and the efficiency designs are on the right. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
+## ----echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package. The accuracy designs are on the left and the efficiency designs are on the right. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."----
 p <- plot(
   two_sample_cal_tbl,
   type1_range = c(0.05, 0.1),
   minimum_power = 0.7,
   plotly = FALSE
   )
-p
+
+# change limits based on reviewer comment
+p1_v2 <- 
+  p[[1]] + 
+  scale_x_continuous(limits = c(0.05, 0.1), breaks = seq(0.05, 0.1, 0.01)) + 
+  scale_y_continuous(limits = c(0.7, 1), breaks = seq(0.7, 1, 0.05))
+p2_v2 <-
+  p[[2]] +
+  scale_x_continuous(limits = c(26, 30)) + 
+  scale_y_continuous(limits = c(43, 47))
+
+p1_v2 + p2_v2
 
 
 ## ----echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with the {plotly} package. The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond. A similar plot is available for the accuracy design options."----
-#> 
-#> # Right now only including the efficiency design plot in the html paper version
 #> p <- plot(
-#>   one_sample_cal_tbl,
+#>   two_sample_cal_tbl,
 #>   type1_range = c(0.05, 0.1),
 #>   minimum_power = 0.7,
-#>   plotly = TRUE
[…]

### 2022-10-14 inst/ppseq-package-paper/zabor-hobbs-kane-1-review-response.Rmd
final review 1 response

@@
 We have reworded this for clarity, thank you.
+The trial would stop at a given look if the number of observed responses is less than or equal to $r$, otherwise the trial would continue enrolling if the number of observed responses is greater than $r$.
+At the end of the trial when the maximum planned sample size is reached, the treatment would be considered promising if the number of observed responses is greater than $r$.
 
@@
 We are sorry that the article in it's original form was difficult to follow.
-We hope the changes have made it more understandable, and that all of the needed detail is now included.
+We hope the changes have made it more understandable, and that all of the needed details are now included.
 Thank you for taking the time to conduct such a thorough review of our paper, we do believe that it has been improved as a result.
@@
 In the algorithm you describe above, using a single estimate $p^r$ will result in a too-narrow predictive distribution.
-In our scheme and using your notation, we rather draw `nsim` iterations $x^r | n, p^r ~ \mbox{Binomial}(n, p^r)$, then we take `S` posterior samples $p^s|x^r \sim \mbox{Beta}(a_0 + x^r, b_0 + n - x^r)$.
+In our scheme and using your notation, we rather draw `nsim` iterations $x^r | n, p^r \sim \mbox{Binomial}(n, p^r)$, then we take `S` posterior samples $p^s|x^r \sim \mbox{Beta}(a_0 + x^r, b_0 + n - x^r)$.
 

### 2022-10-19 README.Rmd
change to me

@@
 The {ppseq} package provides functions to design clinical trials using Bayesian sequential predictive probability monitoring.
-Functionality is available to design [one-arm](https://www.emilyzabor.com/ppseq/articles/one_sample_expansion.html) or [two-arm](https://www.emilyzabor.com/ppseq/articles/two_sample_randomized.html) trials by searching over a grid of combinations of posterior and predictive thresholds and identifying the optimal design according to two criteria: accuracy and efficiency.
+Functionality is available to design [one-arm](https://www.emilyzabor.me/ppseq/articles/one_sample_expansion.html) or [two-arm](https://www.emilyzabor.me/ppseq/articles/two_sample_randomized.html) trials by searching over a grid of combinations of posterior and predictive thresholds and identifying the optimal design according to two criteria: accuracy and efficiency.
 Interactive plotting allows easy comparison of the various design options and easy trial implementation through decision rule plots.
@@
 
-See the vignettes for the [one-sample](https://www.emilyzabor.com/ppseq/articles/one_sample_expansion.html) and [two-sample](https://www.emilyzabor.com/ppseq/articles/two_sample_randomized.html) cases for additional details about available features and options.
+See the vignettes for the [one-sample](https://www.emilyzabor.me/ppseq/articles/one_sample_expansion.html) and [two-sample](https://www.emilyzabor.me/ppseq/articles/two_sample_randomized.html) cases for additional details about available features and options.

### 2022-10-19 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
change to me

@@
     - Cleveland, OH 44195 USA
-    url: http://www.emilyzabor.com/
+    url: http://www.emilyzabor.me/
     orcid: 0000-0002-1402-4498
@@
 library(ggplot2)
+library(patchwork)
 ```
@@
 As a result, phase I trials were centered on identifying the maximum tolerated dose (MTD), defined as the highest dose that did not exceed a pre-specified toxicity threshold.
-Phase I dose-escalation trials were designed using the rule-based 3+3 method or the model-based continual reassessment method, among others.
 But advances in drug discovery have produced numerous biomarker-guided non-cytotoxic therapies such as small molecule inhibitors, antibody drug conjugates, immune checkpoint inhibitors, and monoclonal antibodies.
@@
 Instead, lower doses may have equivalent efficacy but lower toxicity as compared to higher doses.
-As a result, these therapies can be difficult to study with traditional dose-escalation designs [@Pestana2020].
-To address this issue, recent phase I trials have included large dose-expansion cohorts, in which additional patients are enrolled in phase 1 after the dose-escalation phase is complete.
-In this setup, the dose-escalation phase is considered phase 1a and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase 1b and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes, or to further characterize the pharmacokinetics and/or pharmacodynamics.
-The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase 2 trials [@Bugano2017].
+As a result, these therapies can be difficult to study with traditional dose-escalation designs such as the rule-based 3+3 and the model-based continual reassessment method, among others [@Pestana2020].
+To address this issue, recent phase I trials have included large dose-expansion cohorts, in which additional patients are enrolled in phase I after the dose-escalation phase is complete.
+In this setup, the dose-escalation phase is considered phase Ia and used to assess the initial safety of multiple doses, then the dose-expansion phase is considered phase Ib and can have a variety of aims including to further refine the safety of one or more doses, to assess preliminary efficacy, to explore the treatment in various disease-specific subtypes, or to further characterize the pharmacokinetics and/or pharmacodynamics.
+The use of dose-expansion cohorts increased from 12\% in 2006 to 38\% in 2011 [@Manji2013] and trials with dose-expansion cohorts led to higher response rates and more frequent success in phase II trials [@Bugano2017].
 
-But despite these successes, recent dose-expansion cohorts have not always been planned in advance, leading to uncertain statistical properties, and have at times included samples sizes that far exceed those typically seen in early phase trials of cytotoxic treatments.
+But despite these successes, recent dose-expansion cohorts have not always been planned in advance, leading to uncertain statistical properties, and have at times included samples sizes that far exceed those typically seen in early phase trials.
 For example, the KEYNOTE-001 trial of pembrolizumab, initially designed as a 3+3 dose-escalation trial, included multiple protocol amendments and ultimately enrolled a total of 655 patients across five melanoma expansion cohorts and 550 patients across four non-small-cell lung cancer expansion cohorts [@Khoja2015].
-In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial bladder cancer ultimately enrolled 97 patients and evaluated 95, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol.
-The expansion cohort in metastatic urothelial carcinoma was rather added later in a protocol amendment in which the sample size was increased from what was initially planned [@Petrylak2018; @Powles2014].
+In a basket trial of atezolizumab, an anti-PD-L1 treatment in patients with a variety of cancers and both with and without PD-L1 expression, an expansion cohort in metastatic urothelial carcinoma ultimately evaluated 95 patients, despite the fact that no expansion cohort in this disease subtype was originally planned in the trial protocol.
+The expansion cohort in metastatic urothelial carcinoma was added through a protocol amendment, and the sample size later was increased from what was initially planned [@Petrylak2018; @Powles2014].
 These enlarged sample sizes raise ethical concerns for patients who enroll in clinical trials, and emphasize the need for rigorous statistical designs to ensure that trials can stop early for futility while maintaining traditional control of type I error and power.
@@
 This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility.
-Interactive plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and the \CRANpkg{plotly} package [@Sievert2020] compare designs based on different thresholds for decision making.
-Moreover, we demonstrate criteria for selecting an ideal predictive probability monitoring design using the \CRANpkg{ppseq} package.
+Static plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and interactive plots produced using the \CRANpkg{plotly} package [@Sievert2020] compare design options based on frequentist type I error and power.
+Moreover, the \CRANpkg{ppseq} package implements two criteria for selecting an ideal predictive probability monitoring design from among the options formed by combinations of posterior and predictive decision thresholds.
 While the \CRANpkg{ppseq} package was developed with early phase oncology clinical trials in mind, the methodology is general and can be applied to any application of sequential futility monitoring in clinical trial design.
@@
 Each patient, denoted by $i$, enrolled in the trial either has a response such that $x_i = 1$ or does not have a response such that $x_i = 0$.
-Then $X = \sum_{i=1}^n x_i$ represents the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients.
+Then $X = \sum_{i=1}^n x_i$ is the number of responses out of $n$ currently observed patients up to a maximum of $N$ total patients.
 Let $p$ represent the true probability of response.
@@
 A posterior probability threshold $\theta$ would be pre-specified during the trial design stage.
-At the end of the trial if the posterior probability exceeded the pre-specified threshold, i.e.
+At the end of the trial, if the posterior probability exceeded the pre-specified threshold, i.e.
 if $\Pr(p>p_0 | X) > \theta$, the trial would be declared a success.
@@
 The hyperparameters of the prior beta distribution are specified with the argument `prior` and default to `c(0.5, 0.5)`, which denotes a $\mbox{Beta}(0.5, 0.5)$ distribution.
-The number of posterior samples are specified with the argument `S`, which defaults to `5000` samples, and the number of simulated trial datasets is specified with the argument `nsim`, which defaults to `1000`.
+The number of posterior samples is specified with the argument `S`, which defaults to `5000` samples, and the number of simulated trial datasets is specified with the argument `nsim`, which defaults to `1000`.
 The additional argument `delta`, which defaults to `NULL` for the one-sample setting, can specify a clinically meaningful difference between groups $\delta = p_1 - p_0$ in the case of a two-sample trial design.
@@
 The proportion of trials simulated under the null hypothesis for which the null hypothesis was rejected is an estimate of the type I error, and the proportion of trials simulated under the alternative hypothesis for which the null hypothesis was rejected is an estimate of the power.
-The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of type I error and minimum power, if desired.
+The `print()` option will print the results summary for each combination of thresholds, filtered by an acceptable range of frequentist type I error and minimum power, if desired.
 Note that the results will be sensitive to the choice of `nsim` and `S`.
@@
 
-Since there are only four design options in this toy example, we print the entire results table using a call to `print()`:
+Since there are only four design options in this toy example, we print the entire results table using a call to `print()`.
+Each row represents a different combination of the considered posterior and predictive thresholds, denotes `pp_threshold` and `ppp_threshold`, respectively.
 
@@
 
-We use `optimize_design()` to identify the optimal accuracy and optimal efficiency designs, subject to type I error between 0.025 and 0.1, specified by `type1_range = c(0.025, 0.1)`, and power of at least 0.75, specified by `minimum_power = 0.75`:
+We use the `optimize_design()` function to identify the optimal accuracy and optimal efficiency designs.
+We constrain consideration to designs with type I error between 0.025 and 0.1, specified by `type1_range = c(0.025, 0.1)`, and power of at least 0.75, specified by `minimum_power = 0.75`.
 
@@
 
-And to ease interim analysis during the course of the trial, we obtain the decision table using the `calc_decision_rules` function, with the thresholds of our selected optimal efficiency design passed as `theta = 0.9` and `ppp_threshold = 0.2`.
+To ease interim analysis during the course of the trial, we obtain the decision table using the `calc_decision_rules()` function, with the thresholds of our selected optimal efficiency design passed as `theta = 0.9` and `ppp_threshold = 0.2`.
 For computational speed in this basic example, we set `S=50`, but in practice we would want to draw many more posterior samples, and values of 5000 or more should be considered.
@@
 
-This table tells us that at the first interim analysis after we have observed the first five patients for response, we would stop the trial if no responses had been observed, and would continue enrolling if at least one response had been observed.
-At the second interim analysis after we have observed the first ten patients for response, we would stop the trial if zero or one responses had been observed, and would continue enrolling if at least two responses had been observed.
-At the end of the trial after all 15 patients have been observed for response, we would declare the new treatment promising if at least three responses had been observed.
+This table tells us that at the first interim analysis after we have observed $n=5$ patients for response, we would stop the trial if $\leq r=0$ responses had been observed, and would continue enrolling if $> r=0$ responses had been observed.
+At the second interim analysis after we have observed $n=10$ patients for response, we would stop the trial if $\leq r=1$ responses had been observed, and would continue enrolling if $> r=1$ responses had been observed.
+At the end of the trial after all $n=N=15$ patients have been observed for response, we would declare the new treatment promising if $> r=2$ responses had been observed.
 
@@
 We use the defaults for all remaining arguments, including the defaults of `S=5000` and `nsim=1000`, which we believe are sufficient to obtain stable results.
-Because of the inherent computational intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2020] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
+Because of the inherent computational intensity in these calculations, this function relies on the \CRANpkg{future} [@Bengtsson2021] and \CRANpkg{furrr} [@Vaughan2021] packages to parallelize computations.
 The user will be responsible for setting up a call to `future::plan()` that is appropriate to their operating environment and simulation setting.
@@
 
-```{r echo = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package.
+```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package.
 The accuracy designs are on the left and the efficiency designs are on the right.
@@
   )
-p
+
+# change limits based on reviewer comment
+p1_v2 <- 
+  p[[1]] + 
+  scale_x_continuous(limits = c(0.05, 0.1), breaks = seq(0.05, 0.1, 0.01)) + 
+  scale_y_continuous(limits = c(0.7, 1), breaks = seq(0.7, 1, 0.05))
+p2_v2 <-
+  p[[2]] +
+  scale_x_continuous(limits = c(26, 30)) + 
+  scale_y_continuous(limits = c(43, 47))
+
+p1_v2 + p2_v2
 ```
@@
 A similar plot is available for the accuracy design options."}
+p <- plot(
+  two_sample_cal_tbl,
+  type1_range = c(0.05, 0.1),
+  minimum_power = 0.7,
+  plotly = FALSE
+  )
+p2_v2 <-
+  p[[2]] +
+  scale_x_continuous(limits = c(26, 30)) + 
+  scale_y_continuous(limits = c(43, 47))
 
 # Right now only including the efficiency design plot in the html paper version
-p <- plot(
-  one_sample_cal_tbl,
-  type1_range = c(0.05, 0.1),
-  minimum_power = 0.7,
-  plotly = TRUE
-  )
-# p[[1]]
-p[[2]]
+ggplotly(p2_v2)
 ```
@@
 
-```{r message = FALSE, fig.height = 6, fig.width = 8, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
+```{r message = FALSE, fig.height = 6, fig.width = 7, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
-plot(two_sample_decision_tbl, plotly = FALSE)
+dtp <- plot(two_sample_decision_tbl, plotly = FALSE)
+
+dtp + 
+  theme(legend.position="bottom")
 ```

### 2022-11-09 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
change back to com
update package paper

@@
     - Cleveland, OH 44195 USA
-    url: http://www.emilyzabor.me/
+    url: http://www.emilyzabor.com/
     orcid: 0000-0002-1402-4498
@@
 The first, called the "optimal accuracy" design, identifies the design that minimizes the Euclidean distance to 0 type I error probability and a power of 1.
-To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_(1-\beta) * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power.
+To accomplish this, the accuracy Euclidean distance (AED) for the design with posterior threshold $k$ and predictive threshold $m$ is calculated as $AED_{km} = w_\alpha * (\alpha_{km} - 0)^2 + w_{(1-\beta)} * ((1-\beta)_{km}-1)^2$, where $w_\alpha$ and $w_{(1-\beta)}$ are optional weights on the type I error and power, respectively, and $\alpha_{km}$ denotes the estimated type I error and $(1-\beta)_{km}$ denotes the estimated power.
 The design with the smallest value of $AED_{km}$ is selected as optimal.
@@
 
-```{r message = FALSE, fig.height = 6, fig.width = 7, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
+```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
 The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}

### 2022-11-09 inst/ppseq-package-paper/zabor-hobbs-kane.R

@@ -145,7 +145,7 @@ optimize_design(
 #>     )
 
 
-## ----message = FALSE, fig.height = 6, fig.width = 7, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."----
+## ----message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot. The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."----
 dtp <- plot(two_sample_decision_tbl, plotly = FALSE)
 
 dtp + 

### 2022-11-15 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
final paper version

@@
 We wish to re-design the phase II trial of atezolizumab using a two-arm randomized design with sequential predictive probability monitoring.
-We willuse the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
+We will use the `calibrate_thresholds()` function to obtain the operating characteristics of designs based on on each combination of posterior and predictive thresholds.
 We focus here on the main biomarker subgroup of interest, the IC2/3 subgroup.

### 2023-04-05 R/calibrate_thresholds.R
fixing bug and deprecated function

@@ -257,7 +257,8 @@ eval_thresh <- function(data, pp_threshold, ppp_threshold, p0, N,
 #'   nsim = 10
 #'   )
 #' 
-#' @importFrom purrr pmap_dfr map cross_df
+#' @importFrom purrr pmap_dfr map 
+#' @importFrom tidyr expand_grid
 #' @importFrom furrr future_map furrr_options
 #' @importFrom dplyr bind_rows full_join select rename ungroup summarize
 #' @export
@@ -275,10 +276,10 @@ calibrate_thresholds <- function(p_null, p_alt, n, N,
     map(seq_len(nsim), ~ sim_dat1(p = p_alt, n = n))
   
   cross_threshold <-
-    cross_df(list(
+    expand_grid(
       pp_threshold = pp_threshold,
       ppp_threshold = ppp_threshold
-    ))
+    )
   
   p0 <- if(length(p_null) == 1) 
     p_null else if(length(p_null) == 2)
@@ -362,11 +363,11 @@ calibrate_thresholds <- function(p_null, p_alt, n, N,
           mean_n0_null = mean(n0_null),
           mean_n1_null = mean(n1_null),
           prop_pos_null = mean(positive_null),
-          prop_stopped_null = mean(sum(n0_null, n1_null) < sum(N)),
+          prop_stopped_null = mean((n0_null + n1_null) < sum(N)),
           mean_n0_alt = mean(n0_alt),
           mean_n1_alt = mean(n1_alt),
           prop_pos_alt = mean(positive_alt),
-          prop_stopped_alt = mean(sum(n0_alt, n1_alt) < sum(N))
+          prop_stopped_alt = mean((n0_alt + n1_alt) < sum(N))
         )
       )
   } else if (length(p_null) == 1) {

### 2023-04-05 inst/ppseq-package-paper/zabor-hobbs-kane.Rmd
fixing bug and deprecated function (#20)
* fixing bug and deprecated function

* Increment version number to 0.2.1.9000

* Update NEWS.md

* updating version

@@
 output: 
-  rjtools::rjournal_web_article
+  rjtools::rjournal_web_article:
+    self_contained: yes
 bibliography: zabor-hobbs-kane.bib
@@
 To our knowledge no such software currently exists in the R programming language.
-This paper introduces the \CRANpkg{ppseq} package for the R software language [@RCT2020], which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility.
+This paper introduces the \CRANpkg{ppseq} package for the R software language, which provides functions to design early phase clinical trials of binary endpoints using sequential predictive probability monitoring for futility.
 Static plots produced using the \CRANpkg{ggplot2} package [@Wickham2016] and interactive plots produced using the \CRANpkg{plotly} package [@Sievert2020] compare design options based on frequentist type I error and power.
@@
 ```{r echo = FALSE, message = FALSE, fig.width = 8, fig.height = 4.4, out.width = "110%", include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of design options made with the {ggplot2} package.
-The accuracy designs are on the left and the efficiency designs are on the right.
-The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond."}
+The accuracy designs are on the left, with type I error on the x-axis and power on the y-axis, and the efficiency designs are on the right, with the average sample size under the null on the x-axis and the average sample size under the alternative on the y-axis.
+Each point represents one potential design based on a specific combination of posterior and predictive thresholds.
+The color represents the Euclidean distance to the top left point.
+The optimal design is indicated by a diamond.
+The optimal accuracy design is the one with posterior threshold 0.86 and predictive threshold 0.1.
+The optimal efficiency design is the one with posterior threshold 0.92 and predictive threshold 0.05."}
 p <- plot(
@@
 ```{r echo = FALSE, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of efficiency design options made with the {plotly} package.
+The average sample size under the null is on the x-axis and the average sample size under the alternative is on the y-axis.
 The color represents the Euclidean distance to the top left point and the optimal design is indicated by a diamond.
+The optimal efficiency design is the one with posterior threshold 0.92 and predictive threshold 0.05.
 A similar plot is available for the accuracy design options."}
@@
 
-```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with ggplot.
-The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
+```{r message = FALSE, fig.height = 6, fig.width = 6, include = knitr::is_latex_output(), eval = knitr::is_latex_output(), fig.cap = "Plot of decision rules made with {ggplot2}.
+Each facet is for the total sample size at a pre-specified interim analysis.
+The number of responses in the control arm is on the x-axis and the number of responses in the experimental arm is on the y-axis.
+The fill color indicates whether the trial should stop (purple) or proceed (yellow) for a given combination of number of responses in the two arms at each interim analysis.
+At a given interim analysis, one can look up the number of observed responses to make a decision about whether the trial should stop or proceed enrollment."}
 dtp <- plot(two_sample_decision_tbl, plotly = FALSE)
@@
 dtp + 
-  theme(legend.position="bottom")
+  theme(legend.position="bottom") + 
+  scale_fill_viridis_d()
 ```
 
-```{r message = FALSE, fig.height = 6, fig.width = 8, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with plotly.
-The color indicates whether the trial should stop or proceed for a given number of responses at each interim analysis."}
-plot(two_sample_decision_tbl)
+```{r message = FALSE, fig.height = 6, fig.width = 8, include=knitr::is_html_output(), eval=knitr::is_html_output(), fig.cap = "Plot of decision rules made with {plotly}.
+Each facet is for the total sample size at a pre-specified interim analysis.
+The number of responses in the control arm is on the x-axis and the number of responses in the experimental arm is on the y-axis.
+The fill color indicates whether the trial should stop (purple) or proceed (yellow) for a given combination of number of responses in the two arms at each interim analysis.
+At a given interim analysis, one can look up the number of observed responses to make a decision about whether the trial should stop or proceed enrollment."}
+ggplotly(
+  plot(two_sample_decision_tbl, plotly = FALSE) + scale_fill_viridis_d()
+)
 ```

### 2024-01-18 README.Rmd
trying to fix github actions issue

--- 2022-08-08 README.Rmd
@@
 [![R-CMD-check](https://github.com/zabore/ppseq/workflows/R-CMD-check/badge.svg)](https://github.com/zabore/ppseq/actions)
-[![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://codecov.io/gh/zabore/ppseq?branch=main)
+[![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zabore/ppseq?branch=main)
 [![CRAN status](https://www.r-pkg.org/badges/version/ppseq)](https://CRAN.R-project.org/package=ppseq)
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
+[![R-CMD-check](https://github.com/zabore/ppseq/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zabore/ppseq/actions/workflows/R-CMD-check.yaml)
 <!-- badges: end -->

### 2024-01-18 README.Rmd
trying to fix github test issue

--- 2022-08-08 README.Rmd
@@
 <!-- badges: start -->
-[![R-CMD-check](https://github.com/zabore/ppseq/workflows/R-CMD-check/badge.svg)](https://github.com/zabore/ppseq/actions)
 [![Codecov test coverage](https://codecov.io/gh/zabore/ppseq/branch/main/graph/badge.svg)](https://app.codecov.io/gh/zabore/ppseq?branch=main)
@@
 [![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
+[![R-CMD-check](https://github.com/zabore/ppseq/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/zabore/ppseq/actions/workflows/R-CMD-check.yaml)
 <!-- badges: end -->
